### 2024-10-19 paper.Rmd
Initial commit

---
title: "No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction"
blinded: 0
authors: 
- name: Adam Bartonicek
  affiliation: Department of Statistics, The University of Auckland
- name: Simon Urbanek
  affiliation: Department of Statistics, The University of Auckland
- name: Paul Murrell
  affiliation: Department of Statistics, The University of Auckland
keywords:
- Interactive data visualization, category theory, abstract algebra, the Gestalt Principles, descriptive statistics
abstract: |
  Interactive data visualization has become a staple of modern data presentation. Yet, despite its growing popularity, there still remains a lack of a formal framework for turning raw data into summary statistics that can then be displayed by interactive graphics. This gap may stem from a subtle yet profound issue: while we would often like treat statistics, graphics, and interaction as independent entities, they are in fact deeply connected. This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids. Specifically, if we want our graphics to support interactive features which break the data into multiple parts and then combine them back together (such as in the case of linked selection), then we need the statistics in our plots to fulfill certain algebraic properties. Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.   
bibliography: references.bib
biblio-style: apalike
nocite: |
  @magritte1929
  @horst2020
output:
  rticles::asa_article:
    includes:
      in_header: "preamble.tex"
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{tikz-cd}
---

```{r}
#| echo: false

# knitr::opts_chunk$set(
#   cache = TRUE
# )

```

# Introduction \label{sec:intro}

> "Effective graphical analysis makes things seem obvious, the effort involved in making the graphical analysis effective is not so obvious." [@unwin2018]

The essence of data visualization is comparison. When we visualize, we split our data into parts, summarize each part one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and finally use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019]. For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent the counts by drawing bars of the corresponding heights. In this way, each bar becomes a visual representation of its own "small data set", a subset of the original data.

To ground this idea in a more concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other. You would construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on. Even from this very simple example, interesting mathematical properties jump out: for example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].

Now, just as we can learn about the contents of our library by organizing books in the way we just described, we can learn about our data by splitting it into parts. A natural question then arises: can we learn more by subdividing our data into even smaller parts and comparing these? For instance, in our book example, we can split each stack of books into two smaller stacks based on the gender of the author. Each smaller stack will then represent the unique combination (Cartesian product) of the two categorical variables (genre and gender). Moreover, there arises a hierarchical relationship between the stacks: by combining two genre-gender stacks (child components) together, we can recover the original genre stack (parent component). Similar situation often arises in interactive graphics, where we use techniques such as linked selection to highlight specific subsets of our data. Either way, by forming these finer partitions, we can represent our data with greater granularity and discover trends that may otherwise hide in the aggregate.

However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves the hierarchical nature of this relationship. In data visualization, there are two popular methods for doing this, known as "stacking" and "dodging" (see Figure \ref{fig:playfair}). These methods have existed at least since the time of William Playfair [-@playfair1801; -@playfair1822]. Let's briefly illustrate them on the example of the classical barplot. In a stacked barplot, bar segments representing the product of the two factor variables are plotted vertically on top of each other ("stacked"; the operation can be also thought of as "highlighting" parts of the bar). In contrast, in a dodged barplot, bars are plotted side-by-side, as clusters of bars. Finally, there is also a third popular method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting (this technique may have been a bit challenging for William Playfair but has been made simple with computers).

```{r playfair}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Examples of the three methods for dealing with bars that have subdivided into segments. Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010)."
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(magick)

p1 <- image_read("./images/playfair1.jpg") |> image_ggplot()
p2 <- image_read("./images/playfair2.jpg") |> image_ggplot()

mpg$manufacturer <- tools::toTitleCase(mpg$manufacturer)

p3 <- ggplot(mpg, aes(manufacturer, fill = factor(year))) +
  geom_bar(position = "identity", alpha = 0.5) +
  scale_x_discrete(labels = abbreviate) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
)

(p1 + p2) / p3
```

Much has been written about the relative merits of stacking, dodging, and layering. For example, layering is only useful with few categories, as blending many colors can make it difficult to tell the categories apart [@franconeri2021; @wilke2019]. Further, in a landmark study, @cleveland1984 showed that people tend to be less accurate when reading information from stacked bar charts as opposed to dodged bar charts. Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below it, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984]. Subsequent research has independently validated these findings and expanded upon them [see e.g. @heer2010; @thudt2016; @quadri2021]. Ultimately, the suboptimal statistical legibility of stacking has led some authors to advise against its use altogether [@kosara2016; @wilke2019], or at least urge caution [see e.g. @byron2008; @cairo2014; @franconeri2021].

However, despite the mixed reception towards stacking in static visualization, it remains a popular choice in interactive data visualization. Among popular interactive data visualization libraries, stacked plots such as bar charts and histograms make frequent appearance [see e.g. @cook2023; @murray2017; @vegalite2022; @vanderplas2018; @sievert2020; @swayne2003; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @wills2008; @ware2019]. What is the reason for this? Has the interactive data visualization community simply failed to catch up with the current best practices? Or is there perhaps more going on?

We argue that not only is stacking a useful technique, but that it can in fact offer valuable insights about the mathematical structure underlying our visualizations. To make the strongest case possible, we bring forward evidence from several levels of the visualization process. We start by discussing the properties of visual perception and how they apply to both static and interactive visualizations. Next, we explore the process of computing statistical summaries that form the basis of our plots. Finally, we dive into the algebraic properties of the functions underlying these statistical summaries. Throughout the text, we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and efficient interactive graphics, we simply cannot treat graphics, statistics, and interaction as independent components. Instead we need to consider them holistically, within the context of their respective visual and mathematical properties. Understanding this deep web of interdependence may in turn empower us to innovate and build new kinds of robust and expressive interactive visualizations.

# Visual Perception and Interactivity

## Visual perception and the Gestalt Principles

Over the course of our evolutionary history, the human brain has evolved a sophisticated pipeline for processing visual information. This pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp. 20]. Many of the findings from this research are directly applicable to data visualization. For example, certain particularly salient visual stimuli are perceived automatically, rapidly (<10 ms), in parallel, and at a constant speed regardless of the number of distractors [@treisman1985]. This phenomenon, known as pre-attentive processing, is key for data visualization, as the right use of visual attributes such as color, contrast, or shape can help us design figures which communicate insights much more effectively [@ware2019].

However, many of the findings from visual perception research are quite specific; to guide effective visualization, we also need some general, high-level rules. One such set of high-level rules are the Gestalt principles of visual perception [see e.g. @cairo2012; @ware2019; @rosli2015; @vanderplas2020; @todorovic2008; @pinker1990]. These principles formulate a set of fundamental "laws" related to the perception of patterns, primarily visual. They were discovered by the founders of the Gestalt school of psychology early in the 20th century, primarily by means of self-observation, and, despite not being grounded in sound understanding of the neural mechanisms of the brain, have nevertheless been later independently supported by findings from cognitive neuroscience [@guberman2017; @ware2019; @wertheimer1938].

Among the Gestalt principles, there are a few which are particularly relevant to data visualization. First, there are the two principles of proximity and similarity. These principles state that objects that which appear in close together and/or similar in appearance tend to be perceived as part of single group. Second, the closure and common region principles state that regions with closed contours tend to be perceived as single, unified objects. Finally, the figure and ground principle states we tend to see one part of a scene as the central object (the "figure") while the rest is perceived as the background (the "ground"). This can be modulated by other Gestalt principle: for example, smaller, darker, and symmetric objects with closed contours are more likely to be perceived as the figure [@ware2019].

Whether we intend them to or not, the Gestalt principles affect how our data visualizations are perceived. As such, it is a good idea to try to use them to our advantage. Sometimes, little effort is needed. Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups by the observer. In other cases, however, our design choices can have a significant impact. For instance, another way to have objects to be perceived as part of a group is to draw a closed shape around them [@cairo2012]. Alternatively, if we want to highlight a particular part of a data set, we can leverage the figure and ground principle and use attributes such as color and saturation to make it stand out [for specific examples, see @franconeri2021; @wilke2019, p. 34].

Finally, ignoring or violating the Gestalt principles can make our graphics less legible or even misleading. @tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements he termed "chartjunk". Similarly, @cairo2019 highlights many cases of graphics that violate the Gestalt principles. Finally, there is also the issue of well-designed graphics that communicate no statistical information at all, or even information that is blatantly false [@cairo2014; @franconeri2021; @cairo2019; @gelman2013; @tufte2001]. Whether intentional or not, poor design can have a significant impact on the legibility and effectiveness of statistical graphics.

## Gestalt and Interactivity

The Gestalt principles apply to both static and interactive visualizations, however, the way they apply is different. Static data visualizations can be designed well or poorly, but this design quality does not change over time. In contrast, interactive visualizations change over time, with user interaction. As such, an interactive visualization may conform to the Gestalt principles at one instance of time and violate them in another. In fact, interactive visualizations can even differ in how well they leverage design principles throughout time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].

Thus, interactivity inherently places further constraints the design of figures. If we want well-designed interactive figures, we need to ensure that they conform to good design principles across all states induced by user-interaction. This may not always be easy. We will investigate this idea in the following section, on the example of linked selection.

## Case study: Linked Selection

Linked selection, also known as linked brushing or linked highlighting, has been consistently ranked as one of the most useful interactive features in data visualization [see e.g. @buja1996; @heer2012; @wilhelm2003; @ware2019; @ward2015]. By clicking or clicking-and-dragging over objects in one plot, users can highlight parts of objects corresponding to the selected cases across all other "linked" plots. This allows the user to rapidly "drill-down" [@dix1998; @theus2002] and explore trends across different dynamically-generated subsets of the data [similar to interactive filtering, see e.g. @heer2012]. The ability to quickly materialize different views of the data makes linked selection a versatile tool for data exploration.

Yet, despite its usefulness, linked selection also imposes some fundamental constraints on the figure. Specifically, all objects we visualize need to be able to:

1. **Be selected**
2. **Represent selection**
3. (both in a visually and statistically sound way).

While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain types of plots and visualization styles. Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection. In fact, this may explain why stacking is preferred over dodging within the interactive data visualization community. Let's explore this idea in more depth by comparing linked selection with a stacked vs. dodged barplot.

```{r display}
#| echo: false
#| fig-height: 3.5
#| fig-cap: "\\label{fig:display}The visual coherence of stacking vs. dodging
#| during linked selection on the `mtcars` data set [@henderson1981]. Bars show
#| the numbers of cars in the 4-, 6-, and 8-cylinder categories. 
#| Dark segments represent selected cases, light segments represent unselected cases, 
#| and the plots' contour is shown via a light outline. Plots left to right show more cars being selected. 
#| Top row: the contour of a stacked barplot remains constant throughout selection
#| (three bars of the same height). Bottom row: the contour of a dodged barplot 
#| changes dramatically with selection. Note the changes to the upper y-axis limit 
#| and the absence of the lighter segment among the 6-cylinder cars
#| in the rightmost plot (since all 6-cylinder cars are selected, 
#| the darker segment has been stretched to fill the available space; 
#| `ggplot2` default for handling empty subcategories)."

library(ggplot2)

cols <- c('grey80', '#56B4E9', '#009E73')
mtcars$cyl <- factor(mtcars$cyl)
plots <- list()

p0 <- ggplot() +
  guides(fill = "none") +
  scale_fill_manual(values = c(cols[1], cols[3])) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(size = 12),
  )

set.seed(89719)
mtcars$group1 <- factor(sample(rep(c(0, 1), c(27, 5))))
mtcars$group2 <- factor(sample(rep(c(0, 1), c(17, 15))))
mtcars$group3 <- factor(sample(rep(c(0, 1), c(7, 25))))

for (i in 1:3) {
  mtcars$group <- mtcars[[paste0("group", i)]]
  plots[[i]] <- p0 + 
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
             col = cols[2]) +
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group)) +
    labs(title = ifelse(i == 1, "Stacking", ""))
  plots[[3 + i]] <- plot_spacer()
  plots[[6 + i]] <- p0 + 
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
             position = "dodge", col = cols[2]) +
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
             position = "dodge") +
    labs(title = ifelse(i == 1, "Dodging", ""))
}

wrap_plots(plots, nrow = 3, heights = c(1, 0.5, 1))
```

First, we propose that stacking is the superior method for *displaying* selection (see Figure \ref{fig:display}). In a stacked barplot, each stacked bar presents itself as a single visual entity, thanks to the four Gestalt principles: proximity, similarity, closure, and common region. Since the stacked segments are placed right on top of each other, and share the same width and closed border, they are perceived as part of a unified whole (the bar). This remains true throughout selection - no matter how the heights of the highlighted segments within a stacked bar change, the height of the whole bar remains constant, and so does the overall outline of the plot.

Dodged barplots lack this degree of visual consistency. In a dodged barplot, bar segments placed side-by-side in clusters, and, since each segment can be of different length, these clusters lack a shared border (common region). This leads to several design challenges. First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as more cases are added or removed. This creates distracting visual clutter that makes it difficult to track changes over time. Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), we need to decide how the handle the upper y-axis limit in response to selection. One option is to make the upper y-axis limit dynamic. If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the upper y-axis limit provides each time selection occurs. Alternatively, we can assign a static value to the upper y-axis limit, however, then we risk the bar segments growing outside of the bounds of the plotting region.

```{r selection, eval=FALSE}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:selection}Selection behavior in stacked vs. dodged barplot. Darker segments represent selected cases, dashed rectangle represent selection region created by clicking and dragging.
#| Left: in a stacked barplot, it is natural to select an entire category by clicking/dragging over a portion of a bar, regardless of the selection status, e.g. here the dashed selection rectangle would select the entire category A.
#| Right: in a dodged barplot, it is not clear what to expect from clicking and dragging over a segment - is the intention to select only the cases corresponding to the segment, or the cases corresponding to the whole bar? How does the plot change when the selection happens?"

library(ggplot2)
library(patchwork)

mtcars$am <- factor(mtcars$am)
mtcars$cyl <- factor(mtcars$cyl, labels = c("A", "B", "C"))

p0 <- ggplot() +
  scale_fill_manual(values = c("grey80", "indianred")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
)

p1 <- p0 + 
  geom_bar(data = mtcars, mapping = aes(cyl, fill = am)) +
  annotate("rect", xmin = 0.5, xmax = 1.25, ymin = 9.5, ymax = 11.5,
           fill = NA, col = "grey20", lty = "dashed", linewidth = 0.4)

p2 <- p0 + geom_bar(data = mtcars, mapping = aes(cyl, fill = am), 
                    position = "dodge") +
  annotate("rect", xmin = 0.75, xmax = 1.25, ymin = 7, ymax = 8.5,
           fill = NA, col = "grey20", lty = "dashed", linewidth = 0.4)

p1 + p2
```

Stacked barplots offer several other advantages over dodged bars within the context of linked selection. First, we propose that stacking also provides a simpler mental model for *selection*. Since each stacked bar is a single visual entity, it makes intuitive sense that clicking it will select the entire category corresponding to it. Indeed, this is how many interactive data visualization systems implement linked selection: clicking on part of an object selects the category corresponding to the entire object [see e.g. @cook2023; @vegalite2022; @vanderplas2018; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @swayne2003]. In dodged barplots, on the other hand, the expectation is less clear: when we click on a bar segment in a dodged barplot, will we select the entire category or just the cases corresponding to the segment? Second, in a stacked barplot, the bar width and inter-bar gap are constant throughout selection. In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty. For example, when no cases are selected, should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink these in response to selection? Finally, by presenting fewer visual entities on the screen, stacked bars may also reduce cognitive load [@sweller2019; for a non-technical overview, see @knaflic2015].

To summarize, stacked bars respond gracefully to selection. In contrast, dodging produces several undesirable behaviors when combined with selection: it produces jagged objects with non-uniform outline that changes dramatically with selection. These difference may explain why stacking is still popular in the interactive data visualization community.

However, a key question remains: when is stacking appropriate? So far, we have focused on the visual aspect of stacking. But stacking goes beyond aesthetics. This will be the subject of the next section.

# Statistics

## From Data to Representation

While data visualizations are made up of geometric objects such as points, lines, or areas, these objects alone are not enough to constitute data visualizations. A coloured rectangle holds no meaning in and of itself; a canvas filled with such rectangles is just an image (see Figure \ref{fig:magritte}). To turn an image of rectangles into a plot, we need to make sure that each rectangle accurately represents some underlying data.

```{r magritte}
#| echo: false
#| fig-height: 2.5
#| fig-show: "hold"
#| fig-align: "center"
#| fig-cap: "An image is not the thing it represents. Left: The Treachery of Images by René Magritte (1929) makes the point that an image of a thing is different from the thing itself. Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."

library(ggplot2)
library(patchwork)
library(figpatch)

mtcars$am <- factor(mtcars$am)

p1 <- fig("images/magritte2.jpg")

p2 <- ggplot(mtcars, aes(cyl, mpg, fill = am)) +
  geom_bar(stat = "summary", fun = "mean", width = 0.8) +
  annotate("text", 2.5, 60, 
           label = "This is not a barplot.", col = "#433A36ff",
           size = 5, family = "serif", fontface = "italic") +
  scale_y_continuous(limits = c(0, 70)) +
  scale_fill_manual(values = c("#815A38ff", "#433A36ff")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#E9DEC8ff"),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.margin = unit(rep(0, 4), units = "cm")
)

fig_wrap(list(p1, p2), ncol = 2)
```

However, we can rarely just "throw" data at the computer screen, in its raw form. Instead, we often first need to translate it into a format that the graphics device can understand. In most data visualization packages, this is done in two steps: aggregation and scaling.

First, the data needs to be aggregated or "wrangled" into sets of statistical summaries. These summaries may be very simple: for instance, in a typical scatterplot, we assign each point the raw values of the x- and y-variables (the summary function is then just identity, $f(x) = x$). More often, however, the data is grouped and aggregated in some way. For example, as was discussed previously, in a barplot, data points are grouped by the levels of the x-axis variable, and each group is then summarized by some descriptive statistic, such as count, sum, mean, or maximum. Similarly, in a typical histogram, the x-axis variable is divided into bins, and each bin is then summarized by the number of cases that fall inside it.

Second, the computed statistics have to be encoded into graphical attributes such as position, length, size, or colour [see e.g. @franconeri2021; @wilkinson2012; @pinker1990]. This is the job of specialized functions known as scales or coordinate systems [see e.g. @murrell2005; @murrell2007; @wickham2010; @wickham2016; @wilkinson2012]. For instance, in a typical scatterplot, the values of the x-axis variable are mapped to the x-position, such that points corresponding to small values of the x-axis variable are placed near the left-hand side of the plot (and larger values are place near the right-hand side). These mappings may be subject to non-linear transformations, such as log or square-root, and may also apply to discrete values such as barplot categories. Scales and coordinate systems are a rich subject and we have barely scratched the surface here; interested reader is advised to see @wilkinson2012, @ziemkiewicz2009, or @hotz2020.

## Combining Statistics

What does stacking mean in the context of the statistics we are trying to represent? To a novice, stacking may appear as a purely graphical operation. If they attempt to implement a data visualization pipeline from scratch, they may leave stacking to the very end, after the statistical summaries have already been translated into screen coordinates.

This view of stacking as a purely graphical operation has a certain appeal. Many modern data visualization systems follow the grammar-based model of visualization [@mcnutt2022; @kim2022; @vanderplas2020; @wickham2010; @satyanarayan2014; @satyanarayan2016; @wilkinson2012]. This model emphasizes compositionality, allowing plots to be built from independent, modular components, which the users can combine freely. For example, in the popular `ggplot2` library [@wickham2010], plots are built by combining components such as geometric objects (called `geoms`) and statistical summaries (`stats`). These grammar-based approach offers a great deal of expressiveness, and has gained widespread popularity as a result.

Thus, it may be tempting to treat stacking just as another modular component of a visualization, such that we may be able to stack any geometric objects and statistics we like. Unfortunately, this approach is deeply flawed. Many data visualization experts have issued explicit warnings about this:

>"Stacking is useful when the sum of the amounts represented by the individual stacked bars is in itself a meaningful amount" [@wilke2019, p. 52].

>"Because this gives the visual impression of one element that is the sum of several others, it is very important that if the element’s size is used to display a statistic, then that statistic must be summable. Stacking bars that represent counts, sums, or percentages are fine, but a stacked bar chart where bars show average values is generally meaningless." [@wills2011, p. 112].

>"[...] We do this to ensure that aggregate statistics are always computed over the input data, and so users do not inadvertantly compute e.g., averages of averages, which can easily lead to misinterpretation." [@wu2022]

The message is clear: stacking some statistics such as counts, sums, and percentages is acceptable, but stacking others is not. As outlined by Wills, while the sum of grouped counts produces a valid overall count, the sum of group means does not yield a meaningful statistic. Even averaging group means, as noted by @wu2022, does not produce a valid summary - the mean of group means differs from the grand mean (unless by coincidence). This problem is extends beyond barplots; for example, concerns about the statistical validity of stacked density plots have also been raised recently [@pu2020].

What exactly makes stacking sums acceptable but stacking other statistics such as means not? Wilke, Wills, and Wu et al. all argue that the aggregated value should be "meaningful". In the case of sums, perhaps another way to phrase this is: "the whole is equal to the sum of its parts". Or, more specifically:

> "the *sum* of the *sums* on the parts is equal to the *sum* of all values"

But what if we replace word "sum" with some placeholder (such as "foo"):

> "the *foo* of *foos* on the parts is equal to the *foo* of all values"

Then, we find there are other mathematical operators that also behave this way. For example, the product of products or the maximum of maximums are also valid overall statistics, see Figure \ref{fig:sumprodmax}. In fact, there is an entire class of mathematical objects with this exact behavior: monoids. We will explore monoids in the following section.

```{r sumprodmax}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:sumprodmax} Different operators can be used for stacking. The values `1.2`, `1.8`, `NA`, and `1.5` were assigned to a 2-by-2 table and stacked using three different operators. Left: stacking using the sum operator. Middle: stacking using the product operator (note that the y-axis starts at 1). Right: stacking using the max operator." 
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)

cols <- c('grey80', '#56B4E9', '#009E73')
label_size <- 10 / .pt

stack <- function(values, fn, initial) {
  values[is.na(values)] <- initial
  for (i in seq_along(values)) values[i] <- fn(values[i - 1], values[i])
  values
}

group <- factor(rep(c("A", "B"), each = 2))
subgroup <- factor(rep(1:2, 2))
value <- c(NA, 1.5, 1.2, 1.8)

sum_value <- rev(unlist(tapply(value, group, function(x) stack(x, sum, 0))))
prod_value <- rev(unlist(tapply(value, group, function(x) stack(x, prod, 1))))
max_value <- rev(unlist(tapply(value, group, function(x) stack(x, max, 0))))

or_na <- function(x) ifelse(is.na(x), "NA", x)
show_value <- or_na(value[c(3, 4, 1, 2)])

d1 <- data.frame(group, subgroup, value = value[4:1], 
                 sum_value, prod_value, max_value)

d2 <- d1
for (i in 4:6) {
  d2[1, i] <- (d2[1, i] + d2[2, i]) / 2
  d2[2, i] <- d2[2, i] / 2
  d2[3, i] <- (d2[3, i] + d2[4, i]) / 2
  d2[4, i] <- d2[4, i] / 2
}

d2[2, 5] <- 1 + (d1[2, 5] - 1) / 2

format_label <- function(value, symbol) {
  ifelse(is.na(value), value, paste0(symbol, value))
}

p0 <- ggplot(d1) +
  scale_fill_manual(values = c(cols[1], cols[3])) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 12, hjust = 0.1, vjust = -3),
    panel.border = element_blank(),
)

p1 <- p0 +
  geom_text(aes(group, subgroup, label = show_value), size = label_size) +
  geom_hline(aes(yintercept = 1.5), linewidth = 0.2) +
  geom_vline(aes(xintercept = 1.5), linewidth = 0.2) +
  scale_y_discrete(labels = c("Group 2", "Group 1")) +
  theme(panel.border = element_rect(fill = NA),
        axis.text = element_text(size = 12))

ran <- function(x) max(x) - min(x)
r <- sapply(d1[, -c(1:3)], ran)
ofs <- 0.04

p2 <- p0 +
  geom_bar(mapping = aes(group, sum_value, fill = subgroup), 
           stat = "identity", position = "identity") +
  geom_text(data = d2, aes(group, sum_value, label = format_label(value, "+")),
            size = label_size, col = "white") +
  scale_y_continuous(limits = c(0, 3.1)) +
  labs(title = "Sum")

p3 <- p0 +
  geom_bar(mapping = aes(group, prod_value, fill = subgroup), 
           stat = "identity", position = "identity") +
  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 0, ymax = 1), 
            fill = "white") + # Use to mask the bar starting from 0
  geom_text(data = d2, aes(group, prod_value, label = format_label(value, "x")),
             size = label_size, col = "white") +
  coord_cartesian(ylim = c(1, 2.25)) + 
  scale_y_continuous(breaks = c(1, 1.5, 2)) +
  labs(title = "Product")

p4 <- p0 +
  geom_bar(mapping = aes(group, max_value, fill = subgroup), 
           stat = "identity", position = "identity") +
  geom_text(data = d2, aes(group, max_value, label = format_label(value, "max\n")),
             size = label_size, col = "white", lineheight = 1) +
  scale_y_continuous(limits = c(0, 2), breaks = 0:2) +
  labs(title = "Maximum")

p2 + p3 + p4 + plot_layout(nrow = 1)
```

## Monoids \label{sec:monoids}

Monoids are a simple yet powerful concept that plays a key role in several areas of mathematics, including abstract algebra and category theory. The short treatment here draws primarily from @fong2019, @lawvere2009, @baez2023, and @milewski2018. For a particularly accessible introduction, readers are encouraged to seek out @fong2019 and @lawvere2009.

While the term "monoid" may sound intimidating, the concept is really quite simple. A monoid represents the idea of a "whole equal to the sum of its parts", if we relax our idea about what it means to "sum". More formally, a monoid is a tuple $(M, e, \otimes)$ consisting of:

a. A collection (set) of objects $M$
b. A neutral element $e \in M$ called the *monoidal unit*
c. A binary operation $\otimes: M \times M \to M$ called the *monoidal product*

Conforming to two rules:

1. *Unitality*: $x \otimes e = e \otimes x = x$
2. *Associativity*: $x \otimes (y \otimes z) = (x \otimes y) \otimes z = x \otimes y \otimes z$

(for all $x, y, z \in M$)

In plain words, when we have a monoid, we have some set of elements $m \in M$ and a way to combine them $\otimes$, such that the order in which we do the combining does not matter. Further, among the set of elements $M$, we also have a special element $e$, called the "neutral" or "identity" element, that, when combined with any other element $m \in M$, does nothing (and simply yields back $m$). This identity element is always unique [the proof is straightforward, for reference see e.g. @fong2019; @lawvere2009].

Let's return to the book example from Section \ref{sec:intro}. Is it a monoid? First of all, what is our set $M$? This could be the set of all book stacks we can form by combining books in our bookshelf. Second, do we have a neutral element? If we are a bit creative, we can imagine an "empty" stack of zero books. Let's use that as our identity element. Third, what is our binary operation? This is just the operation of taking one stack of books and putting it on top of another stack.

Now we just need to verify that our structure conforms to the two rules. First, is the operation associative? Clearly the answer is "yes", since a stack of detective novels will be the same height regardless of whether we put the novels by Agatha Christie on top of Jo Nesbø novels or vice versa. Second, is the operation unital? Again, this is self-evidently true, since adding zero books on top of a stack does not change its height. Therefore, we do have a monoid.

We can also describe the book example a bit more formally. Specifically, summation on natural numbers $(\mathbb{N}, 0, +)$ is a typical example of a monoid, because, as we have shown, it is associative and unital. Other examples of monoids include the previously mentioned products of real numbers $(\mathbb{R}, 1, \times)$, and the min and max operators $(\mathbb{R}, \infty, \min)$ and $(\mathbb{R}, -\infty, \max)$. As a counterexample, exponentiation is not a monoid, since it is not associative: $x^{(y^z)} \neq (x^y)^z$.

The definition of a monoid is quite broad however, and applies to more exotic structures than just operations on numbers. For example, the set of booleans $\mathbb{B} = \{ True, False \}$ equipped with either of the logical operators $\mathbf{AND}$ or $\mathbf{OR}$ is also a monoid, and so is the multiplication of $n \times n$ square matrices. Likewise, the operation of concatenating vectors and computing the euclidean distance is also a monoid [@stepanov2009]:

$$|| \big( || (x, y) ||_2, z \big) ||_2 = \sqrt{ \bigg(\sqrt{x^2 + y^2} \bigg)^2 + z^2} = \sqrt{x^2 + y^2 + z^2} = ||(x, y, z)||_2$$

Even very "non-number-like" operations such as concatenation of strings can be a monoid:

$$\texttt{"hello"} + \texttt{""} = \texttt{""} + \texttt{"hello"} = \texttt{"hello"}$$
$$(\texttt{"quick"} + \texttt{"brown"}) + \texttt{"fox"} = \texttt{"quick"} + (\texttt{"brown"} + \texttt{"fox"}) = \texttt{"quick brown fox"}$$

## Groups

Groups are another fundamental concept in mathematics, representing the idea of symmetry and reversible transformations. While introducing two algebraic concepts in quick succession may seem like a big ask, a group is really just a monoid with the additional property. Specifically, for all $x, y, z \in M$, we have:

3. *Inverse operator*: There exists $\otimes^{-1}$ such that if $x \otimes y = z$ then $z \otimes^{-1} x = y$

In simpler terms, a group embodies the idea that "the whole is equal to the sum of its parts, *in a non-destructive way*". That is, when we combine two elements, we can always recover either one by "subtracting" the contribution of the other. Of note, like the identity element for monoids, the inverse operator in a group is also always unique [see e.g. @fong2019; @lawvere2009].

Turning back to the book example, we can remove Agatha Christie novels off the detective novel stack and recover the height of the rest of the detective novels stack. Thus, summation on natural numbers is a group: $2 + 3 = 5 \implies 5 - 2 = 3$ (technically speaking, we now have to identify the underlying set as integers $\mathbb{Z}$). But, for example, the min and max operators lack an inverse and so do not constitute groups: if $\max{(x, 6)} = 6$, then there is no way to recover $x$ from the combined result ($6$) alone.

## Groups, Monoids, and Stacking

We now arrive at the core idea of the present paper. Suppose that we have some data set $D$, a way of summarizing this data via some associative binary operation $\otimes$, and a neutral value with respect to this operation $e$. In other words, $(D, \otimes, e)$ forms a monoid. Define $F(A)$ as the result of summarizing or "folding" some subset $A \subseteq D$ using the binary operation:

$$
F(A) = a_1 \otimes a_2 \otimes a_3 \ldots \otimes a_n
$$

Note that we could have written this as $F(A) = (a_1 \otimes (a_2 \otimes (a_3 \otimes ( \ldots \otimes a_n ))))$, however, since the operation is associative, we do not have to worry about brackets. Also, if $A$ is empty, then clearly:

$$F(\varnothing) = e$$

Now comes the most important idea: combining summaries of two disjoint subsets of the data set gives us the same result as summarizing the union of the original subsets:

$$
\begin{aligned}
F(A) \otimes F(B) &= (a_1 \otimes a_2 \otimes \ldots a_n) \otimes (b_1 \otimes b_2 \otimes \ldots b_n) \\
&= a_1 \otimes a_2 \otimes \ldots \otimes a_n \otimes b_1 \otimes b_2 \otimes \ldots \otimes b_n \qquad \text{(by associativity)} \\
&= F(A \cup B)
\end{aligned}
$$

Another way to show the same idea is the following commutative diagram:

$$
% https://q.uiver.app/#q=WzAsOCxbMSwwLCJBIFxcY3VwIEIiXSxbMCwxLCJBIl0sWzIsMSwiQSBcXGN1cCBCIFxcY3VwIEMiXSxbMSwzLCJGKEEgXFxjdXAgQikiXSxbMCw0LCJGKEEpIl0sWzIsNCwiRihBIFxcY3VwIEIgXFxjdXAgQykiXSxbMiw2XSxbMSw0XSxbMCwyLCItXFxjdXAgQyJdLFswLDMsIiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9fX1dLFsxLDIsInstIFxcY3VwIChCIFxcY3VwIEMpfSIsMl0sWzEsNCwiIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn19fV0sWzIsNSwiIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn19fV0sWzMsNSwiey0gXFxvdGltZXMgRihDfSJdLFs0LDUsInstIFxcb3RpbWVzIEYoQSBcXGN1cCBDKX0iLDJdLFsxLDAsIi0gXFxjdXAgQiJdLFs0LDMsIi0gXFxvdGltZXMgRihCKSJdXQ==
\begin{tikzcd}
	& A \\
	{A \cup B} && {A \cup B \cup C} \\
	& {F(A)} \\
	{F(A \cup B)} && {F(A \cup B \cup C)}
	\arrow["{- \cup B}"', from=1-2, to=2-1]
	\arrow["{- \cup (B \cup C)}", from=1-2, to=2-3]
	\arrow[dashed, from=1-2, to=3-2]
	\arrow["{- \cup C}"', near start, from=2-1, to=2-3]
	\arrow[dashed, from=2-1, to=4-1]
	\arrow[dashed, from=2-3, to=4-3]
	\arrow["{- \otimes F(B)}"', near start, from=3-2, to=4-1]
	\arrow["{- \otimes F(B \cup C)}", near start, from=3-2, to=4-3]
	\arrow["{- \otimes F(C)}"', from=4-1, to=4-3]
\end{tikzcd}
$$
The top and bottom triangles in the diagram both indicate associativity: in the top triangle, we can either first take the union of $A$ and $B$ and then union that with $C$, or immediately union $A$ with $B \cup C$ and the result will be identical. The same applies for the monoidal summary in the bottom triangle. By saying that the diagram commutes, we mean that any two parallel paths (the same start and endpoint) always produce the same result. Practically, this means we can either first summarize each set and then combine the summaries, or first union the sets and then summarize the union, or do some mix of the two, and we will always end up with a valid summary. As a sidenote, the diagram above actually corresponds to something called a functor or monoid homomorphism, however, that is beyond the scope of the present paper [interested reader should see @fong2019; @lawvere2009].

But this is precisely the idea of a stackable statistic. If our statistic conforms to the diagram, then the combined summary of subsets of the data is always a valid summary of the whole data set. In Wilke's words, we have discovered what makes the combined summary a "meaningful amount"!

It might seem that we are done. However, there is still one subtle issue left. Suppose we summarize three parts of the data $A$, $B$, and $C$, and combine these summaries by repeatedly applying the operation, as a kind of running total $F^*$:

$$
\begin{aligned}
F^* \big( F(A), F(B), F(C) \big) &= \big( F(A), F(A) \otimes F(B), F(A) \otimes F(B) \otimes F(C) \big) \\
&= \big( F(A), F(A \cup B), F(A \cup B \cup C) \big)
\end{aligned}
$$

Notice that we end up with summaries on nested subsets of the data, i.e. $F(A)$, $F(A \cup B)$, and $F(A \cup B \cup C)$. This is fine, for example, when we do linked selection with a single group and compare the summary on the highlighted subset ($A$) vs. the entirety the data ($A \cup B$).

However, what if we want to compare the subsets $A$ and $B$ directly? For example, suppose we do linked selection and with two highlighted groups ($A$ and $B$). Unfortunately, the properties of monoids do not guarantee that we will be able to recover $F(B)$ from $F(A \cup B)$ using $F(A)$. Put plainly, the monoidal operation may "collapse" the information contained in the disjoint parts, see Figure \ref{fig:summax}. To ensure that $F(B)$ is recoverable from $F(A \cup B)$, we also need the inverse operator $\otimes^{-1}$:

$$F(B) = F(A \cup B) \otimes^{-1} F(A)$$

Importantly, $F(A) \otimes F(B)$ is still a valid summary of $A \cup B$ even when the inverse does not exist. It is just that $F(A) \otimes F(B)$ may "forget" the information contained in either of its two parts. Note also that this is not a problem of different permutations of parts producing different sequences of accumulated summaries - this will happen regardless of whether or not there is an inverse.

```{r summax}
#| echo: false
#| fig-height: 5
#| fig-cap: "\\label{fig:summax} Monoids can \"collapse\" information about disjoint parts. Top: since sum is a group with an inverse (minus), we can recover the value of each disjoint part by subtracting the heights of the preceding segments (i.e. the part order does not matter). Bottom: since maximum lacks an inverse, we may lose information about the individual parts under certain permutations (note the absence of segments in the middle and rightmost plot)."
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)

cols <- c('grey80', '#56B4E9', '#009E73')
label_size <- 10 / .pt

stack <- function(values, fn, initial) {
  values[is.na(values)] <- initial
  for (i in seq_along(values)) values[i] <- fn(values[i - 1], values[i])
  values
}

two_mean <- function(x) (c(0, x[-length(x)]) + x) / 2

format_label <- function(value, symbol) {
  ifelse(is.na(value), value, paste0(symbol, value))
}

or_na <- function(x) ifelse(is.na(x), "NA", x)

group <- factor(rep(c("A", "B"), each = 3))
original_subgroups <- rep(3:1, 2)
original_values <- matrix(c(0.8, 1.2, 1.6, NA, 1.1, 1.5), ncol = 2)

plots <- list()
perms <- list(1:3, c(2, 1, 3), 3:1)

for (i in 1:3) {

  perm <- perms[[i]]
  values <- apply(original_values, 2, function(x) x[perm]) 
  subgroup <- original_subgroups[perm]

  value <- as.numeric(values)
  value[is.na(value)] <- 0

  sums <- apply(values, 2, function(x) stack(x, sum, 0))
  maximums <- apply(values, 2, function(x) stack(x, max, 0))
  sums_midpoint <- apply(sums, 2, two_mean)
  maximums_midpoint <- apply(maximums, 2, two_mean)

  sums_df <- data.frame(group, subgroup, value, 
                        midpoint = as.numeric(sums_midpoint),
                        cum_value = as.numeric(sums))

  maximums_df <- data.frame(group, subgroup, value, 
                            midpoint = as.numeric(maximums_midpoint),
                            cum_value = as.numeric(maximums))

  p0 <- ggplot() +
    scale_fill_manual(values = cols) +
    guides(fill = "none") +
    labs(x = NULL, y = NULL) +
    theme_bw() +
    theme(
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(size = 12, hjust = 0.1),
      panel.border = element_blank(),
      plot.margin = unit(c(0, 1, 0, 0), units = "cm")
  )

  dups <- duplicated(subset(maximums_df, select = c("cum_value")))

  # Need to rearrange for ggplot2 to stack the bars in the correct order  
  sums_df <- sums_df[c(3:1, 6:4), ]
  maximums_df <- maximums_df[c(3:1, 6:4), ]

  sums_df <- sums_df[-which(sums_df$value == 0), ]
  maximums_df <- maximums_df[-which(maximums_df$value == 0), ]

  # Remove rows with duplicate values  
  last_value <- -Inf
  for (j in rev(seq_len(nrow(maximums_df)))) {
    if (maximums_df$cum_value[j] == last_value) {
      maximums_df <- maximums_df[-j, ]
    }
    last_value <- maximums_df$cum_value[j]
  }

  plots[[3 * i - 2]] <- p0 +
    geom_bar(data = sums_df, aes(group, cum_value, 
                                 fill = factor(subgroup)), 
             stat = "identity", position = "identity") +
    geom_text(data = sums_df, aes(group, midpoint, 
                                  label = format_label(value, "+")),
              size = label_size, col = "white") +
    labs(title = ifelse(i == 1, "Sum", ""))

  plots[[3 * i - 1]] <- plot_spacer()

  plots[[3 * i]] <- p0 +
    geom_bar(data = maximums_df, 
             mapping = aes(group, cum_value, 
                           fill = factor(subgroup)), 
             stat = "identity", position = "identity") +
    geom_text(data = maximums_df, aes(group, midpoint, 
                                      label = format_label(value, "max\n")),
              size = label_size, col = "white", lineheight = 1) +
    labs(title = ifelse(i == 1, "Maximum", ""))
}

wrap_plots(plots, nrow = 3, byrow = FALSE,
           heights = c(1, 1/8, 1))
```

To summarize, whether our statistic is a monoid or a group determines what kinds of comparisons we can make on stacked objects:

- Monoids: comparing nested subsets, part vs. whole comparisons
- Groups: comparing disjoint subsets, part vs. part comparisons

This becomes important when implementing interactive data visualization features such as linked selection. Specifically, if we want to compare a single selected group vs. the rest of the data, then it's enough for our summary statistic to be a monoid. If we want to implement multiple selection groups, then our statistic needs to have an inverse as well. If we fail either of these conditions, then we may produce data representations which are either inaccurate (in the case of a non-monoidal statistic), or accurate but not "full" (in the case of a monoid but no inverse).

## Groups, Monoids, and Graphics

The connection between summary statistics and algebraic structures is important because it allows us to predict which new combinations of summaries and figures will "work". For example, while not generally thought of as a summary statistic, the convex hull algorithm [see e.g. @barber1996] provides a way to reduce a set of points to a simpler "summary", see Figure \ref{fig:convexhull}. We can recast the operation of finding the convex hull as a monoid: starting with an empty set, we iterate through the data, adding points if they lie outside of the current hull (and removing any points in the interior of the new hull). This may not be the most computationally efficient method for finding convex hulls, however, since it does meet the definition of a monoid, this tells us *what convex hulls can be used for*. If we combine two convex hulls, we know that the new hull is a valid summary for the union of the underlying data points. In turn, this allows us to implement interactive features such as single-group highlighting/selection, safe in the knowledge that the underlying structure is preserved.

However, does the operation of adding a point to a convex hull have an inverse? No. Once we add a point to the hull, we cannot recover the old hull by "subtracting" that point - we do not know how many points from the interior of the old hull we removed. This tells us that merging two convex hulls may irrevocably collapse information. As such, we cannot use convex hulls to display multiple selections, without resorting to some ad-hoc methods (e.g. keeping separate representation of the hulls and using partial transparency).

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:convexhull}Convex hull can be used as a valid representation of single-group selection but not multi-group selection. Both plots show data from Palmer penguins data set (Horst 2020). The light polygon shows the convex hull of the whole dataset, the dark polygon shows the hull of the Gentoo species, the medium-brightness polygon shows the combined hull of Gentoo and Chinstrap species, and the white dashed line shows the outline the hull of the Chinstrap penguins alone. Points show individual cases. Left: since convex hull is a monoid, we can use it to show single group selection/highlighting. Right: because of the lack of an inverse, we cannot guarantee that hull of the Chinstrap penguins (white-dashed line) will be recoverable from the combined hull of Gentoo and Chinstrap - we do not know how many interior points were \"forgotten\"." 
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(palmerpenguins)

penguins <- na.omit(penguins)[, c("species", "body_mass_g", "flipper_length_mm")]
chull_all <- penguins[with(penguins, chull(body_mass_g, flipper_length_mm)), ]

subset_chull <- function(subset) {
  cases <- penguins$species %in% subset
  penguins[which(cases)[chull(penguins[cases, -1])], ]
}

species <- unique(penguins$species)

chull1 <- subset_chull(c(species[2]))
chull2 <- subset_chull(c(species[3]))
chull3 <- subset_chull(c(species[2], species[3]))

p0 <- ggplot(data = penguins, aes(body_mass_g, flipper_length_mm)) +
  guides(fill = "none") +
  theme_bw() +
  guides(col = "none") +
  labs(x = "Body mass", y = "Flipper length") +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.margin = unit(c(0, 1, 0, 1), "cm")
)

p1 <- p0 + 
  geom_polygon(data = chull_all, fill = cols[1]) +
  geom_polygon(data = chull1, fill = cols[3]) +
  geom_point(col = "grey20", size = 0.25) +
  labs(title = "Single group: Valid")

p2 <- p0 + 
  geom_polygon(data = chull_all, fill = cols[1]) +
  geom_polygon(data = chull3, fill = cols[2]) +
  geom_polygon(data = chull1, fill = cols[3]) +
  geom_point(col = "grey20", size = 0.25) +
  geom_polygon(data = chull2, lty = "dashed", 
               col = "white", fill = NA) +
  labs(title = "Multiple groups: Invalid")

p1 + p2
```

## Groups, Monoids, and the Neutral Element \label{sec:advantages}

There are also other advantages that groups and monoids offer for visualizing data. One of these is the existence of the neutral element $e$ (see Section \ref{sec:monoids}). While seemingly inconspicuous, the neutral element can actually be quite useful.

Specifically, the neutral element ensures that we can always represent empty subsets of our data. Turning once more to the book example from Section \ref{sec:intro}, suppose we split the books into stacks based on the book's genre and the author's gender, and then summarize each stack by its total and average word-count. What if there are no books by male sci-fi authors? For total word-count, we have a meaningful default value - zero - as in, there are zero words across the zero books by male sci-fi authors. However, what about the average word-count? Here we run into trouble, since the mean of an empty set is not defined. It simply does not make sense to talk about the "average word-count across zero books".

A common approach is to omit the objects representing empty subsets. However, with non-monoidal summaries, this approach is flawed and leads ambiguity. Suppose we are drawing the bars of averages. If we omit bars representing the empty categories, then the missingness of a bar can indicate either of two things: either there are *no* cases in the category, or there *are* cases and their average is equal to the lower y-axis limit. There is just no way to tell from the graphic alone. In contrast, with monoidal summaries such as sums, the message of an object's absence is unambiguous. Regardless of whether there are zero or hundred cases in a category, if the bar representing them is missing, this means that their sum is zero.

The neutral element can also provide a new perspective on a perennial debate in data visualization: should the base of a barplot always start at zero? Some authors argue that starting bars from a non-zero base makes the lengths of the bars meaningless [@cleveland1985]. However, @wilkinson2012, for example, suggests that the bar base is really a statement about scales rather than about graphics, and as such there is no compulsion for starting it at zero. We offer the following new perspective: often, it might make sense for the base of the barplot to start at $e$, the neutral element, since this represents the default state of doing nothing. In other words, the bar base is tied to the properties of the *statistic*, rather than the just the graphic or scale.

In a barplot of sums or counts, this approach aligns with the traditional prescription, since the neutra element is zero. However, in a barplot of products, the neutral element is one. Thus, we may should our bars at one. One might object that this invalidates the y-axis as a ratio scale [@stevens1946], since a bar representing the value 4 will be no longer twice as tall as the bar representing the value 2. And indeed, if we start the base at one, the bar representing the value 2 will be only 2 - 1 = 1 units tall, whereas the bar representing the value 4 will be 4 - 1 = 3 units tall. However, we need to think about what the values represent. The bar representing the value 2 shows a 200% change or 100% increase; the bar representing 4 shows 400% change or 300% increase. Therefore, it makes sense that the bar representing 4 is three times taller. Also, this scale still fits the @stevens1946 definition of a ratio scale, since we can still determine equality of ratios: 1.5 is to 2 (100% / 50% = 2, double the increase) as 2 is to 3 (200% / 100% = 2, double the increase).

This example provides further evidence to support the claim that we cannot separate statistics from other data visualization components such as graphics and scales. Even before we apply any transformations, the method we have used to derive our summary statistics determines important properties of our visualization. We need to be mindful of this.

## Groups, Monoids, and Computation

Groups and monoids also come with some inherent computational advantages. Specifically, if a summary statistic is a monoid, associativity guarantees that it can be computed in a single pass through the data (i.e. $O(n)$ complexity), and in a distributed/parallel fashion [see e.g. @lammel2008; @lin2013]. This is why monoids are popular in functional programming, where they are used alongside the so-called `fold` or `reduce` functions [see e.g. @milewski2018; @abelson2022]. To be fair, computational efficiency is usually not the primary concern in data visualization: most data visualization systems are designed to work in a single-threaded environment, and rely on summary statistics that are fairly cheap to compute. Nevertheless, the guarantee of inherent efficiency can still be useful when visualizing larger data sets.

The computational advantages of groups and monoids become significant in interactive data visualization, however. Specifically, when designing an interactive data visualization, we may be able to exploit the fact that summaries on fine-grained subsets of the data "add up" to the summaries on coarse-grained subsets (their unions) to avoid unnecessary computation. Consider a linked histogram with reactive axes and interactive manipulation of binwidth and anchor. When the user engages in selection, the counts within the individual highlighted segments may change but the counts across the whole stacked bars will not. Thus, there is no need to recompute and update the upper y-axis limit - we know this will stay constant throughout selection. However, if the user interactively changes the binwidth or anchor, then both the counts within the whole bars *and* the highlighted segments may change and so we need to update the graphic appropriately.

In other words, associativity allows us to implement strategic updating. Fine-grained summaries (individual segments) need to be updated when the more coarse-grained summaries (whole stacked bars) change, but not vice versa. By pre-computing summaries on different levels of aggregation, we may be able to reduce computational effort.

In fact, if our summary statistic has an inverse, we can be even more computationally efficient. Suppose that one data point out of a thousand is removed from selection. If our summary has an inverse, we can use it to subtract the contribution of that single data point, instead of having to recompute the whole summary from scratch. This is, in fact, the principle behind the fast updates in the popular JavaScript library Crossfilter, which is unfortunately no longer actively maintained [@crossfilter2023; see also the Github discussion initiated by @crossfilter2024]^[Note that, as the present date, we are not aware of any explicit discussion of monoids or groups in the Crossfilter documentation].

However, while the inverse does bring advantages, it is also the case that fewer summary statistics have it. For instance, as was mentioned before, while maximum operator is a valid monoid, it lacks an inverse and as such does not meet the definition of a group. Once we summarize some data with the maximum, we cannot retrieve the second or third highest value - that information is gone forever. And there are many other operations (such as the convex hull in Figure \ref{fig:convexhull}) which lack an inverse. As such, when designing and interactive data visualization library, we need to carefully weigh the ability to support different kinds of summary statistics (general, monoids, groups) versus supporting different kinds of interactions (e.g. single-group selection vs. multi-group selection).

# Conclusion

We hope we have demonstrated that, when designing interactive graphics, algebraic thinking is invaluable. To support specific interactive features, the statistics underlying our plots must possess appropriate algebraic properties. For instance, if we want to display the result of splitting our data into parts and combining these parts back together, then we need either of two classes of mathematical structures. If we want to compare nested subsets of our data (for example, via single-group linked selection) then our summaries need to be monoids. If we want to compare disjoint parts of our data (multi-group linked selection), then our summaries need to be (algebraic) groups. The properties of these structures ensure that our figures are statistically sound, computationally efficient, and adhere to good design principles.

We have focused on unitality, associativity, and inverses, however, it may be useful to consider other algebraic properties as well. For instance, many of the visual attributes that we use to represent our data can only *grow* in one direction. Such is the case with, for example, width, height, and area (there is no such thing as "negative area"). If the visual attribute can only grow in one direction, then our summary statistic should be *monotonic*, such that if $x_1 \leq y_1$ and and $x_2 \leq y_2$ then $x_1 \otimes x_2 \leq y_1 \otimes y_2$. Similarly, in many types of plots, the order of the data points does not matter, however, in some, such as time-series plots, it does. Thus, it may be important to determine whether or not the statistics underlying our plots are *commutative*, such that $x \otimes y = y \otimes x$. Further exploration of these properties in relation to interactive graphics may yield useful insights, however, that is outside of the scope of the present paper.

We do not advocate for exclusively visualizing groups and monoids. Rather, we want to highlight the trade-off present in the choice of the summary statistic. Few summary statistics meet the definition of a group, but, if we choose such a statistic, then we can do a lot of things with it. With a monoid, our options are more limited, and with a non-monoidal statistic, even more so. Thus, like many past authors [such as @tufte2001; @tukey1977; @wilkinson2012], we simply want to point to the importance of thinking critically about the mathematical structure underlying our graphics. Geometric objects, statistics, and interaction are all deeply connected, and a full elaboration on this web of interdependence may require a lot more work. Groups and monoids merely allow us to carve out a small space of well-behaved graphics.

On a more practical note, we also need a framework for efficiently transforming raw data into summary statistics that can be displayed by interactive graphics. A decade ago, @wickham2009 noted the lack of such a formal pipeline (or "plumbing"), and, despite some attempts at implementing such pipelines [see e.g. @lawrence2009], that same sentiment was echoed again recently [@vanderplas2020]. Ideally, such a framework would minimize recomputation during user interaction. The computational "shortcuts" that groups and monoids provide may be invaluable here. We envision a visualization pipeline structured as a hierarchy of reactive partitions, where only the downstream (finer) partitions need to be recomputed when the upstream (coarser) partitions change, but not vice versa. We are currently exploring this concept in an R package called `plotscaper`^[`https://github.com/bartonicek/plotscaper`] (the project still has far to go to be truly generic).

Data visualization, in practice, is rarely about mathematical purity. Real world data is often flawed or incomplete, and it takes cunning and experience to dredge up valuable insights from it. Nevertheless, we believe that simple algebraic thinking offers a valuable foundation. Concepts like monoids and groups provide a rigorous framework for reasoning about graphics, interactive or otherwise. Ultimately, these concepts may empower us to build a broader class of robust interactive data visualizations.

# Disclosure Statement

The authors report there are no competing interests to declare.

# Bibliography

### 2024-10-19 paper.tex
Initial commit

--- 2024-10-19 paper.Rmd
@@
----
-title: "No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction"
-blinded: 0
-authors: 
-- name: Adam Bartonicek
-  affiliation: Department of Statistics, The University of Auckland
-- name: Simon Urbanek
-  affiliation: Department of Statistics, The University of Auckland
-- name: Paul Murrell
-  affiliation: Department of Statistics, The University of Auckland
-keywords:
-- Interactive data visualization, category theory, abstract algebra, the Gestalt Principles, descriptive statistics
-abstract: |
-  Interactive data visualization has become a staple of modern data presentation.
-Yet, despite its growing popularity, there still remains a lack of a formal framework for turning raw data into summary statistics that can then be displayed by interactive graphics.
-This gap may stem from a subtle yet profound issue: while we would often like treat statistics, graphics, and interaction as independent entities, they are in fact deeply connected.
-This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids.
-Specifically, if we want our graphics to support interactive features which break the data into multiple parts and then combine them back together (such as in the case of linked selection), then we need the statistics in our plots to fulfill certain algebraic properties.
-Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.
-bibliography: references.bib
-biblio-style: apalike
-nocite: |
-  @magritte1929
-  @horst2020
-output:
-  rticles::asa_article:
-    includes:
-      in_header: "preamble.tex"
-editor_options: 
-  chunk_output_type: console
-header-includes:
-  - \usepackage{tikz-cd}
----
-
-```{r}
-#| echo: false
-
-# knitr::opts_chunk$set(
-#   cache = TRUE
-# )
-
-```
-
-# Introduction \label{sec:intro}
-
-> "Effective graphical analysis makes things seem obvious, the effort involved in making the graphical analysis effective is not so obvious." [@unwin2018]
-
-The essence of data visualization is comparison.
-When we visualize, we split our data into parts, summarize each part one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and finally use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019].
-For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent the counts by drawing bars of the corresponding heights.
-In this way, each bar becomes a visual representation of its own "small data set", a subset of the original data.
-
-To ground this idea in a more concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other.
-You would construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on.
-Even from this very simple example, interesting mathematical properties jump out: for example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].
-
-Now, just as we can learn about the contents of our library by organizing books in the way we just described, we can learn about our data by splitting it into parts.
-A natural question then arises: can we learn more by subdividing our data into even smaller parts and comparing these?
-For instance, in our book example, we can split each stack of books into two smaller stacks based on the gender of the author.
-Each smaller stack will then represent the unique combination (Cartesian product) of the two categorical variables (genre and gender).
-Moreover, there arises a hierarchical relationship between the stacks: by combining two genre-gender stacks (child components) together, we can recover the original genre stack (parent component).
-Similar situation often arises in interactive graphics, where we use techniques such as linked selection to highlight specific subsets of our data.
-Either way, by forming these finer partitions, we can represent our data with greater granularity and discover trends that may otherwise hide in the aggregate.
-
-However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves the hierarchical nature of this relationship.
-In data visualization, there are two popular methods for doing this, known as "stacking" and "dodging" (see Figure \ref{fig:playfair}).
-These methods have existed at least since the time of William Playfair [-@playfair1801; -@playfair1822].
-Let's briefly illustrate them on the example of the classical barplot.
-In a stacked barplot, bar segments representing the product of the two factor variables are plotted vertically on top of each other ("stacked"; the operation can be also thought of as "highlighting" parts of the bar).
-In contrast, in a dodged barplot, bars are plotted side-by-side, as clusters of bars.
-Finally, there is also a third popular method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting (this technique may have been a bit challenging for William Playfair but has been made simple with computers).
-
-```{r playfair}
-#| echo: false
-#| fig-align: "center"
-#| fig-cap: "Examples of the three methods for dealing with bars that have subdivided into segments.
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+% !TeX program = pdfLaTeX
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx,psfrag,epsf}
+\usepackage{enumerate}
+\usepackage[]{natbib}
+\usepackage{textcomp}
+
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-1in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.7in}%
+\addtolength{\topmargin}{-1in}%
+
+%% load any required packages here
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{amssymb}
+\usepackage{tikz-cd}
+
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\hypersetup{
+  pdftitle={No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction},
+  pdfkeywords={Interactive data visualization, category theory, abstract
+algebra, the Gestalt Principles, descriptive statistics},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+
+\begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+{
+  \title{\bf No More, No Less Than Sum of its Parts: The Algebra of
+Graphics, Statistics, and Interaction}
+
+  \author{
+        Adam Bartonicek \\
+    Department of Statistics, The University of Auckland\\
+     and \\     Simon Urbanek \\
+    Department of Statistics, The University of Auckland\\
+     and \\     Paul Murrell \\
+    Department of Statistics, The University of Auckland\\
+      }
+  \maketitle
+} \fi
+
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf No More, No Less Than Sum of its Parts: The Algebra of
+Graphics, Statistics, and Interaction}
+  \end{center}
+  \medskip
+} \fi
+
+\bigskip
+\begin{abstract}
+Interactive data visualization has become a staple of modern data
+presentation. Yet, despite its growing popularity, there still remains a
+lack of a formal framework for turning raw data into summary statistics
+that can then be displayed by interactive graphics. This gap may stem
+from a subtle yet profound issue: while we would often like treat
+statistics, graphics, and interaction as independent entities, they are
+in fact deeply connected. This paper aims to shed light on this
+interdependence by relating it to two fundamental concepts from category
+theory: groups and monoids. Specifically, if we want our graphics to
+support interactive features which break the data into multiple parts
+and then combine them back together (such as in the case of linked
+selection), then we need the statistics in our plots to fulfill certain
+algebraic properties. Ultimately, understanding this hidden structure
+may empower data visualization experts to build more flexible and
+expressive interactive data visualization systems.
+\end{abstract}
+
+\noindent%
+{\it Keywords:} Interactive data visualization, category theory,
+abstract algebra, the Gestalt Principles, descriptive statistics
+
+\vfill
+
+\newpage
+\spacingset{1.9} % DON'T change the spacing!
+
+\section{\texorpdfstring{Introduction
+\label{sec:intro}}{Introduction }}\label{introduction}
+
+\begin{quote}
+``Effective graphical analysis makes things seem obvious, the effort
+involved in making the graphical analysis effective is not so obvious.''
+\citep{unwin2018}
+\end{quote}
+
+The essence of data visualization is comparison. When we visualize, we
+split our data into parts, summarize each part by a set of one or more
+statistics, and then encode these summaries into visual channels such as
+position, size, or color that we then use to draw geometric objects
+\citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
+recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
+example, when drawing a typical barplot, we split our data based on the
+levels of some categorical variable, count or sum up the values of some
+other variable, and finally represent these summaries by drawing bars of
+the corresponding heights. In this way, each bar becomes a visual
+representation of its own ``small data set'', a subset of the original
+data.
+
+To ground this idea in a more concrete example, one that will be used
+throughout the rest of this paper, imagine going through your bookshelf,
+sorting the books by genre, and stacking them on top of each other. You
+would construct a physical version of the barplot, forming one ``bar''
+(stack) per genre: one bar for sci-fi novels, one bar for romance, one
+bar for detective fiction, and so on. Even from this very simple
+example, interesting mathematical properties jump out: for example, a
+single book cannot go into two stacks (the subsets of the data are
+\emph{disjoint}) and if all of the stacks are combined back together, we
+should end up with the bookshelf we started with \citep[the subsets are
+\emph{surjective} and cover the entirety of the data set,
+see][]{ziemkiewicz2009}.
+
+Now, just as we can learn about the contents of our library by
+organizing books in the way we just described, we can learn about our
+data by splitting it into parts. A natural question that arises is: can
+we learn more by subdividing our data into even smaller parts and
+comparing these? For instance, in our book example, we could split each
+stack of books into two smaller stacks based on the gender of the
+author. Each smaller stack will then represent the unique combination
+(Cartesian product) of the two categorical variables (genre-gender).
+Moreover, there arises a hierarchical relationship between the stacks:
+by combining two genre-gender stacks (child components) together, we can
+recover the original genre stack (parent component). Similar situation
+often arises in interactive graphics, where we use techniques such as
+linked selection to highlight specific subsets of our data. Either way,
+by forming these finer partitions, we can represent our data with
+greater granularity and discover trends that may otherwise hide in the
+aggregate.
+
+However, once we have subdivided each category into its child
+components, we need a way of representing this information that visually
+preserves the hierarchical nature of the relationship. In data
+visualization, there are two popular methods for doing this, known as
+``stacking'' and ``dodging'' (see Figure \ref{fig:playfair}). These
+methods have existed at least since the time of William Playfair
+\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
+briefly illustrate them on the example of the classical barplot. In a
+stacked barplot, bar segments representing the product of the two factor
+variables are plotted vertically on top of each other (``stacked''; the
+operation can be also thought of as ``highlighting'' parts of the bar).
+In contrast, in a dodged barplot, bars are plotted side-by-side, as
+clusters of bars. Finally, there is also a third popular method, known
+as layering, where we plot bar segments in separate graphical layers and
+use partial transparency to mitigate overplotting (this technique may
+have been a bit challenging for William Playfair but has been made
+simple with computers).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/playfair-1}
+
+}
+
+\caption{Examples of the three methods for dealing with bars that have subdivided into segments.
 Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822).
-Bottom row: layering (created with ggplot2; Wickham, 2010)."
-#| message: false
-#| warning: false
-
-library(ggplot2)
-library(patchwork)
-library(magick)
-
-p1 <- image_read("./images/playfair1.jpg") |> image_ggplot()
-p2 <- image_read("./images/playfair2.jpg") |> image_ggplot()
-
-mpg$manufacturer <- tools::toTitleCase(mpg$manufacturer)
-
-p3 <- ggplot(mpg, aes(manufacturer, fill = factor(year))) +
-  geom_bar(position = "identity", alpha = 0.5) +
-  scale_x_discrete(labels = abbreviate) +
-  guides(fill = "none") +
-  labs(x = NULL, y = NULL) +
-  theme_bw() +
-  theme(
-    axis.ticks = element_blank(),
-    panel.grid = element_blank(),
-    panel.border = element_blank()
-)
-
-(p1 + p2) / p3
-```
-
-Much has been written about the relative merits of stacking, dodging, and layering.
-For example, layering is only useful with few categories, as blending many colors can make it difficult to tell the categories apart [@franconeri2021; @wilke2019].
-Further, in a landmark study, @cleveland1984 showed that people tend to be less accurate when reading information from stacked bar charts as opposed to dodged bar charts.
-Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below it, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984].
-Subsequent research has independently validated these findings and expanded upon them [see e.g.
-@heer2010; @thudt2016; @quadri2021].
-Ultimately, the suboptimal statistical legibility of stacking has led some authors to advise against its use altogether [@kosara2016; @wilke2019], or at least urge caution [see e.g.
-@byron2008; @cairo2014; @franconeri2021].
-
-However, despite the mixed reception towards stacking in static visualization, it remains a popular choice in interactive data visualization.
-Among popular interactive data visualization libraries, stacked plots such as bar charts and histograms make frequent appearance [see e.g.
-@cook2023; @murray2017; @vegalite2022; @vanderplas2018; @sievert2020; @swayne2003; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @wills2008; @ware2019].
-What is the reason for this?
-Has the interactive data visualization community simply failed to catch up with the current best practices?
-Or is there perhaps more going on?
-
-We argue that not only is stacking a useful technique, but that it can in fact offer valuable insights about the mathematical structure underlying our visualizations.
-To make the strongest case possible, we bring forward evidence from several levels of the visualization process.
-We start by discussing the properties of visual perception and how they apply to both static and interactive visualizations.
-Next, we explore the process of computing statistical summaries that form the basis of our plots.
-Finally, we dive into the algebraic properties of the functions underlying these statistical summaries.
-Throughout the text, we use the example of the humble barplot to ground our discussion.
-
-Our goal is to convince you that, in order to produce coherent and efficient interactive graphics, we simply cannot treat graphics, statistics, and interaction as independent components.
-Instead we need to consider them holistically, within the context of their respective visual and mathematical properties.
-Understanding this deep web of interdependence may in turn empower us to innovate and build new kinds of robust and expressive interactive visualizations.
-
-# Visual Perception and Interactivity
-
-## Visual perception and the Gestalt Principles
-
-Over the course of our evolutionary history, the human brain has evolved a sophisticated pipeline for processing visual information.
-This pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp.
-20].
-Many of the findings from this research are directly applicable to data visualization.
-For example, certain particularly salient visual stimuli are perceived automatically, rapidly (<10 ms), in parallel, and at a constant speed regardless of the number of distractors [@treisman1985].
-This phenomenon, known as pre-attentive processing, is key for data visualization, as the right use of visual attributes such as color, contrast, or shape can help us design figures which communicate insights much more effectively [@ware2019].
-
-However, many of the findings from visual perception research are quite specific; to guide effective visualization, we also need some general, high-level rules.
-One such set of high-level rules are the Gestalt principles of visual perception [see e.g.
-@cairo2012; @ware2019; @rosli2015; @vanderplas2020; @todorovic2008; @pinker1990].
-These principles formulate a set of fundamental "laws" related to the perception of patterns, primarily visual.
-They were discovered by the founders of the Gestalt school of psychology early in the 20th century, primarily by means of self-observation, and, despite not being grounded in sound understanding of the neural mechanisms of the brain, have nevertheless been later independently supported by findings from cognitive neuroscience [@guberman2017; @ware2019; @wertheimer1938].
-
-Among the Gestalt principles, there are a few which are particularly relevant to data visualization.
-First, there are the two principles of proximity and similarity.
-These principles state that objects that which appear in close together and/or similar in appearance tend to be perceived as part of single group.
-Second, the closure and common region principles state that regions with closed contours tend to be perceived as single, unified objects.
-Finally, the figure and ground principle states we tend to see one part of a scene as the central object (the "figure") while the rest is perceived as the background (the "ground").
-This can be modulated by other Gestalt principle: for example, smaller, darker, and symmetric objects with closed contours are more likely to be perceived as the figure [@ware2019].
-
-Whether we intend them to or not, the Gestalt principles affect how our data visualizations are perceived.
-As such, it is a good idea to try to use them to our advantage.
-Sometimes, little effort is needed.
-Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups by the observer.
-In other cases, however, our design choices can have a significant impact.
-For instance, another way to have objects to be perceived as part of a group is to draw a closed shape around them [@cairo2012].
-Alternatively, if we want to highlight a particular part of a data set, we can leverage the figure and ground principle and use attributes such as color and saturation to make it stand out [for specific examples, see @franconeri2021; @wilke2019, p.
-34].
-
-Finally, ignoring or violating the Gestalt principles can make our graphics less legible or even misleading.
-@tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements he termed "chartjunk".
-Similarly, @cairo2019 highlights many cases of graphics that violate the Gestalt principles.
-Finally, there is also the issue of well-designed graphics that communicate no statistical information at all, or even information that is blatantly false [@cairo2014; @franconeri2021; @cairo2019; @gelman2013; @tufte2001].
-Whether intentional or not, poor design can have a significant impact on the legibility and effectiveness of statistical graphics.
-
-## Gestalt and Interactivity
-
-The Gestalt principles apply to both static and interactive visualizations, however, the way they apply is different.
-Static data visualizations can be designed well or poorly, but this design quality does not change over time.
-In contrast, interactive visualizations change over time, with user interaction.
-As such, an interactive visualization may conform to the Gestalt principles at one instance of time and violate them in another.
-In fact, interactive visualizations can even differ in how well they leverage design principles throughout time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].
-
-Thus, interactivity inherently places further constraints the design of figures.
-If we want well-designed interactive figures, we need to ensure that they conform to good design principles across all states induced by user-interaction.
-This may not always be easy.
-We will investigate this idea in the following section, on the example of linked selection.
-
-## Case study: Linked Selection
-
-Linked selection, also known as linked brushing or linked highlighting, has been consistently ranked as one of the most useful interactive features in data visualization [see e.g.
-@buja1996; @heer2012; @wilhelm2003; @ware2019; @ward2015].
-By clicking or clicking-and-dragging over objects in one plot, users can highlight parts of objects corresponding to the selected cases across all other "linked" plots.
-This allows the user to rapidly "drill-down" [@dix1998; @theus2002] and explore trends across different dynamically-generated subsets of the data [similar to interactive filtering, see e.g.
-@heer2012].
-The ability to quickly materialize different views of the data makes linked selection a versatile tool for data exploration.
-
-Yet, despite its usefulness, linked selection also imposes some fundamental constraints on the figure. Specifically, all objects we visualize need to be able to:
-
-1. **Be selected**
-2. **Represent selection**
-3. (both in a visually and statistically sound way).
-
-While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain types of plots and visualization styles.
-Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection.
-In fact, this may explain why stacking is preferred over dodging within the interactive data visualization community.
-Let's explore this idea in more depth by comparing linked selection with a stacked vs.
-dodged barplot.
-
-```{r display}
-#| echo: false
-#| fig-height: 3.5
-#| fig-cap: "\\label{fig:display}The visual coherence of stacking vs. dodging
-#| during linked selection on the `mtcars` data set [@henderson1981]. Bars show
-#| the numbers of cars in the 4-, 6-, and 8-cylinder categories. 
-#| Dark segments represent selected cases, light segments represent unselected cases, 
-#| and the plots' contour is shown via a light outline. Plots left to right show more cars being selected. 
-#| Top row: the contour of a stacked barplot remains constant throughout selection
-#| (three bars of the same height). Bottom row: the contour of a dodged barplot 
-#| changes dramatically with selection. Note the changes to the upper y-axis limit 
-#| and the absence of the lighter segment among the 6-cylinder cars
-#| in the rightmost plot (since all 6-cylinder cars are selected, 
-#| the darker segment has been stretched to fill the available space; 
-#| `ggplot2` default for handling empty subcategories)."
-
-library(ggplot2)
-
-cols <- c('grey80', '#56B4E9', '#009E73')
-mtcars$cyl <- factor(mtcars$cyl)
-plots <- list()
-
-p0 <- ggplot() +
-  guides(fill = "none") +
-  scale_fill_manual(values = c(cols[1], cols[3])) +
-  labs(x = NULL, y = NULL) +
-  theme_bw() +
-  theme(
-    axis.ticks = element_blank(),
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    plot.title = element_text(size = 12),
-  )
-
-set.seed(89719)
-mtcars$group1 <- factor(sample(rep(c(0, 1), c(27, 5))))
-mtcars$group2 <- factor(sample(rep(c(0, 1), c(17, 15))))
-mtcars$group3 <- factor(sample(rep(c(0, 1), c(7, 25))))
-
-for (i in 1:3) {
-  mtcars$group <- mtcars[[paste0("group", i)]]
-  plots[[i]] <- p0 + 
-    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
-             col = cols[2]) +
-    geom_bar(data = mtcars, mapping = aes(cyl, fill = group)) +
-    labs(title = ifelse(i == 1, "Stacking", ""))
-  plots[[3 + i]] <- plot_spacer()
-  plots[[6 + i]] <- p0 + 
-    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
-             position = "dodge", col = cols[2]) +
-    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
-             position = "dodge") +
-    labs(title = ifelse(i == 1, "Dodging", ""))
+Bottom row: layering (created with ggplot2; Wickham, 2010).}\label{fig:playfair}
+\end{figure}
+
+Much has been written about the relative merits of stacking, dodging,
+and layering. For example, layering is only useful with few categories,
+as blending many colors can make it difficult to tell the categories
+apart \citep{franconeri2021, wilke2019}. Further, in a landmark study,
+\citet{cleveland1984} showed that people tend to be less accurate when
+reading information from stacked bar charts as opposed to dodged bar
+charts. Specifically, since the lower y-axis coordinate of a stacked
+segment is pushed up by the cumulative height of the segments below it,
+it becomes difficult to accurately compare segments' length, both within
+and across bars \citep{cleveland1984}. Subsequent research has
+independently validated and expanded upon these findings \citep[see
+e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
+statistical legibility of stacking has led some authors to advise
+against its use altogether \citep{kosara2016, wilke2019}, or at least
+urge caution \citep[see e.g.][]{byron2008, cairo2014, franconeri2021}.
+
+However, despite the mixed reception of stacked plots in static
+visualization, they remain a popular choice in interactive data
+visualization. Among popular interactive data visualization libraries,
+stacked plots such as bar charts and histograms make frequent appearance
+\citep[see
+e.g.][]{cook2023, murray2017, vegalite2022, vanderplas2018, sievert2020, swayne2003, xie2014, theus2002, urbanek2003, urbanek2011, wills2008, ware2019}.
+What is the reason for this? Has the interactive data visualization
+community simply failed to catch up with the current best practices? Or
+is there perhaps more going on?
+
+We argue that not only is stacking a useful technique, but that it can
+in fact offer valuable insights about the mathematical structure
+underlying our visualizations. To make the strongest case possible, we
+bring forward evidence from several levels of the visualization process.
+We start by discussing the properties of visual perception and how they
+apply to both static and interactive data visualizations. Next, we
+explore the process of computing statistical summaries that form the
+basis of our plots. Finally, we dive into the algebraic properties of
+the functions underlying these statistical summaries. Throughout the
+text, we use the example of the humble barplot to ground our discussion.
+
+Our goal is to convince you that, in order to produce coherent and
+efficient interactive graphics, we simply cannot treat graphics,
+statistics, and interaction as independent components. Instead we need
+to consider them holistically, within the context of their respective
+visual and mathematical properties. Understanding this deep web of
+interdependence may in turn empower us to innovate and build new kinds
+of robust interactive visualizations.
+
+\section{Visual Perception and
+Interactivity}\label{visual-perception-and-interactivity}
+
+\subsection{Visual perception and the Gestalt
+Principles}\label{visual-perception-and-the-gestalt-principles}
+
+Over the course of our evolutionary history, the human brain has evolved
+a sophisticated pipeline for processing visual information. This
+pipeline has been the subject of a great body of research in cognitive
+neuroscience
+\citetext{\citealp{goebel2004}; \citealp{knudsen2020}; \citealp[for a
+brief overview, see][20]{ware2019}}. Many of the findings from this
+research are directly applicable to data visualization. For example,
+certain particularly salient visual stimuli are perceived automatically,
+rapidly (\textless10 ms), in parallel, and at a constant speed
+regardless of the number of distractors \citep{treisman1985}. This
+phenomenon, known as pre-attentive processing, is key for data
+visualization, as the right use of visual attributes such as color,
+contrast, or shape can help us design figures which communicate insights
+much more effectively \citep{ware2019}.
+
+However, many of the findings from visual perception research are quite
+specific; to guide effective visualization, we also need some general,
+high-level rules. One such set of high-level rules are the Gestalt
+principles of visual perception \citep[see
+e.g.][]{cairo2012, ware2019, rosli2015, vanderplas2020, todorovic2008, pinker1990}.
+These principles formulate a set of fundamental ``laws'' related to the
+perception of patterns, primarily visual. They were discovered by the
+founders of the Gestalt school of psychology early in the 20th century,
+primarily by means of self-observation, and, despite not being grounded
+in sound understanding of the neural mechanisms of the brain, have
+nevertheless been later independently supported by findings from
+cognitive neuroscience \citep{guberman2017, ware2019, wertheimer1938}.
+
+Among the Gestalt principles, there are a few which are particularly
+relevant to data visualization. First, there are the two principles of
+proximity and similarity. These principles state that objects that which
+appear in close together and/or similar in appearance tend to be
+perceived as part of single group. Second, the closure and common region
+principles state that regions with closed contours tend to be perceived
+as single, unified objects. Finally, the figure and ground principle
+states we tend to see one part of a scene as the central object (the
+``figure'') while the rest is perceived as the background (the
+``ground''). This can be modulated by other Gestalt principle: for
+example, smaller, darker, and symmetric objects with closed contours are
+more likely to be perceived as the figure \citep{ware2019}.
+
+Whether we intend them to or not, the Gestalt principles how we perceive
+data visualizations. As such, it is a good idea to try to use them to
+our advantage. Sometimes, little effort is needed. Such is the case, for
+example, when identifying clusters in a scatterplot \citep{rosli2015}:
+the principles of proximity and similarity ensure that clouds of
+similarly shaped and colored points will be perceived as discrete groups
+by the observer. In other cases, however, our design choices can have a
+significant impact. For instance, another way to have objects to be
+perceived as part of a group is to draw a closed shape around them
+\citep{cairo2012}. Alternatively, if we want to highlight a particular
+part of a data set, we can leverage the figure and ground principle and
+use attributes such as color and saturation to make it stand out
+\citep[for specific examples, see][34]{franconeri2021, wilke2019}.
+
+Finally, ignoring or violating the Gestalt principles can make our
+graphics less legible or even misleading. \citet{tufte2001} had
+documented many such violations, such as deceptive 3D effects,
+inadvertent optical illusions caused by dense stripe patterns, and the
+overuse of superfluous graphical elements he termed ``chartjunk''.
+Similarly, \citet{cairo2019} highlights many cases of graphics that
+violate the Gestalt principles. Finally, there is also the issue of
+well-designed graphics that communicate no statistical information at
+all, or even information that is blatantly false
+\citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
+Either way, whether intentional or not, poor design can have a
+significant impact on the legibility and effectiveness of statistical
+graphics.
+
+\subsection{Gestalt and Interactivity}\label{gestalt-and-interactivity}
+
+The Gestalt principles apply to both static and interactive
+visualizations, however, the way they apply is different. Static data
+visualizations can be designed well or poorly, but this design quality
+does not change over time. In contrast, interactive visualizations
+change over time, with user interaction. As such, an interactive
+visualization may conform to the Gestalt principles at one instance of
+time and violate them in another. In fact, interactive visualizations
+can even differ in how well they leverage design principles throughout
+time: the animations that interactions induce may themselves conform to
+good or poor design \citep[see][]{hullman2013, ware2019}.
+
+Thus, interactivity inherently places further constraints the design of
+figures. If we want well-designed interactive figures, we need to ensure
+that they conform to good design principles across all states induced by
+user-interaction. This may not always be easy. We will investigate this
+idea in the following section, on the example of linked selection.
+
+\subsection{Case study: Linked
+Selection}\label{case-study-linked-selection}
+
+Linked selection, also known as linked brushing or linked highlighting,
+has been consistently ranked as one of the most useful interactive
+features in data visualization \citep[see
+e.g.][]{buja1996, heer2012, wilhelm2003, ware2019, ward2015}. By
+clicking or clicking-and-dragging over objects in one plot, users can
+highlight parts of objects corresponding to the selected cases across
+all other ``linked'' plots. This allows the user to rapidly
+``drill-down'' \citep{dix1998, theus2002} and explore trends across
+different dynamically-generated subsets of the data \citep[similar to
+interactive filtering, see e.g.][]{heer2012}. The ability to rapidly
+materialize different views of the data makes linked selection a
+particularly versatile tool for data exploration.
+
+Yet, despite its versatility, linked selection also imposes some
+fundamental constraints on the visualization. Specifically, all objects
+we visualize need to be able to:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Be selected}
+\item
+  \textbf{Represent selection}
+\item
+  (both in a visually and statistically sound way).
+\end{enumerate}
+
+While these constraints may not seem too restrictive, we argue that they
+pose significant challenges for certain types of plots and visualization
+styles. Specifically, some kinds of plots may produce undesirable visual
+behavior when combined with linked selection. In fact, this may explain
+why stacking is preferred over dodging within the interactive data
+visualization community. Let's explore this idea in more depth by
+comparing linked selection with a stacked vs.~dodged barplot.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/display-1.pdf}
+\caption{\label{fig:display}The visual coherence of stacking vs.~dodging
+during linked selection on the \texttt{mtcars} data set
+\citep{henderson1981}. Bars show the numbers of cars in the 4-, 6-, and
+8-cylinder categories. Dark segments represent selected cases, light
+segments represent unselected cases, and the plots' contour is shown via
+a light outline. Plots left to right show more cars being selected. Top
+row: the contour of a stacked barplot remains constant throughout
+selection (three bars of the same height). Bottom row: the contour of a
+dodged barplot changes dramatically with selection. Note the changes to
+the upper y-axis limit and the absence of the lighter segment among the
+6-cylinder cars in the rightmost plot (since all 6-cylinder cars are
+selected, the darker segment has been stretched to fill the available
+space; \texttt{ggplot2} default for handling empty subcategories).}
+\end{figure}
+
+First, we propose that stacking is the superior method for
+\emph{displaying} selection (see Figure \ref{fig:display}). In a stacked
+barplot, each stacked bar presents itself as a single visual entity,
+thanks to the four Gestalt principles: proximity, similarity, closure,
+and common region. Since the stacked segments are placed right on top of
+each other, and share the same width and closed border, they are
+perceived as part of a unified whole (the bar). This remains true
+throughout selection - no matter how the heights of the highlighted
+segments within a stacked bar change, the height of the whole bar
+remains constant, and so does the overall outline of the plot.
+
+Dodged barplots lack the degree of visual consistency found in stacked
+barplots. In a dodged barplot, the clusters of segments placed
+side-by-side lack a shared border (common region), leading to several
+design challenges. First, when selection occurs, the jagged contour of
+the dodged barplot can change dramatically as more cases are added or
+removed. This creates distracting visual clutter that makes it difficult
+to track changes over time. Second, selection may alter the height of
+the tallest bar segment in the plot. Since the height of the bar
+segments is not bound by the height of the tallest bar (as is the case
+in stacked barplot), this presents us with a dilemma. Either either we
+make the upper y-axis limit dynamic, meaning that the bar segments will
+always fit the plotting region, but then the context that the upper
+y-axis limit provides is lost with each time selection happens.
+Alternatively, we keep the upper y-axis limit static, but then we risk
+the bar segments growing outside of the bounds of the plotting region.
+
+Stacked barplots offer several other advantages over dodged bars within
+the context of linked selection. First, we propose that stacking also
+provides a simpler mental model for \emph{selection}. Since each stacked
+bar is a single visual entity, it makes intuitive sense that clicking it
+will select the entire category corresponding to it. Indeed, this is how
+many interactive data visualization systems implement linked selection:
+clicking on part of an object selects the category corresponding to the
+entire object \citep[see
+e.g.][]{cook2023, vegalite2022, vanderplas2018, xie2014, theus2002, urbanek2003, urbanek2011, swayne2003}.
+In dodged barplots, on the other hand, the expectation is less clear:
+when we click on a bar segment in a dodged barplot, will we select the
+entire category or just the cases corresponding to the segment? Second,
+in a stacked barplot, the bar width and inter-bar gap are constant
+throughout selection. In a dodged barplot, it is not clear how to handle
+the gaps when some selection groups are empty. For example, when no
+cases are selected, should we leave gaps for the (currently absent)
+highlighted segments, or draw wide bars and only shrink these in
+response to selection? Finally, by presenting fewer visual entities on
+the screen, stacked bars may also reduce cognitive load
+\citetext{\citealp{sweller2019}; \citealp[for a non-technical overview,
+see][]{knaflic2015}}.
+
+To summarize, stacked bars respond gracefully to selection. In contrast,
+dodging produces several undesirable behaviors when combined with
+selection: it produces jagged objects with non-uniform outline that
+changes dramatically with selection. These difference may explain why
+stacking is still popular in the interactive data visualization
+community.
+
+However, a key question remains: when is stacking appropriate? So far,
+we have focused on the visual aspect of stacking. But stacking goes
+beyond aesthetics. This will be the subject of the next section.
+
+\section{Statistics}\label{statistics}
+
+\subsection{From Data to
+Representation}\label{from-data-to-representation}
+
+While data visualizations are made up of geometric objects such as
+points, lines, or areas, these objects alone are not enough to
+constitute data visualizations. A coloured rectangle holds no meaning in
+and of itself; a canvas filled with such rectangles is just an image
+(see Figure \ref{fig:magritte}). To turn an image of rectangles into a
+plot, we need to make sure that each rectangle accurately represents
+some underlying data.
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/magritte-1}
+
 }
 
-wrap_plots(plots, nrow = 3, heights = c(1, 0.5, 1))
-```
-
-First, we propose that stacking is the superior method for *displaying* selection (see Figure \ref{fig:display}).
-In a stacked barplot, each stacked bar presents itself as a single visual entity, thanks to the four Gestalt principles: proximity, similarity, closure, and common region.
-Since the stacked segments are placed right on top of each other, and share the same width and closed border, they are perceived as part of a unified whole (the bar).
-This remains true throughout selection - no matter how the heights of the highlighted segments within a stacked bar change, the height of the whole bar remains constant, and so does the overall outline of the plot.
-
-Dodged barplots lack this degree of visual consistency.
-In a dodged barplot, bar segments placed side-by-side in clusters, and, since each segment can be of different length, these clusters lack a shared border (common region).
-This leads to several design challenges.
-First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as more cases are added or removed.
-This creates distracting visual clutter that makes it difficult to track changes over time.
-Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), we need to decide how the handle the upper y-axis limit in response to selection.
-One option is to make the upper y-axis limit dynamic.
-If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the upper y-axis limit provides each time selection occurs.
-Alternatively, we can assign a static value to the upper y-axis limit, however, then we risk the bar segments growing outside of the bounds of the plotting region.
-
-```{r selection, eval=FALSE}
-#| echo: false
-#| fig-height: 2.5
-#| fig-cap: "\\label{fig:selection}Selection behavior in stacked vs.
-dodged barplot.
-Darker segments represent selected cases, dashed rectangle represent selection region created by clicking and dragging.
-#| Left: in a stacked barplot, it is natural to select an entire category by clicking/dragging over a portion of a bar, regardless of the selection status, e.g.
-here the dashed selection rectangle would select the entire category A.
-#| Right: in a dodged barplot, it is not clear what to expect from clicking and dragging over a segment - is the intention to select only the cases corresponding to the segment, or the cases corresponding to the whole bar?
-How does the plot change when the selection happens?"
-
-library(ggplot2)
-library(patchwork)
-
-mtcars$am <- factor(mtcars$am)
-mtcars$cyl <- factor(mtcars$cyl, labels = c("A", "B", "C"))
-
-p0 <- ggplot() +
-  scale_fill_manual(values = c("grey80", "indianred")) +
-  guides(fill = "none") +
-  labs(x = NULL, y = NULL) +
-  theme_bw() +
-  theme(
-    axis.ticks = element_blank(),
-    panel.grid = element_blank(),
-    panel.border = element_blank()
-)
-
-p1 <- p0 + 
-  geom_bar(data = mtcars, mapping = aes(cyl, fill = am)) +
-  annotate("rect", xmin = 0.5, xmax = 1.25, ymin = 9.5, ymax = 11.5,
-           fill = NA, col = "grey20", lty = "dashed", linewidth = 0.4)
-
-p2 <- p0 + geom_bar(data = mtcars, mapping = aes(cyl, fill = am), 
-                    position = "dodge") +
-  annotate("rect", xmin = 0.75, xmax = 1.25, ymin = 7, ymax = 8.5,
-           fill = NA, col = "grey20", lty = "dashed", linewidth = 0.4)
-
-p1 + p2
-```
-
-Stacked barplots offer several other advantages over dodged bars within the context of linked selection.
-First, we propose that stacking also provides a simpler mental model for *selection*.
-Since each stacked bar is a single visual entity, it makes intuitive sense that clicking it will select the entire category corresponding to it.
-Indeed, this is how many interactive data visualization systems implement linked selection: clicking on part of an object selects the category corresponding to the entire object [see e.g.
-@cook2023; @vegalite2022; @vanderplas2018; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @swayne2003].
-In dodged barplots, on the other hand, the expectation is less clear: when we click on a bar segment in a dodged barplot, will we select the entire category or just the cases corresponding to the segment?
-Second, in a stacked barplot, the bar width and inter-bar gap are constant throughout selection.
-In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty.
-For example, when no cases are selected, should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink these in response to selection?
-Finally, by presenting fewer visual entities on the screen, stacked bars may also reduce cognitive load [@sweller2019; for a non-technical overview, see @knaflic2015].
-
-To summarize, stacked bars respond gracefully to selection.
-In contrast, dodging produces several undesirable behaviors when combined with selection: it produces jagged objects with non-uniform outline that changes dramatically with selection.
-These difference may explain why stacking is still popular in the interactive data visualization community.
-
-However, a key question remains: when is stacking appropriate?
-So far, we have focused on the visual aspect of stacking.
-But stacking goes beyond aesthetics.
-This will be the subject of the next section.
-
-# Statistics
-
-## From Data to Representation
-
-While data visualizations are made up of geometric objects such as points, lines, or areas, these objects alone are not enough to constitute data visualizations.
-A coloured rectangle holds no meaning in and of itself; a canvas filled with such rectangles is just an image (see Figure \ref{fig:magritte}).
-To turn an image of rectangles into a plot, we need to make sure that each rectangle accurately represents some underlying data.
-
-```{r magritte}
-#| echo: false
-#| fig-height: 2.5
-#| fig-show: "hold"
-#| fig-align: "center"
-#| fig-cap: "An image is not the thing it represents.
+\caption{An image is not the thing it represents.
 Left: The Treachery of Images by René Magritte (1929) makes the point that an image of a thing is different from the thing itself.
-Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."
-
-library(ggplot2)
-library(patchwork)
-library(figpatch)
-
-mtcars$am <- factor(mtcars$am)
-
-p1 <- fig("images/magritte2.jpg")
-
-p2 <- ggplot(mtcars, aes(cyl, mpg, fill = am)) +
-  geom_bar(stat = "summary", fun = "mean", width = 0.8) +
-  annotate("text", 2.5, 60, 
-           label = "This is not a barplot.", col = "#433A36ff",
-           size = 5, family = "serif", fontface = "italic") +
-  scale_y_continuous(limits = c(0, 70)) +
-  scale_fill_manual(values = c("#815A38ff", "#433A36ff")) +
-  guides(fill = "none") +
-  labs(x = NULL, y = NULL) +
-  theme(
-    panel.background = element_rect(fill = "#E9DEC8ff"),
-    axis.ticks = element_blank(),
-    axis.text = element_blank(),
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    plot.margin = unit(rep(0, 4), units = "cm")
-)
-
-fig_wrap(list(p1, p2), ncol = 2)
-```
-
-However, we can rarely just "throw" data at the computer screen, in its raw form.
-Instead, we often first need to translate it into a format that the graphics device can understand.
-In most data visualization packages, this is done in two steps: aggregation and scaling.
-
-First, the data needs to be aggregated or "wrangled" into sets of statistical summaries.
-These summaries may be very simple: for instance, in a typical scatterplot, we assign each point the raw values of the x- and y-variables (the summary function is then just identity, $f(x) = x$).
-More often, however, the data is grouped and aggregated in some way.
-For example, as was discussed previously, in a barplot, data points are grouped by the levels of the x-axis variable, and each group is then summarized by some descriptive statistic, such as count, sum, mean, or maximum.
-Similarly, in a typical histogram, the x-axis variable is divided into bins, and each bin is then summarized by the number of cases that fall inside it.
-
-Second, the computed statistics have to be encoded into graphical attributes such as position, length, size, or colour [see e.g.
-@franconeri2021; @wilkinson2012; @pinker1990].
-This is the job of specialized functions known as scales or coordinate systems [see e.g.
-@murrell2005; @murrell2007; @wickham2010; @wickham2016; @wilkinson2012].
-For instance, in a typical scatterplot, the values of the x-axis variable are mapped to the x-position, such that points corresponding to small values of the x-axis variable are placed near the left-hand side of the plot (and larger values are place near the right-hand side).
-These mappings may be subject to non-linear transformations, such as log or square-root, and may also apply to discrete values such as barplot categories.
-Scales and coordinate systems are a rich subject and we have barely scratched the surface here; interested reader is advised to see @wilkinson2012, @ziemkiewicz2009, or @hotz2020.
-
-## Combining Statistics
-
-What does stacking mean in the context of the statistics we are trying to represent?
-To a novice, stacking may appear as a purely graphical operation.
-If they attempt to implement a data visualization pipeline from scratch, they may leave stacking to the very end, after the statistical summaries have already been translated into screen coordinates.
-
-This view of stacking as a purely graphical operation has a certain appeal.
-Many modern data visualization systems follow the grammar-based model of visualization [@mcnutt2022; @kim2022; @vanderplas2020; @wickham2010; @satyanarayan2014; @satyanarayan2016; @wilkinson2012].
-This model emphasizes compositionality, allowing plots to be built from independent, modular components, which the users can combine freely.
-For example, in the popular `ggplot2` library [@wickham2010], plots are built by combining components such as geometric objects (called `geoms`) and statistical summaries (`stats`).
-These grammar-based approach offers a great deal of expressiveness, and has gained widespread popularity as a result.
-
-Thus, it may be tempting to treat stacking just as another modular component of a visualization, such that we may be able to stack any geometric objects and statistics we like.
-Unfortunately, this approach is deeply flawed.
-Many data visualization experts have issued explicit warnings about this:
-
->"Stacking is useful when the sum of the amounts represented by the individual stacked bars is in itself a meaningful amount" [@wilke2019, p. 52].
-
->"Because this gives the visual impression of one element that is the sum of several others, it is very important that if the element’s size is used to display a statistic, then that statistic must be summable.
-Stacking bars that represent counts, sums, or percentages are fine, but a stacked bar chart where bars show average values is generally meaningless." [@wills2011, p.
-112].
-
->"[...] We do this to ensure that aggregate statistics are always computed over the input data, and so users do not inadvertantly compute e.g., averages of averages, which can easily lead to misinterpretation." [@wu2022]
-
-The message is clear: stacking some statistics such as counts, sums, and percentages is acceptable, but stacking others is not.
-As outlined by Wills, while the sum of grouped counts produces a valid overall count, the sum of group means does not yield a meaningful statistic.
-Even averaging group means, as noted by @wu2022, does not produce a valid summary - the mean of group means differs from the grand mean (unless by coincidence).
-This problem is extends beyond barplots; for example, concerns about the statistical validity of stacked density plots have also been raised recently [@pu2020].
-
-What exactly makes stacking sums acceptable but stacking other statistics such as means not?
-Wilke, Wills, and Wu et al.
-all argue that the aggregated value should be "meaningful".
-In the case of sums, perhaps another way to phrase this is: "the whole is equal to the sum of its parts".
-Or, more specifically:
-
-> "the *sum* of the *sums* on the parts is equal to the *sum* of all values"
-
-But what if we replace word "sum" with some placeholder (such as "foo"):
-
-> "the *foo* of *foos* on the parts is equal to the *foo* of all values"
-
-Then, we find there are other mathematical operators that also behave this way.
-For example, the product of products or the maximum of maximums are also valid overall statistics, see Figure \ref{fig:sumprodmax}.
-In fact, there is an entire class of mathematical objects with this exact behavior: monoids.
-We will explore monoids in the following section.
-
-```{r sumprodmax}
-#| echo: false
-#| fig-height: 2.5
-#| fig-cap: "\\label{fig:sumprodmax} Different operators can be used for stacking.
-The values `1.2`, `1.8`, `NA`, and `1.5` were assigned to a 2-by-2 table and stacked using three different operators.
-Left: stacking using the sum operator.
-Middle: stacking using the product operator (note that the y-axis starts at 1).
-Right: stacking using the max operator."
-#| message: false
-#| warning: false
-
-library(ggplot2)
-library(patchwork)
-
-cols <- c('grey80', '#56B4E9', '#009E73')
-label_size <- 10 / .pt
-
-stack <- function(values, fn, initial) {
-  values[is.na(values)] <- initial
-  for (i in seq_along(values)) values[i] <- fn(values[i - 1], values[i])
-  values
-}
-
-group <- factor(rep(c("A", "B"), each = 2))
-subgroup <- factor(rep(1:2, 2))
-value <- c(NA, 1.5, 1.2, 1.8)
-
-sum_value <- rev(unlist(tapply(value, group, function(x) stack(x, sum, 0))))
-prod_value <- rev(unlist(tapply(value, group, function(x) stack(x, prod, 1))))
-max_value <- rev(unlist(tapply(value, group, function(x) stack(x, max, 0))))
-
-or_na <- function(x) ifelse(is.na(x), "NA", x)
-show_value <- or_na(value[c(3, 4, 1, 2)])
-
-d1 <- data.frame(group, subgroup, value = value[4:1], 
-                 sum_value, prod_value, max_value)
-
-d2 <- d1
-for (i in 4:6) {
-  d2[1, i] <- (d2[1, i] + d2[2, i]) / 2
-  d2[2, i] <- d2[2, i] / 2
-  d2[3, i] <- (d2[3, i] + d2[4, i]) / 2
-  d2[4, i] <- d2[4, i] / 2
-}
-
-d2[2, 5] <- 1 + (d1[2, 5] - 1) / 2
-
-format_label <- function(value, symbol) {
-  ifelse(is.na(value), value, paste0(symbol, value))
-}
-
-p0 <- ggplot(d1) +
-  scale_fill_manual(values = c(cols[1], cols[3])) +
-  guides(fill = "none") +
-  labs(x = NULL, y = NULL) +
-  theme_bw() +
-  theme(
-    axis.ticks = element_blank(),
-    panel.grid = element_blank(),
-    plot.title = element_text(size = 12, hjust = 0.1, vjust = -3),
-    panel.border = element_blank(),
-)
-
-p1 <- p0 +
-  geom_text(aes(group, subgroup, label = show_value), size = label_size) +
-  geom_hline(aes(yintercept = 1.5), linewidth = 0.2) +
-  geom_vline(aes(xintercept = 1.5), linewidth = 0.2) +
-  scale_y_discrete(labels = c("Group 2", "Group 1")) +
-  theme(panel.border = element_rect(fill = NA),
-        axis.text = element_text(size = 12))
-
-ran <- function(x) max(x) - min(x)
-r <- sapply(d1[, -c(1:3)], ran)
-ofs <- 0.04
-
-p2 <- p0 +
-  geom_bar(mapping = aes(group, sum_value, fill = subgroup), 
-           stat = "identity", position = "identity") +
-  geom_text(data = d2, aes(group, sum_value, label = format_label(value, "+")),
-            size = label_size, col = "white") +
-  scale_y_continuous(limits = c(0, 3.1)) +
-  labs(title = "Sum")
-
-p3 <- p0 +
-  geom_bar(mapping = aes(group, prod_value, fill = subgroup), 
-           stat = "identity", position = "identity") +
-  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 0, ymax = 1), 
-            fill = "white") + # Use to mask the bar starting from 0
-  geom_text(data = d2, aes(group, prod_value, label = format_label(value, "x")),
-             size = label_size, col = "white") +
-  coord_cartesian(ylim = c(1, 2.25)) + 
-  scale_y_continuous(breaks = c(1, 1.5, 2)) +
-  labs(title = "Product")
-
-p4 <- p0 +
-  geom_bar(mapping = aes(group, max_value, fill = subgroup), 
-           stat = "identity", position = "identity") +
-  geom_text(data = d2, aes(group, max_value, label = format_label(value, "max\n")),
-             size = label_size, col = "white", lineheight = 1) +
-  scale_y_continuous(limits = c(0, 2), breaks = 0:2) +
-  labs(title = "Maximum")
-
-p2 + p3 + p4 + plot_layout(nrow = 1)
-```
-
-## Monoids \label{sec:monoids}
-
-Monoids are a simple yet powerful concept that plays a key role in several areas of mathematics, including abstract algebra and category theory.
-The short treatment here draws primarily from @fong2019, @lawvere2009, @baez2023, and @milewski2018.
-For a particularly accessible introduction, readers are encouraged to seek out @fong2019 and @lawvere2009.
-
-While the term "monoid" may sound intimidating, the concept is really quite simple.
-A monoid represents the idea of a "whole equal to the sum of its parts", if we relax our idea about what it means to "sum".
-More formally, a monoid is a tuple $(M, e, \otimes)$ consisting of:
-
-a. A collection (set) of objects $M$
-b. A neutral element $e \in M$ called the *monoidal unit*
-c. A binary operation $\otimes: M \times M \to M$ called the *monoidal product*
+Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
+\end{figure}
+
+However, we can rarely just ``throw'' data at the computer screen, in
+its raw form. Instead, we often first need to translate it into a format
+that the graphics device can understand. In most data visualization
+packages, this is done in two steps: aggregation and scaling.
+
+First, the data needs to be aggregated or ``wrangled'' into sets of
+statistical summaries. These summaries may be very simple: for instance,
+in a typical scatterplot, we assign each point the raw values of the x-
+and y-variables (the summary function is then just identity,
+\(f(x) = x\)). More often, however, the data is grouped and aggregated
+in some way. For example, as was discussed previously, in a barplot,
+data points are grouped by the levels of the x-axis variable, and each
+group is then summarized by some descriptive statistic, such as count,
+sum, mean, or maximum. Similarly, in a typical histogram, the x-axis
+variable is divided into bins, and each bin is then summarized by the
+number of cases that fall inside it.
+
+Second, the computed statistics have to be encoded into graphical
+attributes such as position, length, size, or colour \citep[see
+e.g.][]{franconeri2021, wilkinson2012, pinker1990}. This is the job of
+specialized functions known as scales or coordinate systems \citep[see
+e.g.][]{murrell2005, murrell2007, wickham2010, wickham2016, wilkinson2012}.
+For instance, in a typical scatterplot, the values of the x-axis
+variable are mapped to the x-position, such that points corresponding to
+small values of the x-axis variable are placed near the left-hand side
+of the plot (and larger values are place near the right-hand side).
+These mappings may be subject to non-linear transformations, such as log
+or square-root, and may also apply to discrete values such as barplot
+categories. Scales and coordinate systems are a rich subject and we have
+barely scratched the surface here; interested reader is advised to see
+\citet{wilkinson2012}, \citet{ziemkiewicz2009}, or \citet{hotz2020}.
+
+\subsection{Combining Statistics}\label{combining-statistics}
+
+What does stacking mean in the context of the statistics we are trying
+to represent? To a novice, stacking may appear as a purely graphical
+operation. If they attempt to implement a data visualization pipeline
+from scratch, they may leave stacking to the very end, after the
+statistical summaries have already been translated into screen
+coordinates.
+
+This view of stacking as a purely graphical operation has a certain
+appeal. Many modern data visualization systems follow the grammar-based
+model of visualization
+\citep{mcnutt2022, kim2022, vanderplas2020, wickham2010, satyanarayan2014, satyanarayan2016, wilkinson2012}.
+This model emphasizes compositionality, allowing plots to be built from
+independent, modular components, which the users can combine freely. For
+example, in the popular \texttt{ggplot2} library \citep{wickham2010},
+plots are built by combining components such as geometric objects
+(called \texttt{geoms}) and statistical summaries (\texttt{stats}).
+These grammar-based approach offers a great deal of expressiveness, and
+has gained widespread popularity as a result.
+
+Thus, it may be tempting to treat stacking just as another modular
+component of a visualization, such that we may be able to stack any
+geometric objects and statistics we like. Unfortunately, this approach
+is deeply flawed. Many data visualization experts have issued explicit
+warnings about this:
+
+\begin{quote}
+``Stacking is useful when the sum of the amounts represented by the
+individual stacked bars is in itself a meaningful amount''
+\citep[52]{wilke2019}.
+\end{quote}
+
+\begin{quote}
+``Because this gives the visual impression of one element that is the
+sum of several others, it is very important that if the element's size
+is used to display a statistic, then that statistic must be summable.
+Stacking bars that represent counts, sums, or percentages are fine, but
+a stacked bar chart where bars show average values is generally
+meaningless.'' \citep[112]{wills2011}.
+\end{quote}
+
+\begin{quote}
+``{[}\ldots{]} We do this to ensure that aggregate statistics are always
+computed over the input data, and so users do not inadvertantly compute
+e.g., averages of averages, which can easily lead to
+misinterpretation.'' \citep{wu2022}
+\end{quote}
+
+The message is clear: stacking some statistics such as counts, sums, and
+percentages is acceptable, but stacking others is not. As outlined by
+Wills, while the sum of grouped counts produces a valid overall count,
+the sum of group means does not yield a meaningful statistic. Even
+averaging group means, as noted by \citet{wu2022}, does not produce a
+valid summary - the mean of group means differs from the grand mean
+(unless by coincidence). This problem is extends beyond barplots; for
+example, concerns about the statistical validity of stacked density
+plots have also been raised recently \citep{pu2020}.
+
+What exactly makes stacking sums acceptable but stacking other
+statistics such as means not? Wilke, Wills, and Wu et al.~all argue that
+the aggregated value should be ``meaningful''. In the case of sums,
+perhaps another way to phrase this is: ``the whole is equal to the sum
+of its parts''. Or, more specifically:
+
+\begin{quote}
+``the \emph{sum} of the \emph{sums} on the parts is equal to the
+\emph{sum} of all values''
+\end{quote}
+
+But what if we replace word ``sum'' with some placeholder (such as
+``foo''):
+
+\begin{quote}
+``the \emph{foo} of \emph{foos} on the parts is equal to the \emph{foo}
+of all values''
+\end{quote}
+
+Then, we find there are other mathematical operators that also behave
+this way. For example, the product of products or the maximum of
+maximums are also valid overall statistics, see Figure
+\ref{fig:sumprodmax}. In fact, there is an entire class of mathematical
+objects with this exact behavior: monoids. We will explore monoids in
+the following section.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/sumprodmax-1.pdf}
+\caption{\label{fig:sumprodmax} Different operators can be used for
+stacking. The values \texttt{1.2}, \texttt{1.8}, \texttt{NA}, and
+\texttt{1.5} were assigned to a 2-by-2 table and stacked using three
+different operators. Left: stacking using the sum operator. Middle:
+stacking using the product operator (note that the y-axis starts at 1).
+Right: stacking using the max operator.}
+\end{figure}
+
+\subsection{\texorpdfstring{Monoids
+\label{sec:monoids}}{Monoids }}\label{monoids}
+
+Monoids are a simple yet powerful concept that plays a key role in
+several areas of mathematics, including abstract algebra and category
+theory. The short treatment here draws primarily from \citet{fong2019},
+\citet{lawvere2009}, \citet{baez2023}, and \citet{milewski2018}. For a
+particularly accessible introduction, readers are encouraged to seek out
+\citet{fong2019} and \citet{lawvere2009}.
+
+While the term ``monoid'' may sound intimidating, the concept is really
+quite simple. A monoid represents the idea of a ``whole equal to the sum
+of its parts'', if we relax our idea about what it means to ``sum''.
+More formally, a monoid is a tuple \((M, e, \otimes)\) consisting of:
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi}.}
+\tightlist
+\item
+  A collection (set) of objects \(M\)
+\item
+  A neutral element \(e \in M\) called the \emph{monoidal unit}
+\item
+  A binary operation \(\otimes: M \times M \to M\) called the
+  \emph{monoidal product}
+\end{enumerate}
 
@@
 
-1. *Unitality*: $x \otimes e = e \otimes x = x$
-2. *Associativity*: $x \otimes (y \otimes z) = (x \otimes y) \otimes z = x \otimes y \otimes z$
-
-(for all $x, y, z \in M$)
-
-In plain words, when we have a monoid, we have some set of elements $m \in M$ and a way to combine them $\otimes$, such that the order in which we do the combining does not matter.
-Further, among the set of elements $M$, we also have a special element $e$, called the "neutral" or "identity" element, that, when combined with any other element $m \in M$, does nothing (and simply yields back $m$).
-This identity element is always unique [the proof is straightforward, for reference see e.g.
-@fong2019; @lawvere2009].
-
-Let's return to the book example from Section \ref{sec:intro}.
-Is it a monoid?
-First of all, what is our set $M$?
-This could be the set of all book stacks we can form by combining books in our bookshelf.
-Second, do we have a neutral element?
-If we are a bit creative, we can imagine an "empty" stack of zero books.
-Let's use that as our identity element.
-Third, what is our binary operation?
-This is just the operation of taking one stack of books and putting it on top of another stack.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \emph{Unitality}: \(x \otimes e = e \otimes x = x\)
+\item
+  \emph{Associativity}:
+  \(x \otimes (y \otimes z) = (x \otimes y) \otimes z = x \otimes y \otimes z\)
+\end{enumerate}
+
+(for all \(x, y, z \in M\))
+
+In plain words, when we have a monoid, we have some set of elements
+\(m \in M\) and a way to combine them \(\otimes\), such that the order
+in which we do the combining does not matter. Further, among the set of
+elements \(M\), we also have a special element \(e\), called the
+``neutral'' or ``identity'' element, that, when combined with any other
+element \(m \in M\), does nothing (and simply yields back \(m\)). This
+identity element is always unique \citep[the proof is straightforward,
+for reference see e.g.][]{fong2019, lawvere2009}.
+
+Let's return to the book example from Section \ref{sec:intro}. Is it a
+monoid? First of all, what is our set \(M\)? This could be the set of
+all book stacks we can form by combining books in our bookshelf. Second,
+do we have a neutral element? If we are a bit creative, we can imagine
+an ``empty'' stack of zero books. Let's use that as our identity
+element. Third, what is our binary operation? This is just the operation
+of taking one stack of books and putting it on top of another stack.
 
 Now we just need to verify that our structure conforms to the two rules.
-First, is the operation associative?
-Clearly the answer is "yes", since a stack of detective novels will be the same height regardless of whether we put the novels by Agatha Christie on top of Jo Nesbø novels or vice versa.
-Second, is the operation unital?
-Again, this is self-evidently true, since adding zero books on top of a stack does not change its height.
-Therefore, we do have a monoid.
-
-We can also describe the book example a bit more formally.
-Specifically, summation on natural numbers $(\mathbb{N}, 0, +)$ is a typical example of a monoid, because, as we have shown, it is associative and unital.
-Other examples of monoids include the previously mentioned products of real numbers $(\mathbb{R}, 1, \times)$, and the min and max operators $(\mathbb{R}, \infty, \min)$ and $(\mathbb{R}, -\infty, \max)$.
-As a counterexample, exponentiation is not a monoid, since it is not associative: $x^{(y^z)} \neq (x^y)^z$.
-
-The definition of a monoid is quite broad however, and applies to more exotic structures than just operations on numbers.
-For example, the set of booleans $\mathbb{B} = \{ True, False \}$ equipped with either of the logical operators $\mathbf{AND}$ or $\mathbf{OR}$ is also a monoid, and so is the multiplication of $n \times n$ square matrices.
-Likewise, the operation of concatenating vectors and computing the euclidean distance is also a monoid [@stepanov2009]:
-
-$$|| \big( || (x, y) ||_2, z \big) ||_2 = \sqrt{ \bigg(\sqrt{x^2 + y^2} \bigg)^2 + z^2} = \sqrt{x^2 + y^2 + z^2} = ||(x, y, z)||_2$$
-
-Even very "non-number-like" operations such as concatenation of strings can be a monoid:
-
-$$\texttt{"hello"} + \texttt{""} = \texttt{""} + \texttt{"hello"} = \texttt{"hello"}$$
-$$(\texttt{"quick"} + \texttt{"brown"}) + \texttt{"fox"} = \texttt{"quick"} + (\texttt{"brown"} + \texttt{"fox"}) = \texttt{"quick brown fox"}$$
-
-## Groups
-
-Groups are another fundamental concept in mathematics, representing the idea of symmetry and reversible transformations.
-While introducing two algebraic concepts in quick succession may seem like a big ask, a group is really just a monoid with the additional property.
-Specifically, for all $x, y, z \in M$, we have:
-
-3. *Inverse operator*: There exists $\otimes^{-1}$ such that if $x \otimes y = z$ then $z \otimes^{-1} x = y$
-
-In simpler terms, a group embodies the idea that "the whole is equal to the sum of its parts, *in a non-destructive way*".
-That is, when we combine two elements, we can always recover either one by "subtracting" the contribution of the other.
-Of note, like the identity element for monoids, the inverse operator in a group is also always unique [see e.g.
-@fong2019; @lawvere2009].
-
-Turning back to the book example, we can remove Agatha Christie novels off the detective novel stack and recover the height of the rest of the detective novels stack.
-Thus, summation on natural numbers is a group: $2 + 3 = 5 \implies 5 - 2 = 3$ (technically speaking, we now have to identify the underlying set as integers $\mathbb{Z}$).
-But, for example, the min and max operators lack an inverse and so do not constitute groups: if $\max{(x, 6)} = 6$, then there is no way to recover $x$ from the combined result ($6$) alone.
-
-## Groups, Monoids, and Stacking
-
-We now arrive at the core idea of the present paper.
-Suppose that we have some data set $D$, a way of summarizing this data via some associative binary operation $\otimes$, and a neutral value with respect to this operation $e$.
-In other words, $(D, \otimes, e)$ forms a monoid.
-Define $F(A)$ as the result of summarizing or "folding" some subset $A \subseteq D$ using the binary operation:
-
-$$
+First, is the operation associative? Clearly the answer is ``yes'',
+since a stack of detective novels will be the same height regardless of
+whether we put the novels by Agatha Christie on top of Jo Nesbø novels
+or vice versa. Second, is the operation unital? Again, this is
+self-evidently true, since adding zero books on top of a stack does not
+change its height. Therefore, we do have a monoid.
+
+We can also describe the book example a bit more formally. Specifically,
+summation on natural numbers \((\mathbb{N}, 0, +)\) is a typical example
+of a monoid, because, as we have shown, it is associative and unital.
+Other examples of monoids include the previously mentioned products of
+real numbers \((\mathbb{R}, 1, \times)\), and the min and max operators
+\((\mathbb{R}, \infty, \min)\) and \((\mathbb{R}, -\infty, \max)\). As a
+counterexample, exponentiation is not a monoid, since it is not
+associative: \(x^{(y^z)} \neq (x^y)^z\).
+
+The definition of a monoid is quite broad however, and applies to more
+exotic structures than just operations on numbers. For example, the set
+of booleans \(\mathbb{B} = \{ True, False \}\) equipped with either of
+the logical operators \(\mathbf{AND}\) or \(\mathbf{OR}\) is also a
+monoid, and so is the multiplication of \(n \times n\) square matrices.
+Likewise, the operation of concatenating vectors and computing the
+euclidean distance is also a monoid \citep{stepanov2009}:
+
+\[|| \big( || (x, y) ||_2, z \big) ||_2 = \sqrt{ \bigg(\sqrt{x^2 + y^2} \bigg)^2 + z^2} = \sqrt{x^2 + y^2 + z^2} = ||(x, y, z)||_2\]
+
+Even very ``non-number-like'' operations such as concatenation of
+strings can be a monoid:
+
+\[\texttt{"hello"} + \texttt{""} = \texttt{""} + \texttt{"hello"} = \texttt{"hello"}\]
+\[(\texttt{"quick"} + \texttt{"brown"}) + \texttt{"fox"} = \texttt{"quick"} + (\texttt{"brown"} + \texttt{"fox"}) = \texttt{"quick brown fox"}\]
+
+\subsection{Groups}\label{groups}
+
+Groups are another fundamental concept in mathematics, representing the
+idea of symmetry and reversible transformations. While introducing two
+algebraic concepts in quick succession may seem like a big ask, a group
+is really just a monoid with the additional property. Specifically, for
+all \(x, y, z \in M\), we have:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  \emph{Inverse operator}: There exists \(\otimes^{-1}\) such that if
+  \(x \otimes y = z\) then \(z \otimes^{-1} x = y\)
+\end{enumerate}
+
+In simpler terms, a group embodies the idea that ``the whole is equal to
+the sum of its parts, \emph{in a non-destructive way}''. That is, when
+we combine two elements, we can always recover either one by
+``subtracting'' the contribution of the other. Of note, like the
+identity element for monoids, the inverse operator in a group is also
+always unique \citep[see e.g.][]{fong2019, lawvere2009}.
+
+Turning back to the book example, we can remove Agatha Christie novels
+off the detective novel stack and recover the height of the rest of the
+detective novels stack. Thus, summation on natural numbers is a group:
+\(2 + 3 = 5 \implies 5 - 2 = 3\) (technically speaking, we now have to
+identify the underlying set as integers \(\mathbb{Z}\)). But, for
+example, the min and max operators lack an inverse and so do not
+constitute groups: if \(\max{(x, 6)} = 6\), then there is no way to
+recover \(x\) from the combined result (\(6\)) alone.
+
+\subsection{Groups, Monoids, and
+Stacking}\label{groups-monoids-and-stacking}
+
+We now arrive at the core idea of the present paper. Suppose that we
+have some data set \(D\), a way of summarizing this data via some
+associative binary operation \(\otimes\), and a neutral value with
+respect to this operation \(e\). In other words, \((D, \otimes, e)\)
+forms a monoid. Define \(F(A)\) as the result of summarizing or
+``folding'' some subset \(A \subseteq D\) using the binary operation:
+
+\[
 F(A) = a_1 \otimes a_2 \otimes a_3 \ldots \otimes a_n
-$$
-
-Note that we could have written this as $F(A) = (a_1 \otimes (a_2 \otimes (a_3 \otimes ( \ldots \otimes a_n ))))$, however, since the operation is associative, we do not have to worry about brackets.
-Also, if $A$ is empty, then clearly:
-
-$$F(\varnothing) = e$$
-
-Now comes the most important idea: combining summaries of two disjoint subsets of the data set gives us the same result as summarizing the union of the original subsets:
-
-$$
+\]
+
+Note that we could have written this as
+\(F(A) = (a_1 \otimes (a_2 \otimes (a_3 \otimes ( \ldots \otimes a_n ))))\),
+however, since the operation is associative, we do not have to worry
+about brackets. Also, if \(A\) is empty, then clearly:
+
+\[F(\varnothing) = e\]
+
+Now comes the most important idea: combining summaries of two disjoint
+subsets of the data set gives us the same result as summarizing the
+union of the original subsets:
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
+\]
 
@@
 
-$$
+\[
 % https://q.uiver.app/#q=WzAsOCxbMSwwLCJBIFxcY3VwIEIiXSxbMCwxLCJBIl0sWzIsMSwiQSBcXGN1cCBCIFxcY3VwIEMiXSxbMSwzLCJGKEEgXFxjdXAgQikiXSxbMCw0LCJGKEEpIl0sWzIsNCwiRihBIFxcY3VwIEIgXFxjdXAgQykiXSxbMiw2XSxbMSw0XSxbMCwyLCItXFxjdXAgQyJdLFswLDMsIiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9fX1dLFsxLDIsInstIFxcY3VwIChCIFxcY3VwIEMpfSIsMl0sWzEsNCwiIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn19fV0sWzIsNSwiIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn19fV0sWzMsNSwiey0gXFxvdGltZXMgRihDfSJdLFs0LDUsInstIFxcb3RpbWVzIEYoQSBcXGN1cCBDKX0iLDJdLFsxLDAsIi0gXFxjdXAgQiJdLFs0LDMsIi0gXFxvdGltZXMgRihCKSJdXQ==
@@
 \end{tikzcd}
-$$
-The top and bottom triangles in the diagram both indicate associativity: in the top triangle, we can either first take the union of $A$ and $B$ and then union that with $C$, or immediately union $A$ with $B \cup C$ and the result will be identical.
-The same applies for the monoidal summary in the bottom triangle.
-By saying that the diagram commutes, we mean that any two parallel paths (the same start and endpoint) always produce the same result.
-Practically, this means we can either first summarize each set and then combine the summaries, or first union the sets and then summarize the union, or do some mix of the two, and we will always end up with a valid summary.
-As a sidenote, the diagram above actually corresponds to something called a functor or monoid homomorphism, however, that is beyond the scope of the present paper [interested reader should see @fong2019; @lawvere2009].
-
-But this is precisely the idea of a stackable statistic.
-If our statistic conforms to the diagram, then the combined summary of subsets of the data is always a valid summary of the whole data set.
-In Wilke's words, we have discovered what makes the combined summary a "meaningful amount"!
-
-It might seem that we are done.
-However, there is still one subtle issue left.
-Suppose we summarize three parts of the data $A$, $B$, and $C$, and combine these summaries by repeatedly applying the operation, as a kind of running total $F^*$:
-
-$$
+\] The top and bottom triangles in the diagram both indicate
+associativity: in the top triangle, we can either first take the union
+of \(A\) and \(B\) and then union that with \(C\), or immediately union
+\(A\) with \(B \cup C\) and the result will be identical. The same
+applies for the monoidal summary in the bottom triangle. By saying that
+the diagram commutes, we mean that any two parallel paths (the same
+start and endpoint) always produce the same result. Practically, this
+means we can either first summarize each set and then combine the
+summaries, or first union the sets and then summarize the union, or do
+some mix of the two, and we will always end up with a valid summary. As
+a sidenote, the diagram above actually corresponds to something called a
+functor or monoid homomorphism, however, that is beyond the scope of the
+present paper \citep[interested reader should
+see][]{fong2019, lawvere2009}.
+
+But this is precisely the idea of a stackable statistic. If our
+statistic conforms to the diagram, then the combined summary of subsets
+of the data is always a valid summary of the whole data set. In Wilke's
+words, we have discovered what makes the combined summary a ``meaningful
+amount''!
+
+It might seem that we are done. However, there is still one subtle issue
+left. Suppose we summarize three parts of the data \(A\), \(B\), and
+\(C\), and combine these summaries by repeatedly applying the operation,
+as a kind of running total \(F^*\):
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-Notice that we end up with summaries on nested subsets of the data, i.e.
-$F(A)$, $F(A \cup B)$, and $F(A \cup B \cup C)$.
-This is fine, for example, when we do linked selection with a single group and compare the summary on the highlighted subset ($A$) vs.
-the entirety the data ($A \cup B$).
-
-However, what if we want to compare the subsets $A$ and $B$ directly?
-For example, suppose we do linked selection and with two highlighted groups ($A$ and $B$).
-Unfortunately, the properties of monoids do not guarantee that we will be able to recover $F(B)$ from $F(A \cup B)$ using $F(A)$.
-Put plainly, the monoidal operation may "collapse" the information contained in the disjoint parts, see Figure \ref{fig:summax}.
-To ensure that $F(B)$ is recoverable from $F(A \cup B)$, we also need the inverse operator $\otimes^{-1}$:
-
-$$F(B) = F(A \cup B) \otimes^{-1} F(A)$$
-
-Importantly, $F(A) \otimes F(B)$ is still a valid summary of $A \cup B$ even when the inverse does not exist.
-It is just that $F(A) \otimes F(B)$ may "forget" the information contained in either of its two parts.
-Note also that this is not a problem of different permutations of parts producing different sequences of accumulated summaries - this will happen regardless of whether or not there is an inverse.
-
-```{r summax}
-#| echo: false
-#| fig-height: 5
-#| fig-cap: "\\label{fig:summax} Monoids can \"collapse\" information about disjoint parts.
-Top: since sum is a group with an inverse (minus), we can recover the value of each disjoint part by subtracting the heights of the preceding segments (i.e.
-the part order does not matter).
-Bottom: since maximum lacks an inverse, we may lose information about the individual parts under certain permutations (note the absence of segments in the middle and rightmost plot)."
-#| message: false
-#| warning: false
-
-library(ggplot2)
-library(patchwork)
-
-cols <- c('grey80', '#56B4E9', '#009E73')
-label_size <- 10 / .pt
-
-stack <- function(values, fn, initial) {
-  values[is.na(values)] <- initial
-  for (i in seq_along(values)) values[i] <- fn(values[i - 1], values[i])
-  values
-}
-
-two_mean <- function(x) (c(0, x[-length(x)]) + x) / 2
-
-format_label <- function(value, symbol) {
-  ifelse(is.na(value), value, paste0(symbol, value))
-}
-
-or_na <- function(x) ifelse(is.na(x), "NA", x)
-
-group <- factor(rep(c("A", "B"), each = 3))
-original_subgroups <- rep(3:1, 2)
-original_values <- matrix(c(0.8, 1.2, 1.6, NA, 1.1, 1.5), ncol = 2)
-
-plots <- list()
-perms <- list(1:3, c(2, 1, 3), 3:1)
-
-for (i in 1:3) {
-
-  perm <- perms[[i]]
-  values <- apply(original_values, 2, function(x) x[perm]) 
-  subgroup <- original_subgroups[perm]
-
-  value <- as.numeric(values)
-  value[is.na(value)] <- 0
-
-  sums <- apply(values, 2, function(x) stack(x, sum, 0))
-  maximums <- apply(values, 2, function(x) stack(x, max, 0))
-  sums_midpoint <- apply(sums, 2, two_mean)
-  maximums_midpoint <- apply(maximums, 2, two_mean)
-
-  sums_df <- data.frame(group, subgroup, value, 
-                        midpoint = as.numeric(sums_midpoint),
-                        cum_value = as.numeric(sums))
-
-  maximums_df <- data.frame(group, subgroup, value, 
-                            midpoint = as.numeric(maximums_midpoint),
-                            cum_value = as.numeric(maximums))
-
-  p0 <- ggplot() +
-    scale_fill_manual(values = cols) +
-    guides(fill = "none") +
-    labs(x = NULL, y = NULL) +
-    theme_bw() +
-    theme(
-      axis.ticks = element_blank(),
-      panel.grid = element_blank(),
-      plot.title = element_text(size = 12, hjust = 0.1),
-      panel.border = element_blank(),
-      plot.margin = unit(c(0, 1, 0, 0), units = "cm")
-  )
-
-  dups <- duplicated(subset(maximums_df, select = c("cum_value")))
-
-  # Need to rearrange for ggplot2 to stack the bars in the correct order  
-  sums_df <- sums_df[c(3:1, 6:4), ]
-  maximums_df <- maximums_df[c(3:1, 6:4), ]
-
-  sums_df <- sums_df[-which(sums_df$value == 0), ]
-  maximums_df <- maximums_df[-which(maximums_df$value == 0), ]
-
-  # Remove rows with duplicate values  
-  last_value <- -Inf
-  for (j in rev(seq_len(nrow(maximums_df)))) {
-    if (maximums_df$cum_value[j] == last_value) {
-      maximums_df <- maximums_df[-j, ]
-    }
-    last_value <- maximums_df$cum_value[j]
-  }
-
-  plots[[3 * i - 2]] <- p0 +
-    geom_bar(data = sums_df, aes(group, cum_value, 
-                                 fill = factor(subgroup)), 
-             stat = "identity", position = "identity") +
-    geom_text(data = sums_df, aes(group, midpoint, 
-                                  label = format_label(value, "+")),
-              size = label_size, col = "white") +
-    labs(title = ifelse(i == 1, "Sum", ""))
-
-  plots[[3 * i - 1]] <- plot_spacer()
-
-  plots[[3 * i]] <- p0 +
-    geom_bar(data = maximums_df, 
-             mapping = aes(group, cum_value, 
-                           fill = factor(subgroup)), 
-             stat = "identity", position = "identity") +
-    geom_text(data = maximums_df, aes(group, midpoint, 
-                                      label = format_label(value, "max\n")),
-              size = label_size, col = "white", lineheight = 1) +
-    labs(title = ifelse(i == 1, "Maximum", ""))
-}
-
-wrap_plots(plots, nrow = 3, byrow = FALSE,
-           heights = c(1, 1/8, 1))
-```
-
-To summarize, whether our statistic is a monoid or a group determines what kinds of comparisons we can make on stacked objects:
-
-- Monoids: comparing nested subsets, part vs. whole comparisons
-- Groups: comparing disjoint subsets, part vs. part comparisons
-
-This becomes important when implementing interactive data visualization features such as linked selection.
-Specifically, if we want to compare a single selected group vs.
-the rest of the data, then it's enough for our summary statistic to be a monoid.
-If we want to implement multiple selection groups, then our statistic needs to have an inverse as well.
-If we fail either of these conditions, then we may produce data representations which are either inaccurate (in the case of a non-monoidal statistic), or accurate but not "full" (in the case of a monoid but no inverse).
-
-## Groups, Monoids, and Graphics
-
-The connection between summary statistics and algebraic structures is important because it allows us to predict which new combinations of summaries and figures will "work".
-For example, while not generally thought of as a summary statistic, the convex hull algorithm [see e.g.
-@barber1996] provides a way to reduce a set of points to a simpler "summary", see Figure \ref{fig:convexhull}.
-We can recast the operation of finding the convex hull as a monoid: starting with an empty set, we iterate through the data, adding points if they lie outside of the current hull (and removing any points in the interior of the new hull).
-This may not be the most computationally efficient method for finding convex hulls, however, since it does meet the definition of a monoid, this tells us *what convex hulls can be used for*.
-If we combine two convex hulls, we know that the new hull is a valid summary for the union of the underlying data points.
-In turn, this allows us to implement interactive features such as single-group highlighting/selection, safe in the knowledge that the underlying structure is preserved.
-
-However, does the operation of adding a point to a convex hull have an inverse?
-No.
-Once we add a point to the hull, we cannot recover the old hull by "subtracting" that point - we do not know how many points from the interior of the old hull we removed.
-This tells us that merging two convex hulls may irrevocably collapse information.
-As such, we cannot use convex hulls to display multiple selections, without resorting to some ad-hoc methods (e.g.
-keeping separate representation of the hulls and using partial transparency).
-
-```{r}
-#| echo: false
-#| fig-height: 2.5
-#| fig-cap: "\\label{fig:convexhull}Convex hull can be used as a valid representation of single-group selection but not multi-group selection.
-Both plots show data from Palmer penguins data set (Horst 2020).
-The light polygon shows the convex hull of the whole dataset, the dark polygon shows the hull of the Gentoo species, the medium-brightness polygon shows the combined hull of Gentoo and Chinstrap species, and the white dashed line shows the outline the hull of the Chinstrap penguins alone.
-Points show individual cases.
-Left: since convex hull is a monoid, we can use it to show single group selection/highlighting.
-Right: because of the lack of an inverse, we cannot guarantee that hull of the Chinstrap penguins (white-dashed line) will be recoverable from the combined hull of Gentoo and Chinstrap - we do not know how many interior points were \"forgotten\"."
-#| message: false
-#| warning: false
-
-library(ggplot2)
-library(patchwork)
-library(palmerpenguins)
-
-penguins <- na.omit(penguins)[, c("species", "body_mass_g", "flipper_length_mm")]
-chull_all <- penguins[with(penguins, chull(body_mass_g, flipper_length_mm)), ]
-
-subset_chull <- function(subset) {
-  cases <- penguins$species %in% subset
-  penguins[which(cases)[chull(penguins[cases, -1])], ]
-}
-
-species <- unique(penguins$species)
-
-chull1 <- subset_chull(c(species[2]))
-chull2 <- subset_chull(c(species[3]))
-chull3 <- subset_chull(c(species[2], species[3]))
-
-p0 <- ggplot(data = penguins, aes(body_mass_g, flipper_length_mm)) +
-  guides(fill = "none") +
-  theme_bw() +
-  guides(col = "none") +
-  labs(x = "Body mass", y = "Flipper length") +
-  theme(
-    axis.ticks = element_blank(),
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    plot.margin = unit(c(0, 1, 0, 1), "cm")
-)
-
-p1 <- p0 + 
-  geom_polygon(data = chull_all, fill = cols[1]) +
-  geom_polygon(data = chull1, fill = cols[3]) +
-  geom_point(col = "grey20", size = 0.25) +
-  labs(title = "Single group: Valid")
-
-p2 <- p0 + 
-  geom_polygon(data = chull_all, fill = cols[1]) +
-  geom_polygon(data = chull3, fill = cols[2]) +
-  geom_polygon(data = chull1, fill = cols[3]) +
-  geom_point(col = "grey20", size = 0.25) +
-  geom_polygon(data = chull2, lty = "dashed", 
-               col = "white", fill = NA) +
-  labs(title = "Multiple groups: Invalid")
-
-p1 + p2
-```
-
-## Groups, Monoids, and the Neutral Element \label{sec:advantages}
-
-There are also other advantages that groups and monoids offer for visualizing data.
-One of these is the existence of the neutral element $e$ (see Section \ref{sec:monoids}).
-While seemingly inconspicuous, the neutral element can actually be quite useful.
-
-Specifically, the neutral element ensures that we can always represent empty subsets of our data.
-Turning once more to the book example from Section \ref{sec:intro}, suppose we split the books into stacks based on the book's genre and the author's gender, and then summarize each stack by its total and average word-count.
-What if there are no books by male sci-fi authors?
-For total word-count, we have a meaningful default value - zero - as in, there are zero words across the zero books by male sci-fi authors.
-However, what about the average word-count?
-Here we run into trouble, since the mean of an empty set is not defined.
-It simply does not make sense to talk about the "average word-count across zero books".
+\]
+
+Notice that we end up with summaries on nested subsets of the data,
+i.e.~\(F(A)\), \(F(A \cup B)\), and \(F(A \cup B \cup C)\). This is
+fine, for example, when we do linked selection with a single group and
+compare the summary on the highlighted subset (\(A\)) vs.~the entirety
+the data (\(A \cup B\)).
+
+However, what if we want to compare the subsets \(A\) and \(B\)
+directly? For example, suppose we do linked selection and with two
+highlighted groups (\(A\) and \(B\)). Unfortunately, the properties of
+monoids do not guarantee that we will be able to recover \(F(B)\) from
+\(F(A \cup B)\) using \(F(A)\). Put plainly, the monoidal operation may
+``collapse'' the information contained in the disjoint parts, see Figure
+\ref{fig:summax}. To ensure that \(F(B)\) is recoverable from
+\(F(A \cup B)\), we also need the inverse operator \(\otimes^{-1}\):
+
+\[F(B) = F(A \cup B) \otimes^{-1} F(A)\]
+
+Importantly, \(F(A) \otimes F(B)\) is still a valid summary of
+\(A \cup B\) even when the inverse does not exist. It is just that
+\(F(A) \otimes F(B)\) may ``forget'' the information contained in either
+of its two parts. Note also that this is not a problem of different
+permutations of parts producing different sequences of accumulated
+summaries - this will happen regardless of whether or not there is an
+inverse.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/summax-1.pdf}
+\caption{\label{fig:summax} Monoids can ``collapse'' information about
+disjoint parts. Top: since sum is a group with an inverse (minus), we
+can recover the value of each disjoint part by subtracting the heights
+of the preceding segments (i.e.~the part order does not matter). Bottom:
+since maximum lacks an inverse, we may lose information about the
+individual parts under certain permutations (note the absence of
+segments in the middle and rightmost plot).}
+\end{figure}
+
+To summarize, whether our statistic is a monoid or a group determines
+what kinds of comparisons we can make on stacked objects:
+
+\begin{itemize}
+\tightlist
+\item
+  Monoids: comparing nested subsets, part vs.~whole comparisons
+\item
+  Groups: comparing disjoint subsets, part vs.~part comparisons
+\end{itemize}
+
+This becomes important when implementing interactive data visualization
+features such as linked selection. Specifically, if we want to compare a
+single selected group vs.~the rest of the data, then it's enough for our
+summary statistic to be a monoid. If we want to implement multiple
+selection groups, then our statistic needs to have an inverse as well.
+If we fail either of these conditions, then we may produce data
+representations which are either inaccurate (in the case of a
+non-monoidal statistic), or accurate but not ``full'' (in the case of a
+monoid but no inverse).
+
+\subsection{Groups, Monoids, and
+Graphics}\label{groups-monoids-and-graphics}
+
+The connection between summary statistics and algebraic structures is
+important because it allows us to predict which new combinations of
+summaries and figures will ``work''. For example, while not generally
+thought of as a summary statistic, the convex hull algorithm \citep[see
+e.g.][]{barber1996} provides a way to reduce a set of points to a
+simpler ``summary'', see Figure \ref{fig:convexhull}. We can recast the
+operation of finding the convex hull as a monoid: starting with an empty
+set, we iterate through the data, adding points if they lie outside of
+the current hull (and removing any points in the interior of the new
+hull). This may not be the most computationally efficient method for
+finding convex hulls, however, since it does meet the definition of a
+monoid, this tells us \emph{what convex hulls can be used for}. If we
+combine two convex hulls, we know that the new hull is a valid summary
+for the union of the underlying data points. In turn, this allows us to
+implement interactive features such as single-group
+highlighting/selection, safe in the knowledge that the underlying
+structure is preserved.
+
+However, does the operation of adding a point to a convex hull have an
+inverse? No.~Once we add a point to the hull, we cannot recover the old
+hull by ``subtracting'' that point - we do not know how many points from
+the interior of the old hull we removed. This tells us that merging two
+convex hulls may irrevocably collapse information. As such, we cannot
+use convex hulls to display multiple selections, without resorting to
+some ad-hoc methods (e.g.~keeping separate representation of the hulls
+and using partial transparency).
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{\label{fig:convexhull}Convex hull can be used as a valid
+representation of single-group selection but not multi-group selection.
+Both plots show data from Palmer penguins data set (Horst 2020). The
+light polygon shows the convex hull of the whole dataset, the dark
+polygon shows the hull of the Gentoo species, the medium-brightness
+polygon shows the combined hull of Gentoo and Chinstrap species, and the
+white dashed line shows the outline the hull of the Chinstrap penguins
+alone. Points show individual cases. Left: since convex hull is a
+monoid, we can use it to show single group selection/highlighting.
+Right: because of the lack of an inverse, we cannot guarantee that hull
+of the Chinstrap penguins (white-dashed line) will be recoverable from
+the combined hull of Gentoo and Chinstrap - we do not know how many
+interior points were ``forgotten''.}
+\end{figure}
+
+\subsection{\texorpdfstring{Groups, Monoids, and the Neutral Element
+\label{sec:advantages}}{Groups, Monoids, and the Neutral Element }}\label{groups-monoids-and-the-neutral-element}
+
+There are also other advantages that groups and monoids offer for
+visualizing data. One of these is the existence of the neutral element
+\(e\) (see Section \ref{sec:monoids}). While seemingly inconspicuous,
+the neutral element can actually be quite useful.
+
+Specifically, the neutral element ensures that we can always represent
+empty subsets of our data. Turning once more to the book example from
+Section \ref{sec:intro}, suppose we split the books into stacks based on
+the book's genre and the author's gender, and then summarize each stack
+by its total and average word-count. What if there are no books by male
+sci-fi authors? For total word-count, we have a meaningful default value
+- zero - as in, there are zero words across the zero books by male
+sci-fi authors. However, what about the average word-count? Here we run
+into trouble, since the mean of an empty set is not defined. It simply
+does not make sense to talk about the ``average word-count across zero
+books''.
 
 A common approach is to omit the objects representing empty subsets.
-However, with non-monoidal summaries, this approach is flawed and leads ambiguity.
-Suppose we are drawing the bars of averages.
-If we omit bars representing the empty categories, then the missingness of a bar can indicate either of two things: either there are *no* cases in the category, or there *are* cases and their average is equal to the lower y-axis limit.
-There is just no way to tell from the graphic alone.
-In contrast, with monoidal summaries such as sums, the message of an object's absence is unambiguous.
-Regardless of whether there are zero or hundred cases in a category, if the bar representing them is missing, this means that their sum is zero.
-
-The neutral element can also provide a new perspective on a perennial debate in data visualization: should the base of a barplot always start at zero?
-Some authors argue that starting bars from a non-zero base makes the lengths of the bars meaningless [@cleveland1985].
-However, @wilkinson2012, for example, suggests that the bar base is really a statement about scales rather than about graphics, and as such there is no compulsion for starting it at zero.
-We offer the following new perspective: often, it might make sense for the base of the barplot to start at $e$, the neutral element, since this represents the default state of doing nothing.
-In other words, the bar base is tied to the properties of the *statistic*, rather than the just the graphic or scale.
-
-In a barplot of sums or counts, this approach aligns with the traditional prescription, since the neutra element is zero.
-However, in a barplot of products, the neutral element is one.
-Thus, we may should our bars at one.
-One might object that this invalidates the y-axis as a ratio scale [@stevens1946], since a bar representing the value 4 will be no longer twice as tall as the bar representing the value 2.
-And indeed, if we start the base at one, the bar representing the value 2 will be only 2 - 1 = 1 units tall, whereas the bar representing the value 4 will be 4 - 1 = 3 units tall.
-However, we need to think about what the values represent.
-The bar representing the value 2 shows a 200% change or 100% increase; the bar representing 4 shows 400% change or 300% increase.
-Therefore, it makes sense that the bar representing 4 is three times taller.
-Also, this scale still fits the @stevens1946 definition of a ratio scale, since we can still determine equality of ratios: 1.5 is to 2 (100% / 50% = 2, double the increase) as 2 is to 3 (200% / 100% = 2, double the increase).
-
-This example provides further evidence to support the claim that we cannot separate statistics from other data visualization components such as graphics and scales.
-Even before we apply any transformations, the method we have used to derive our summary statistics determines important properties of our visualization.
-We need to be mindful of this.
-
-## Groups, Monoids, and Computation
-
-Groups and monoids also come with some inherent computational advantages.
-Specifically, if a summary statistic is a monoid, associativity guarantees that it can be computed in a single pass through the data (i.e.
-$O(n)$ complexity), and in a distributed/parallel fashion [see e.g.
-@lammel2008; @lin2013].
-This is why monoids are popular in functional programming, where they are used alongside the so-called `fold` or `reduce` functions [see e.g.
-@milewski2018; @abelson2022].
-To be fair, computational efficiency is usually not the primary concern in data visualization: most data visualization systems are designed to work in a single-threaded environment, and rely on summary statistics that are fairly cheap to compute.
-Nevertheless, the guarantee of inherent efficiency can still be useful when visualizing larger data sets.
-
-The computational advantages of groups and monoids become significant in interactive data visualization, however.
-Specifically, when designing an interactive data visualization, we may be able to exploit the fact that summaries on fine-grained subsets of the data "add up" to the summaries on coarse-grained subsets (their unions) to avoid unnecessary computation.
-Consider a linked histogram with reactive axes and interactive manipulation of binwidth and anchor.
-When the user engages in selection, the counts within the individual highlighted segments may change but the counts across the whole stacked bars will not.
-Thus, there is no need to recompute and update the upper y-axis limit - we know this will stay constant throughout selection.
-However, if the user interactively changes the binwidth or anchor, then both the counts within the whole bars *and* the highlighted segments may change and so we need to update the graphic appropriately.
+However, with non-monoidal summaries, this approach is flawed and leads
+ambiguity. Suppose we are drawing the bars of averages. If we omit bars
+representing the empty categories, then the missingness of a bar can
+indicate either of two things: either there are \emph{no} cases in the
+category, or there \emph{are} cases and their average is equal to the
+lower y-axis limit. There is just no way to tell from the graphic alone.
+In contrast, with monoidal summaries such as sums, the message of an
+object's absence is unambiguous. Regardless of whether there are zero or
+hundred cases in a category, if the bar representing them is missing,
+this means that their sum is zero.
+
+The neutral element can also provide a new perspective on a perennial
+debate in data visualization: should the base of a barplot always start
+at zero? Some authors argue that starting bars from a non-zero base
+makes the lengths of the bars meaningless \citep{cleveland1985}.
+However, \citet{wilkinson2012}, for example, suggests that the bar base
+is really a statement about scales rather than about graphics, and as
+such there is no compulsion for starting it at zero. We offer the
+following new perspective: often, it might make sense for the base of
+the barplot to start at \(e\), the neutral element, since this
+represents the default state of doing nothing. In other words, the bar
+base is tied to the properties of the \emph{statistic}, rather than the
+just the graphic or scale.
+
+In a barplot of sums or counts, this approach aligns with the
+traditional prescription, since the neutra element is zero. However, in
+a barplot of products, the neutral element is one. Thus, we may should
+our bars at one. One might object that this invalidates the y-axis as a
+ratio scale \citep{stevens1946}, since a bar representing the value 4
+will be no longer twice as tall as the bar representing the value 2. And
+indeed, if we start the base at one, the bar representing the value 2
+will be only 2 - 1 = 1 units tall, whereas the bar representing the
+value 4 will be 4 - 1 = 3 units tall. However, we need to think about
+what the values represent. The bar representing the value 2 shows a
+200\% change or 100\% increase; the bar representing 4 shows 400\%
+change or 300\% increase. Therefore, it makes sense that the bar
+representing 4 is three times taller. Also, this scale still fits the
+\citet{stevens1946} definition of a ratio scale, since we can still
+determine equality of ratios: 1.5 is to 2 (100\% / 50\% = 2, double the
+increase) as 2 is to 3 (200\% / 100\% = 2, double the increase).
+
+This example provides further evidence to support the claim that we
+cannot separate statistics from other data visualization components such
+as graphics and scales. Even before we apply any transformations, the
+method we have used to derive our summary statistics determines
+important properties of our visualization. We need to be mindful of
+this.
+
+\subsection{Groups, Monoids, and
+Computation}\label{groups-monoids-and-computation}
+
+Groups and monoids also come with some inherent computational
+advantages. Specifically, if a summary statistic is a monoid,
+associativity guarantees that it can be computed in a single pass
+through the data (i.e.~\(O(n)\) complexity), and in a
+distributed/parallel fashion \citep[see e.g.][]{lammel2008, lin2013}.
+This is why monoids are popular in functional programming, where they
+are used alongside the so-called \texttt{fold} or \texttt{reduce}
+functions \citep[see e.g.][]{milewski2018, abelson2022}. To be fair,
+computational efficiency is usually not the primary concern in data
+visualization: most data visualization systems are designed to work in a
+single-threaded environment, and rely on summary statistics that are
+fairly cheap to compute. Nevertheless, the guarantee of inherent
+efficiency can still be useful when visualizing larger data sets.
+
+The computational advantages of groups and monoids become significant in
+interactive data visualization, however. Specifically, when designing an
+interactive data visualization, we may be able to exploit the fact that
+summaries on fine-grained subsets of the data ``add up'' to the
+summaries on coarse-grained subsets (their unions) to avoid unnecessary
+computation. Consider a linked histogram with reactive axes and
+interactive manipulation of binwidth and anchor. When the user engages
+in selection, the counts within the individual highlighted segments may
+change but the counts across the whole stacked bars will not. Thus,
+there is no need to recompute and update the upper y-axis limit - we
+know this will stay constant throughout selection. However, if the user
+interactively changes the binwidth or anchor, then both the counts
+within the whole bars \emph{and} the highlighted segments may change and
+so we need to update the graphic appropriately.
 
 In other words, associativity allows us to implement strategic updating.
-Fine-grained summaries (individual segments) need to be updated when the more coarse-grained summaries (whole stacked bars) change, but not vice versa.
-By pre-computing summaries on different levels of aggregation, we may be able to reduce computational effort.
-
-In fact, if our summary statistic has an inverse, we can be even more computationally efficient.
-Suppose that one data point out of a thousand is removed from selection.
-If our summary has an inverse, we can use it to subtract the contribution of that single data point, instead of having to recompute the whole summary from scratch.
-This is, in fact, the principle behind the fast updates in the popular JavaScript library Crossfilter, which is unfortunately no longer actively maintained [@crossfilter2023; see also the Github discussion initiated by @crossfilter2024]^[Note that, as the present date, we are not aware of any explicit discussion of monoids or groups in the Crossfilter documentation].
-
-However, while the inverse does bring advantages, it is also the case that fewer summary statistics have it.
-For instance, as was mentioned before, while maximum operator is a valid monoid, it lacks an inverse and as such does not meet the definition of a group.
-Once we summarize some data with the maximum, we cannot retrieve the second or third highest value - that information is gone forever.
-And there are many other operations (such as the convex hull in Figure \ref{fig:convexhull}) which lack an inverse.
-As such, when designing and interactive data visualization library, we need to carefully weigh the ability to support different kinds of summary statistics (general, monoids, groups) versus supporting different kinds of interactions (e.g.
-single-group selection vs.
-multi-group selection).
-
-# Conclusion
-
-We hope we have demonstrated that, when designing interactive graphics, algebraic thinking is invaluable.
-To support specific interactive features, the statistics underlying our plots must possess appropriate algebraic properties.
-For instance, if we want to display the result of splitting our data into parts and combining these parts back together, then we need either of two classes of mathematical structures.
-If we want to compare nested subsets of our data (for example, via single-group linked selection) then our summaries need to be monoids.
-If we want to compare disjoint parts of our data (multi-group linked selection), then our summaries need to be (algebraic) groups.
-The properties of these structures ensure that our figures are statistically sound, computationally efficient, and adhere to good design principles.
-
-We have focused on unitality, associativity, and inverses, however, it may be useful to consider other algebraic properties as well.
-For instance, many of the visual attributes that we use to represent our data can only *grow* in one direction.
-Such is the case with, for example, width, height, and area (there is no such thing as "negative area").
-If the visual attribute can only grow in one direction, then our summary statistic should be *monotonic*, such that if $x_1 \leq y_1$ and and $x_2 \leq y_2$ then $x_1 \otimes x_2 \leq y_1 \otimes y_2$.
-Similarly, in many types of plots, the order of the data points does not matter, however, in some, such as time-series plots, it does.
-Thus, it may be important to determine whether or not the statistics underlying our plots are *commutative*, such that $x \otimes y = y \otimes x$.
-Further exploration of these properties in relation to interactive graphics may yield useful insights, however, that is outside of the scope of the present paper.
+Fine-grained summaries (individual segments) need to be updated when the
+more coarse-grained summaries (whole stacked bars) change, but not vice
+versa. By pre-computing summaries on different levels of aggregation, we
+may be able to reduce computational effort.
+
+In fact, if our summary statistic has an inverse, we can be even more
+computationally efficient. Suppose that one data point out of a thousand
+is removed from selection. If our summary has an inverse, we can use it
+to subtract the contribution of that single data point, instead of
+having to recompute the whole summary from scratch. This is, in fact,
+the principle behind the fast updates in the popular JavaScript library
+Crossfilter, which is unfortunately no longer actively maintained
+\citetext{\citealp{crossfilter2023}; \citealp[see also the Github
+discussion initiated by][]{crossfilter2024}}\footnote{Note that, as the
+  present date, we are not aware of any explicit discussion of monoids
+  or groups in the Crossfilter documentation}.
+
+However, while the inverse does bring advantages, it is also the case
+that fewer summary statistics have it. For instance, as was mentioned
+before, while maximum operator is a valid monoid, it lacks an inverse
+and as such does not meet the definition of a group. Once we summarize
+some data with the maximum, we cannot retrieve the second or third
+highest value - that information is gone forever. And there are many
+other operations (such as the convex hull in Figure
+\ref{fig:convexhull}) which lack an inverse. As such, when designing and
+interactive data visualization library, we need to carefully weigh the
+ability to support different kinds of summary statistics (general,
+monoids, groups) versus supporting different kinds of interactions
+(e.g.~single-group selection vs.~multi-group selection).
+
+\section{Conclusion}\label{conclusion}
+
+We hope we have demonstrated that, when designing interactive graphics,
+algebraic thinking is invaluable. To support specific interactive
+features, the statistics underlying our plots must possess appropriate
+algebraic properties. For instance, if we want to display the result of
+splitting our data into parts and combining these parts back together,
+then we need either of two classes of mathematical structures. If we
+want to compare nested subsets of our data (for example, via
+single-group linked selection) then our summaries need to be monoids. If
+we want to compare disjoint parts of our data (multi-group linked
+selection), then our summaries need to be (algebraic) groups. The
+properties of these structures ensure that our figures are statistically
+sound, computationally efficient, and adhere to good design principles.
+
+We have focused on unitality, associativity, and inverses, however, it
+may be useful to consider other algebraic properties as well. For
+instance, many of the visual attributes that we use to represent our
+data can only \emph{grow} in one direction. Such is the case with, for
+example, width, height, and area (there is no such thing as ``negative
+area''). If the visual attribute can only grow in one direction, then
+our summary statistic should be \emph{monotonic}, such that if
+\(x_1 \leq y_1\) and and \(x_2 \leq y_2\) then
+\(x_1 \otimes x_2 \leq y_1 \otimes y_2\). Similarly, in many types of
+plots, the order of the data points does not matter, however, in some,
+such as time-series plots, it does. Thus, it may be important to
+determine whether or not the statistics underlying our plots are
+\emph{commutative}, such that \(x \otimes y = y \otimes x\). Further
+exploration of these properties in relation to interactive graphics may
+yield useful insights, however, that is outside of the scope of the
+present paper.
 
 We do not advocate for exclusively visualizing groups and monoids.
-Rather, we want to highlight the trade-off present in the choice of the summary statistic.
-Few summary statistics meet the definition of a group, but, if we choose such a statistic, then we can do a lot of things with it.
-With a monoid, our options are more limited, and with a non-monoidal statistic, even more so.
-Thus, like many past authors [such as @tufte2001; @tukey1977; @wilkinson2012], we simply want to point to the importance of thinking critically about the mathematical structure underlying our graphics.
-Geometric objects, statistics, and interaction are all deeply connected, and a full elaboration on this web of interdependence may require a lot more work.
-Groups and monoids merely allow us to carve out a small space of well-behaved graphics.
-
-On a more practical note, we also need a framework for efficiently transforming raw data into summary statistics that can be displayed by interactive graphics.
-A decade ago, @wickham2009 noted the lack of such a formal pipeline (or "plumbing"), and, despite some attempts at implementing such pipelines [see e.g.
-@lawrence2009], that same sentiment was echoed again recently [@vanderplas2020].
-Ideally, such a framework would minimize recomputation during user interaction.
-The computational "shortcuts" that groups and monoids provide may be invaluable here.
-We envision a visualization pipeline structured as a hierarchy of reactive partitions, where only the downstream (finer) partitions need to be recomputed when the upstream (coarser) partitions change, but not vice versa.
-We are currently exploring this concept in an R package called `plotscaper`^[`https://github.com/bartonicek/plotscaper`] (the project still has far to go to be truly generic).
+Rather, we want to highlight the trade-off present in the choice of the
+summary statistic. Few summary statistics meet the definition of a
+group, but, if we choose such a statistic, then we can do a lot of
+things with it. With a monoid, our options are more limited, and with a
+non-monoidal statistic, even more so. Thus, like many past authors
+\citep[such as][]{tufte2001, tukey1977, wilkinson2012}, we simply want
+to point to the importance of thinking critically about the mathematical
+structure underlying our graphics. Geometric objects, statistics, and
+interaction are all deeply connected, and a full elaboration on this web
+of interdependence may require a lot more work. Groups and monoids
+merely allow us to carve out a small space of well-behaved graphics.
+
+On a more practical note, we also need a framework for efficiently
+transforming raw data into summary statistics that can be displayed by
+interactive graphics. A decade ago, \citet{wickham2009} noted the lack
+of such a formal pipeline (or ``plumbing''), and, despite some attempts
+at implementing such pipelines \citep[see e.g.][]{lawrence2009}, that
+same sentiment was echoed again recently \citep{vanderplas2020}.
+Ideally, such a framework would minimize recomputation during user
+interaction. The computational ``shortcuts'' that groups and monoids
+provide may be invaluable here. We envision a visualization pipeline
+structured as a hierarchy of reactive partitions, where only the
+downstream (finer) partitions need to be recomputed when the upstream
+(coarser) partitions change, but not vice versa. We are currently
+exploring this concept in an R package called
+\texttt{plotscaper}\footnote{\texttt{https://github.com/bartonicek/plotscaper}}
+(the project still has far to go to be truly generic).
 
 Data visualization, in practice, is rarely about mathematical purity.
-Real world data is often flawed or incomplete, and it takes cunning and experience to dredge up valuable insights from it.
-Nevertheless, we believe that simple algebraic thinking offers a valuable foundation.
-Concepts like monoids and groups provide a rigorous framework for reasoning about graphics, interactive or otherwise.
-Ultimately, these concepts may empower us to build a broader class of robust interactive data visualizations.
-
-# Disclosure Statement
+Real world data is often flawed or incomplete, and it takes cunning and
+experience to dredge up valuable insights from it. Nevertheless, we
+believe that simple algebraic thinking offers a valuable foundation.
+Concepts like monoids and groups provide a rigorous framework for
+reasoning about graphics, interactive or otherwise. Ultimately, these
+concepts may empower us to build a broader class of robust interactive
+data visualizations.
+
+\section{Disclosure Statement}\label{disclosure-statement}
 
@@
 
-# Bibliography
+\bibliographystyle{apalike}
+\renewcommand\refname{Bibliography}
+\bibliography{references.bib}
+
+\end{document}

### 2024-10-19 paper.tex
Wrote some more

@@
 The essence of data visualization is comparison. When we visualize, we
-split our data into parts, summarize each part by a set of one or more
-statistics, and then encode these summaries into visual channels such as
-position, size, or color that we then use to draw geometric objects
+split our data into parts, summarize each part one or more summary
+statistics, encode the summaries into visual attributes such as
+position, size, or color, and finally use these attributes to draw
+geometric objects
 \citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
 recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
-example, when drawing a typical barplot, we split our data based on the
-levels of some categorical variable, count or sum up the values of some
-other variable, and finally represent these summaries by drawing bars of
+example, when drawing a typical barplot, we split our data into parts
+based on the levels of some categorical variable, count the number of
+cases in each part, and finally represent the counts by drawing bars of
 the corresponding heights. In this way, each bar becomes a visual
@@
 
-To ground this idea in a more concrete example, one that will be used
-throughout the rest of this paper, imagine going through your bookshelf,
-sorting the books by genre, and stacking them on top of each other. You
-would construct a physical version of the barplot, forming one ``bar''
-(stack) per genre: one bar for sci-fi novels, one bar for romance, one
-bar for detective fiction, and so on. Even from this very simple
-example, interesting mathematical properties jump out: for example, a
-single book cannot go into two stacks (the subsets of the data are
-\emph{disjoint}) and if all of the stacks are combined back together, we
-should end up with the bookshelf we started with \citep[the subsets are
+To ground this idea in a more concrete example that will be used
+throughout this paper, imagine going through your bookshelf, sorting the
+books by genre, and stacking them on top of each other. You would
+construct a physical version of the barplot, forming one ``bar'' (stack)
+per genre: one bar for sci-fi novels, one bar for romance, one bar for
+detective fiction, and so on. Even from this very simple example,
+interesting mathematical properties jump out: for example, a single book
+cannot go into two stacks (the subsets of the data are \emph{disjoint})
+and if all of the stacks are combined back together, we should end up
+with the bookshelf we started with \citep[the subsets are
 \emph{surjective} and cover the entirety of the data set,
@@
 organizing books in the way we just described, we can learn about our
-data by splitting it into parts. A natural question that arises is: can
-we learn more by subdividing our data into even smaller parts and
-comparing these? For instance, in our book example, we could split each
-stack of books into two smaller stacks based on the gender of the
-author. Each smaller stack will then represent the unique combination
-(Cartesian product) of the two categorical variables (genre-gender).
-Moreover, there arises a hierarchical relationship between the stacks:
-by combining two genre-gender stacks (child components) together, we can
+data by splitting it into parts. A natural question then arises: can we
+learn more by subdividing our data into even smaller parts and comparing
+these? For instance, in our book example, we can split each stack of
+books into two smaller stacks based on the gender of the author. Each
+smaller stack will then represent the unique combination (Cartesian
+product) of the two categorical variables (genre and gender). Moreover,
+there arises a hierarchical relationship between the stacks: by
+combining two genre-gender stacks (child components) together, we can
 recover the original genre stack (parent component). Similar situation
@@
 However, once we have subdivided each category into its child
-components, we need a way of representing this information that visually
-preserves the hierarchical nature of the relationship. In data
+components, we need to represent the information in a way that visually
+preserves the hierarchical nature of this relationship. In data
 visualization, there are two popular methods for doing this, known as
@@
 and across bars \citep{cleveland1984}. Subsequent research has
-independently validated and expanded upon these findings \citep[see
+independently validated these findings and expanded upon them \citep[see
 e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
@@
 
-However, despite the mixed reception of stacked plots in static
-visualization, they remain a popular choice in interactive data
+However, despite the mixed reception towards stacking in static
+visualization, it remains a popular choice in interactive data
 visualization. Among popular interactive data visualization libraries,
@@
 We start by discussing the properties of visual perception and how they
-apply to both static and interactive data visualizations. Next, we
-explore the process of computing statistical summaries that form the
-basis of our plots. Finally, we dive into the algebraic properties of
-the functions underlying these statistical summaries. Throughout the
-text, we use the example of the humble barplot to ground our discussion.
+apply to both static and interactive visualizations. Next, we explore
+the process of computing statistical summaries that form the basis of
+our plots. Finally, we dive into the algebraic properties of the
+functions underlying these statistical summaries. Throughout the text,
+we use the example of the humble barplot to ground our discussion.
 
@@
 interdependence may in turn empower us to innovate and build new kinds
-of robust interactive visualizations.
+of robust and expressive interactive visualizations.
 
@@
 
-Whether we intend them to or not, the Gestalt principles how we perceive
-data visualizations. As such, it is a good idea to try to use them to
-our advantage. Sometimes, little effort is needed. Such is the case, for
-example, when identifying clusters in a scatterplot \citep{rosli2015}:
-the principles of proximity and similarity ensure that clouds of
-similarly shaped and colored points will be perceived as discrete groups
-by the observer. In other cases, however, our design choices can have a
-significant impact. For instance, another way to have objects to be
-perceived as part of a group is to draw a closed shape around them
-\citep{cairo2012}. Alternatively, if we want to highlight a particular
-part of a data set, we can leverage the figure and ground principle and
-use attributes such as color and saturation to make it stand out
-\citep[for specific examples, see][34]{franconeri2021, wilke2019}.
+Whether we intend them to or not, the Gestalt principles affect how our
+data visualizations are perceived. As such, it is a good idea to try to
+use them to our advantage. Sometimes, little effort is needed. Such is
+the case, for example, when identifying clusters in a scatterplot
+\citep{rosli2015}: the principles of proximity and similarity ensure
+that clouds of similarly shaped and colored points will be perceived as
+discrete groups by the observer. In other cases, however, our design
+choices can have a significant impact. For instance, another way to have
+objects to be perceived as part of a group is to draw a closed shape
+around them \citep{cairo2012}. Alternatively, if we want to highlight a
+particular part of a data set, we can leverage the figure and ground
+principle and use attributes such as color and saturation to make it
+stand out \citep[for specific examples,
+see][34]{franconeri2021, wilke2019}.
 
@@
 \citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
-Either way, whether intentional or not, poor design can have a
-significant impact on the legibility and effectiveness of statistical
-graphics.
+Whether intentional or not, poor design can have a significant impact on
+the legibility and effectiveness of statistical graphics.
 
@@
 different dynamically-generated subsets of the data \citep[similar to
-interactive filtering, see e.g.][]{heer2012}. The ability to rapidly
+interactive filtering, see e.g.][]{heer2012}. The ability to quickly
 materialize different views of the data makes linked selection a
-particularly versatile tool for data exploration.
-
-Yet, despite its versatility, linked selection also imposes some
-fundamental constraints on the visualization. Specifically, all objects
-we visualize need to be able to:
+versatile tool for data exploration.
+
+Yet, despite its usefulness, linked selection also imposes some
+fundamental constraints on the figure. Specifically, all objects we
+visualize need to be able to:
 
@@
 
-Dodged barplots lack the degree of visual consistency found in stacked
-barplots. In a dodged barplot, the clusters of segments placed
-side-by-side lack a shared border (common region), leading to several
-design challenges. First, when selection occurs, the jagged contour of
-the dodged barplot can change dramatically as more cases are added or
-removed. This creates distracting visual clutter that makes it difficult
-to track changes over time. Second, selection may alter the height of
-the tallest bar segment in the plot. Since the height of the bar
-segments is not bound by the height of the tallest bar (as is the case
-in stacked barplot), this presents us with a dilemma. Either either we
-make the upper y-axis limit dynamic, meaning that the bar segments will
-always fit the plotting region, but then the context that the upper
-y-axis limit provides is lost with each time selection happens.
-Alternatively, we keep the upper y-axis limit static, but then we risk
-the bar segments growing outside of the bounds of the plotting region.
+Dodged barplots lack this degree of visual consistency. In a dodged
+barplot, bar segments placed side-by-side in clusters, and, since each
+segment can be of different length, these clusters lack a shared border
+(common region). This leads to several design challenges. First, when
+selection occurs, the jagged contour of the dodged barplot can change
+dramatically as more cases are added or removed. This creates
+distracting visual clutter that makes it difficult to track changes over
+time. Second, since the heights of the bar segments are not bound by any
+value present in the plot (unlike the heights of the segments in the
+stacked barplot, which are bound by the height of the tallest bar), we
+need to decide how the handle the upper y-axis limit in response to
+selection. One option is to make the upper y-axis limit dynamic. If we
+do this, we ensure that the bar segments always fit the plotting region,
+however, we also lose the context that the upper y-axis limit provides
+each time selection occurs. Alternatively, we can assign a static value
+to the upper y-axis limit, however, then we risk the bar segments
+growing outside of the bounds of the plotting region.
 

### 2024-10-23 paper.Rmd
Rewrote some sections

@@
 ---
-title: "No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction"
+title: "No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction"
 blinded: 0
@@
 This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids.
-Specifically, if we want our graphics to support interactive features which break the data into multiple parts and then combine them back together (such as in the case of linked selection), then we need the statistics in our plots to fulfill certain algebraic properties.
+Specifically, if we want our graphics to support interactive features which split our data into parts and then combine these parts back together (such as linked selection), then we need the statistics in our plots to fulfill certain algebraic properties.
 Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.
@@
 The essence of data visualization is comparison.
-When we visualize, we split our data into parts, summarize each part one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and finally use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019].
-For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent the counts by drawing bars of the corresponding heights.
-In this way, each bar becomes a visual representation of its own "small data set", a subset of the original data.
-
-To ground this idea in a more concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other.
-You would construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on.
-Even from this very simple example, interesting mathematical properties jump out: for example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].
-
-Now, just as we can learn about the contents of our library by organizing books in the way we just described, we can learn about our data by splitting it into parts.
-A natural question then arises: can we learn more by subdividing our data into even smaller parts and comparing these?
-For instance, in our book example, we can split each stack of books into two smaller stacks based on the gender of the author.
-Each smaller stack will then represent the unique combination (Cartesian product) of the two categorical variables (genre and gender).
-Moreover, there arises a hierarchical relationship between the stacks: by combining two genre-gender stacks (child components) together, we can recover the original genre stack (parent component).
+When we visualize, we split our data into parts, summarize each part by one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and, finally, use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019].
+For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent these counts by drawing bars of corresponding heights.
+In this way, each geometric object (bar) becomes a representation of its own small data set, a subset of the original data.
+
+To ground this idea in a concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other.
+You will construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on.
+Even this very simple example reveals some interesting mathematical properties.
+For example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].
+
+Now, just as we can learn about the contents of our library by sorting books into stacks, we can learn about our data by splitting it into parts.
+A natural question then arises: can we learn more by subdividing our data into even smaller parts?
+For instance, in our book example, we can split each stack of books into smaller stacks based on the gender of the author.
+Each smaller stack then represents the unique combination (Cartesian product) of the two categorical variables (genre and gender).
+Moreover, there arises a hierarchical relationship between the stacks: by combining genre-gender stacks (child components), we can recover the original genre stack (parent component).
 Similar situation often arises in interactive graphics, where we use techniques such as linked selection to highlight specific subsets of our data.
-Either way, by forming these finer partitions, we can represent our data with greater granularity and discover trends that may otherwise hide in the aggregate.
-
-However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves the hierarchical nature of this relationship.
+Either way, by forming these finer partitions of our data, we can represent it with greater granularity and discover trends that may otherwise hide in the aggregate.
+
+However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves this hierarchical relationship.
 In data visualization, there are two popular methods for doing this, known as "stacking" and "dodging" (see Figure \ref{fig:playfair}).
 These methods have existed at least since the time of William Playfair [-@playfair1801; -@playfair1822].
-Let's briefly illustrate them on the example of the classical barplot.
+Let's briefly illustrate them on the example of the barplot.
 In a stacked barplot, bar segments representing the product of the two factor variables are plotted vertically on top of each other ("stacked"; the operation can be also thought of as "highlighting" parts of the bar).
-In contrast, in a dodged barplot, bars are plotted side-by-side, as clusters of bars.
-Finally, there is also a third popular method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting (this technique may have been a bit challenging for William Playfair but has been made simple with computers).
+In contrast, in a dodged barplot, bars are plotted side-by-side, as "clusters".
+Finally, there is also a third more recent method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting.
 
@@
 #| fig-align: "center"
-#| fig-cap: "Examples of the three methods for dealing with bars that have subdivided into segments.
+#| fig-cap: "Examples of three methods for dealing with bars that have split into child components (segments).
 Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822).
@@
   geom_bar(position = "identity", alpha = 0.5) +
-  scale_x_discrete(labels = abbreviate) +
+  scale_x_discrete(labels = NULL) +
   guides(fill = "none") +
@@
 Further, in a landmark study, @cleveland1984 showed that people tend to be less accurate when reading information from stacked bar charts as opposed to dodged bar charts.
-Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below it, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984].
+Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984].
 Subsequent research has independently validated these findings and expanded upon them [see e.g.
 @heer2010; @thudt2016; @quadri2021].
-Ultimately, the suboptimal statistical legibility of stacking has led some authors to advise against its use altogether [@kosara2016; @wilke2019], or at least urge caution [see e.g.
-@byron2008; @cairo2014; @franconeri2021].
+Due to this suboptimal statistical legibility, many data visualization researchers have urged caution about stacking [see e.g.
+@byron2008; @cairo2014; @franconeri2021], and some have even discouraged its use altogether [@kosara2016; @wilke2019].
 
@@
 
-We argue that not only is stacking a useful technique, but that it can in fact offer valuable insights about the mathematical structure underlying our visualizations.
+We argue that not only is stacking a useful technique, despite its limitations, but that it can also offer valuable insights into the mathematical structure underlying our visualizations.
 To make the strongest case possible, we bring forward evidence from several levels of the visualization process.
@@
 
-Our goal is to convince you that, in order to produce coherent and efficient interactive graphics, we simply cannot treat graphics, statistics, and interaction as independent components.
+Our goal is to convince you that, in order to produce coherent and well-behaved interactive visualizations, we simply cannot treat graphics, statistics, and interaction as independent components.
 Instead we need to consider them holistically, within the context of their respective visual and mathematical properties.
@@
 
-Over the course of our evolutionary history, the human brain has evolved a sophisticated pipeline for processing visual information.
-This pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp.
-20].
-Many of the findings from this research are directly applicable to data visualization.
+Over the course of our evolutionary history, the human brain has evolved a sophisticated network for processing visual input.
+This visual processing pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp.
+20], and many of its findings are directly applicable to data visualization.
 For example, certain particularly salient visual stimuli are perceived automatically, rapidly (<10 ms), in parallel, and at a constant speed regardless of the number of distractors [@treisman1985].
@@
 
-However, many of the findings from visual perception research are quite specific; to guide effective visualization, we also need some general, high-level rules.
+However, many of the visual perception findings are quite specific and niche; to guide effective visualization, we also need some general, high-level rules.
 One such set of high-level rules are the Gestalt principles of visual perception [see e.g.
 @cairo2012; @ware2019; @rosli2015; @vanderplas2020; @todorovic2008; @pinker1990].
-These principles formulate a set of fundamental "laws" related to the perception of patterns, primarily visual.
+These principles formulate a set of fundamental "laws" related to the perception of (primarily visual) patterns.
 They were discovered by the founders of the Gestalt school of psychology early in the 20th century, primarily by means of self-observation, and, despite not being grounded in sound understanding of the neural mechanisms of the brain, have nevertheless been later independently supported by findings from cognitive neuroscience [@guberman2017; @ware2019; @wertheimer1938].
@@
 Second, the closure and common region principles state that regions with closed contours tend to be perceived as single, unified objects.
-Finally, the figure and ground principle states we tend to see one part of a scene as the central object (the "figure") while the rest is perceived as the background (the "ground").
+Finally, the figure and ground principle states that we tend to perceive one part of a visual scene as the central object (the "figure") and the rest as background (the "ground").
 This can be modulated by other Gestalt principle: for example, smaller, darker, and symmetric objects with closed contours are more likely to be perceived as the figure [@ware2019].
@@
 Sometimes, little effort is needed.
-Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups by the observer.
+Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups.
 In other cases, however, our design choices can have a significant impact.
-For instance, another way to have objects to be perceived as part of a group is to draw a closed shape around them [@cairo2012].
+For instance, another way to make objects be perceived as part of a group is to draw a closed shape around them [@cairo2012].
 Alternatively, if we want to highlight a particular part of a data set, we can leverage the figure and ground principle and use attributes such as color and saturation to make it stand out [for specific examples, see @franconeri2021; @wilke2019, p.
@@
 Finally, ignoring or violating the Gestalt principles can make our graphics less legible or even misleading.
-@tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements he termed "chartjunk".
+@tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements termed "chartjunk".
 Similarly, @cairo2019 highlights many cases of graphics that violate the Gestalt principles.
@@
 As such, an interactive visualization may conform to the Gestalt principles at one instance of time and violate them in another.
-In fact, interactive visualizations can even differ in how well they leverage design principles throughout time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].
-
-Thus, interactivity inherently places further constraints the design of figures.
+In fact, interactive visualizations can even differ in how well they leverage design principles *across* time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].
+
+Thus, interactivity inherently places additional constraints the design of our figures.
 If we want well-designed interactive figures, we need to ensure that they conform to good design principles across all states induced by user-interaction.
 This may not always be easy.
-We will investigate this idea in the following section, on the example of linked selection.
+We will investigate the design challenged posed by interaction in the following section, on the example of linked selection.
 
@@
 
-While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain types of plots and visualization styles.
-Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection.
-In fact, this may explain why stacking is preferred over dodging within the interactive data visualization community.
+While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain visualization styles.
+Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection, and this may in fact explain why stacking is preferred over dodging within the interactive data visualization community.
 Let's explore this idea in more depth by comparing linked selection with a stacked vs.
@@
 #| fig-cap: "\\label{fig:display}The visual coherence of stacking vs. dodging
-#| during linked selection on the `mtcars` data set [@henderson1981]. Bars show
-#| the numbers of cars in the 4-, 6-, and 8-cylinder categories. 
-#| Dark segments represent selected cases, light segments represent unselected cases, 
-#| and the plots' contour is shown via a light outline. Plots left to right show more cars being selected. 
+#| during linked selection. Dark segments represent selected cases, 
+#| light segments represent unselected cases, 
+#| and the plots' contour is shown via a light outline. 
+#| Plots left to right show more cases being selected. 
 #| Top row: the contour of a stacked barplot remains constant throughout selection
@@
 #| changes dramatically with selection. Note the changes to the upper y-axis limit 
-#| and the absence of the lighter segment among the 6-cylinder cars
-#| in the rightmost plot (since all 6-cylinder cars are selected, 
-#| the darker segment has been stretched to fill the available space; 
-#| `ggplot2` default for handling empty subcategories)."
+#| and the absence of the lighter segment among the bar representing category B
+#| in the rightmost plot (the darker segment has been stretched to fill the available space; 
+#| `ggplot2` default for handling empty subcategories). Data source: @henderson1981."
 
@@
 mtcars$cyl <- factor(mtcars$cyl)
+levels(mtcars$cyl) <- c("A", "B", "C")
 plots <- list()
@@
 First, we propose that stacking is the superior method for *displaying* selection (see Figure \ref{fig:display}).
-In a stacked barplot, each stacked bar presents itself as a single visual entity, thanks to the four Gestalt principles: proximity, similarity, closure, and common region.
-Since the stacked segments are placed right on top of each other, and share the same width and closed border, they are perceived as part of a unified whole (the bar).
-This remains true throughout selection - no matter how the heights of the highlighted segments within a stacked bar change, the height of the whole bar remains constant, and so does the overall outline of the plot.
+Thanks to the four Gestalt principles of proximity, similarity, closure, and common region, in a stacked barplot, each bar presents itself as a single visual entity.
+In other words, since the stacked segments are placed right on top of each other, and share width and a closed border, they are perceived as part of a unified whole (the bar).
+This remains true throughout selection - no matter how the heights of the highlighted segments change, the height of the whole stacked bar remains constant, and so does the overall outline of the plot.
 
@@
 This leads to several design challenges.
-First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as more cases are added or removed.
+First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as cases are added or removed.
 This creates distracting visual clutter that makes it difficult to track changes over time.
-Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), we need to decide how the handle the upper y-axis limit in response to selection.
-One option is to make the upper y-axis limit dynamic.
-If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the upper y-axis limit provides each time selection occurs.
-Alternatively, we can assign a static value to the upper y-axis limit, however, then we risk the bar segments growing outside of the bounds of the plotting region.
+Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), it is not clear how to handle the upper y-axis limit.
+On one hand, we could make the upper y-axis limit dynamic and have it change in response to selection.
+If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the limit provides each time selection occurs.
+Alternatively, we can set the limit to a static value, however, then we are at the risk of the bar segments growing outside of the bounds of the plotting region.
 
@@
 
-Stacked barplots offer several other advantages over dodged bars within the context of linked selection.
+Stacked bars may offer several other advantages over dodged bars within the context of linked selection.
 First, we propose that stacking also provides a simpler mental model for *selection*.
@@
 @cook2023; @vegalite2022; @vanderplas2018; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @swayne2003].
-In dodged barplots, on the other hand, the expectation is less clear: when we click on a bar segment in a dodged barplot, will we select the entire category or just the cases corresponding to the segment?
+On the other hand, with dodged bars, the expectation is less clear: when we click on a segment in a dodged barplot, do we select the entire category or just the cases corresponding to the segment?
 Second, in a stacked barplot, the bar width and inter-bar gap are constant throughout selection.
-In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty.
-For example, when no cases are selected, should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink these in response to selection?
+In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty - should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink them in response to selection?
+If we do the former, the barplot may show asymmetric gaps and illusory clustering due to the absence of some segments.
+If we do the latter, we break the principle that the area of the bar should be proportional to the number of cases it represents [@cleveland1985].
 Finally, by presenting fewer visual entities on the screen, stacked bars may also reduce cognitive load [@sweller2019; for a non-technical overview, see @knaflic2015].
 
-To summarize, stacked bars respond gracefully to selection.
-In contrast, dodging produces several undesirable behaviors when combined with selection: it produces jagged objects with non-uniform outline that changes dramatically with selection.
+To summarize, stacked bars exhibit a more graceful and visually pleasing behavior when combined with selection.
+In contrast, dodging produces non-uniform objects with jagged outline that changes significantly with selection.
 These difference may explain why stacking is still popular in the interactive data visualization community.
@@
 
-While data visualizations are made up of geometric objects such as points, lines, or areas, these objects alone are not enough to constitute data visualizations.
-A coloured rectangle holds no meaning in and of itself; a canvas filled with such rectangles is just an image (see Figure \ref{fig:magritte}).
-To turn an image of rectangles into a plot, we need to make sure that each rectangle accurately represents some underlying data.
+Data visualizations are made up of geometric objects such as points, lines, or areas.
+However, these objects alone hold no meaning.
+Without context, an image filled with coloured rectangles is just that - an image (see Figure \ref{fig:magritte}).
+To turn the into a plot, we need to make sure that each rectangle accurately represents some underlying data.
 
@@
 #| fig-align: "center"
-#| fig-cap: "An image is not the thing it represents.
-Left: The Treachery of Images by René Magritte (1929) makes the point that an image of a thing is different from the thing itself.
-Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."
+#| fig-cap: "A representation is not the thing itself.
+Left: The Treachery of Images by René Magritte (1929) makes the point that a painting of a pipe is not the thing it represents (a pipe).
+Right: similarly, an filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."
 
@@
 
-However, we can rarely just "throw" data at the computer screen, in its raw form.
-Instead, we often first need to translate it into a format that the graphics device can understand.
+Thus, we always have to start with the data.
+But the raw data is rarely something we can just "throw" at the computer screen.
+Instead, we often first need to massage and translate it into a format that can be understood by the graphics device.
 In most data visualization packages, this is done in two steps: aggregation and scaling.
@@
 For example, as was discussed previously, in a barplot, data points are grouped by the levels of the x-axis variable, and each group is then summarized by some descriptive statistic, such as count, sum, mean, or maximum.
-Similarly, in a typical histogram, the x-axis variable is divided into bins, and each bin is then summarized by the number of cases that fall inside it.
+Likewise, in a typical histogram, the x-axis variable is divided into bins, and each bin is then summarized by the number of cases that fall inside it.
 
@@
 For instance, in a typical scatterplot, the values of the x-axis variable are mapped to the x-position, such that points corresponding to small values of the x-axis variable are placed near the left-hand side of the plot (and larger values are place near the right-hand side).
-These mappings may be subject to non-linear transformations, such as log or square-root, and may also apply to discrete values such as barplot categories.
+These mappings may be subject to non-linear transformations, such as log or square-root, and may also translate discrete values, such as barplot categories.
 Scales and coordinate systems are a rich subject and we have barely scratched the surface here; interested reader is advised to see @wilkinson2012, @ziemkiewicz2009, or @hotz2020.
@@
 
-What does stacking mean in the context of the statistics we are trying to represent?
-To a novice, stacking may appear as a purely graphical operation.
+> “This system cannot produce a meaningless graphic, however.
+This is a strong claim, vulnerable to a single counter-example.
+It is a claim based on the formal rules of the system, however, not on the evaluation of specific graphics it may produce.”
+>
+> “Some of the combinations of graphs and statistical methods may be degenerate or bizarre, but there is no moral reason to restrict them.”
+>
+> @wilkinson2012, The Grammar of Graphics, pp. 15 and 112.
+
+But what does stacking mean in the context of the statistics we are trying to represent?
+To a data visualization novice, stacking may appear as a purely graphical operation.
 If they attempt to implement a data visualization pipeline from scratch, they may leave stacking to the very end, after the statistical summaries have already been translated into screen coordinates.
 
-This view of stacking as a purely graphical operation has a certain appeal.
+The view of stacking as a purely graphical operation has its appeal.
 Many modern data visualization systems follow the grammar-based model of visualization [@mcnutt2022; @kim2022; @vanderplas2020; @wickham2010; @satyanarayan2014; @satyanarayan2016; @wilkinson2012].
-This model emphasizes compositionality, allowing plots to be built from independent, modular components, which the users can combine freely.
+This model emphasizes compositionality, and enables the user to build plots from independent, modular components.
 For example, in the popular `ggplot2` library [@wickham2010], plots are built by combining components such as geometric objects (called `geoms`) and statistical summaries (`stats`).
-These grammar-based approach offers a great deal of expressiveness, and has gained widespread popularity as a result.
-
-Thus, it may be tempting to treat stacking just as another modular component of a visualization, such that we may be able to stack any geometric objects and statistics we like.
-Unfortunately, this approach is deeply flawed.
-Many data visualization experts have issued explicit warnings about this:
+This grammar-based approach offers a great deal of flexibility and expressiveness, and has gained widespread popularity as a result.
+
+Thus, it may be tempting to treat stacking just as another modular component of a visualization, such that we may be able to freely stack any geometric objects and statistics we like.
+Unfortunately, there is a deep flaw in this approach.
+Many data visualization experts have warned about this:
 
@@
 As outlined by Wills, while the sum of grouped counts produces a valid overall count, the sum of group means does not yield a meaningful statistic.
-Even averaging group means, as noted by @wu2022, does not produce a valid summary - the mean of group means differs from the grand mean (unless by coincidence).
-This problem is extends beyond barplots; for example, concerns about the statistical validity of stacked density plots have also been raised recently [@pu2020].
-
-What exactly makes stacking sums acceptable but stacking other statistics such as means not?
+Even averaging group means, as noted by @wu2022, does not produce a valid summary statistic - the mean of group means differs from the grand mean (unless by coincidence).
+This problem is general and applies beyond barplots; for example, concerns about the statistical validity of stacked density plots have also been recently raised [@pu2020].
+
+What exactly makes stacking counts and sums acceptable but stacking other statistics such as means not?
 Wilke, Wills, and Wu et al.
 all argue that the aggregated value should be "meaningful".
-In the case of sums, perhaps another way to phrase this is: "the whole is equal to the sum of its parts".
+In the case of sums, there is perhaps a colloquial way to put this: "the whole is equal to the sum of its parts".
 Or, more specifically:
@@
 
-Then, we find there are other mathematical operators that also behave this way.
+Then, we find there are other mathematical operators that satisfy this condition.
 For example, the product of products or the maximum of maximums are also valid overall statistics, see Figure \ref{fig:sumprodmax}.
@@
 The short treatment here draws primarily from @fong2019, @lawvere2009, @baez2023, and @milewski2018.
-For a particularly accessible introduction, readers are encouraged to seek out @fong2019 and @lawvere2009.
+For a particularly accessible introduction, readers are encouraged to seek out @fong2019 or @lawvere2009.
 
 While the term "monoid" may sound intimidating, the concept is really quite simple.
-A monoid represents the idea of a "whole equal to the sum of its parts", if we relax our idea about what it means to "sum".
+A monoid represents a "whole equal to the sum of its parts", if we relax our notion of what it means to "sum".
 More formally, a monoid is a tuple $(M, e, \otimes)$ consisting of:
@@
 
-In plain words, when we have a monoid, we have some set of elements $m \in M$ and a way to combine them $\otimes$, such that the order in which we do the combining does not matter.
-Further, among the set of elements $M$, we also have a special element $e$, called the "neutral" or "identity" element, that, when combined with any other element $m \in M$, does nothing (and simply yields back $m$).
+In plain words, when we have a monoid, we have some set of elements ($m \in M$), and a way to combine these elements ($\otimes$), such that the order in which we do the combining does not matter (associativity).
+Further, among the elements, we also have a special element ($e$), called the "neutral" or "identity" element, that, when combined with any other element, does nothing.
 This identity element is always unique [the proof is straightforward, for reference see e.g.
@@
 Let's return to the book example from Section \ref{sec:intro}.
-Is it a monoid?
+Does it represent a monoid?
 First of all, what is our set $M$?
-This could be the set of all book stacks we can form by combining books in our bookshelf.
+This could be the set of all book stacks we can form by combining books in our bookshelf, including an "empty" stack of zero books, "short" stacks made up of a single book each, and the tallest stack formed by taking all of the books in the bookshelf.
 Second, do we have a neutral element?
-If we are a bit creative, we can imagine an "empty" stack of zero books.
-Let's use that as our identity element.
+Sure, this is the empty stack of zero books.
 Third, what is our binary operation?
-This is just the operation of taking one stack of books and putting it on top of another stack.
-
-Now we just need to verify that our structure conforms to the two rules.
+This is just the operation of taking one stack of books and putting it on top of another.
+
+Now we just need to verify that our example conforms to the two rules.
 First, is the operation associative?
-Clearly the answer is "yes", since a stack of detective novels will be the same height regardless of whether we put the novels by Agatha Christie on top of Jo Nesbø novels or vice versa.
+Yes, since a stack of detective novels will be the same height regardless of whether we put the novels by Agatha Christie on top of Jo Nesbø books or vice versa.
 Second, is the operation unital?
-Again, this is self-evidently true, since adding zero books on top of a stack does not change its height.
+Also yes, since adding zero books on top of a stack does not change its height.
 Therefore, we do have a monoid.
 
-We can also describe the book example a bit more formally.
+We can describe the book example a bit more formally.
 Specifically, summation on natural numbers $(\mathbb{N}, 0, +)$ is a typical example of a monoid, because, as we have shown, it is associative and unital.
-Other examples of monoids include the previously mentioned products of real numbers $(\mathbb{R}, 1, \times)$, and the min and max operators $(\mathbb{R}, \infty, \min)$ and $(\mathbb{R}, -\infty, \max)$.
+Other typical examples of monoids include the previously mentioned products of real numbers $(\mathbb{R}, 1, \times)$, and the min and max operators $(\mathbb{R}, \infty, \min)$ and $(\mathbb{R}, -\infty, \max)$.
 As a counterexample, exponentiation is not a monoid, since it is not associative: $x^{(y^z)} \neq (x^y)^z$.
 
-The definition of a monoid is quite broad however, and applies to more exotic structures than just operations on numbers.
+The definition of a monoid is quite broad however, and encompasses more exotic structures than just operations on numbers.
 For example, the set of booleans $\mathbb{B} = \{ True, False \}$ equipped with either of the logical operators $\mathbf{AND}$ or $\mathbf{OR}$ is also a monoid, and so is the multiplication of $n \times n$ square matrices.
@@
 
-Even very "non-number-like" operations such as concatenation of strings can be a monoid:
+Even very "non-number-like" things such as strings can be a monoid:
 
@@
 Groups are another fundamental concept in mathematics, representing the idea of symmetry and reversible transformations.
-While introducing two algebraic concepts in quick succession may seem like a big ask, a group is really just a monoid with the additional property.
+While introducing two algebraic concepts in quick succession may seem like a big ask, a group is essentially just a monoid with one additional property.
 Specifically, for all $x, y, z \in M$, we have:
@@
 
-In simpler terms, a group embodies the idea that "the whole is equal to the sum of its parts, *in a non-destructive way*".
+In simple terms, a group embodies the idea that "the whole is equal to the sum of its parts, *in a non-destructive way*".
 That is, when we combine two elements, we can always recover either one by "subtracting" the contribution of the other.
@@
 
-Turning back to the book example, we can remove Agatha Christie novels off the detective novel stack and recover the height of the rest of the detective novels stack.
-Thus, summation on natural numbers is a group: $2 + 3 = 5 \implies 5 - 2 = 3$ (technically speaking, we now have to identify the underlying set as integers $\mathbb{Z}$).
+Turning back to the book example, we can remove Agatha Christie novels off the detective novel stack and recover the height of the rest of the books.
+Thus, summation on natural numbers is a group: $2 + 3 = 5 \implies 5 - 2 = 3$ (technically, we now have to identify the underlying set as the set of integers $\mathbb{Z}$).
 But, for example, the min and max operators lack an inverse and so do not constitute groups: if $\max{(x, 6)} = 6$, then there is no way to recover $x$ from the combined result ($6$) alone.
@@
 We now arrive at the core idea of the present paper.
-Suppose that we have some data set $D$, a way of summarizing this data via some associative binary operation $\otimes$, and a neutral value with respect to this operation $e$.
+Suppose that we have some data set $D$, a way of summarizing this data set via an associative binary operation $\otimes$, and a neutral value with respect to this operation $e$.
 In other words, $(D, \otimes, e)$ forms a monoid.
@@
 
-Now comes the most important idea: combining summaries of two disjoint subsets of the data set gives us the same result as summarizing the union of the original subsets:
+Now comes the key concept: if the conditions above hold, then combining summaries of two disjoint subsets of the data set gives us the same result as summarizing the union of the original subsets:
 
@@
 
-Another way to show the same idea is the following commutative diagram:
+Another way to show this is the following commutative diagram:
 
@@
 $$
-The top and bottom triangles in the diagram both indicate associativity: in the top triangle, we can either first take the union of $A$ and $B$ and then union that with $C$, or immediately union $A$ with $B \cup C$ and the result will be identical.
+The top and bottom triangles in the diagram both show associativity: in the top triangle, we can either first take the union of $A$ and $B$ and then union the result with $C$, or immediately union $A$ with $B \cup C$, and the result will be identical.
 The same applies for the monoidal summary in the bottom triangle.
 By saying that the diagram commutes, we mean that any two parallel paths (the same start and endpoint) always produce the same result.
-Practically, this means we can either first summarize each set and then combine the summaries, or first union the sets and then summarize the union, or do some mix of the two, and we will always end up with a valid summary.
+Practically, this means we can either first summarize each set and then combine the summaries, first union the sets and then summarize the union, or do some mix of both, and we always end up with a valid summary.
 As a sidenote, the diagram above actually corresponds to something called a functor or monoid homomorphism, however, that is beyond the scope of the present paper [interested reader should see @fong2019; @lawvere2009].
 
-But this is precisely the idea of a stackable statistic.
+But what we have now is precisely the idea of a stackable statistic.
 If our statistic conforms to the diagram, then the combined summary of subsets of the data is always a valid summary of the whole data set.
@@
 $F(A)$, $F(A \cup B)$, and $F(A \cup B \cup C)$.
-This is fine, for example, when we do linked selection with a single group and compare the summary on the highlighted subset ($A$) vs.
+This is fine, for example, when we do linked selection with a single highlighted group and want to compare the summary on the highlighted subset ($A$) vs.
 the entirety the data ($A \cup B$).
+The resulting statistics give us what we need.
 
@@
 
-Importantly, $F(A) \otimes F(B)$ is still a valid summary of $A \cup B$ even when the inverse does not exist.
-It is just that $F(A) \otimes F(B)$ may "forget" the information contained in either of its two parts.
-Note also that this is not a problem of different permutations of parts producing different sequences of accumulated summaries - this will happen regardless of whether or not there is an inverse.
+Importantly, even when the operation does not have an inverse, $F(A) \otimes F(B)$ is still a valid summary of $A \cup B$.
+It is just that $F(A) \otimes F(B)$ may "forget" the information contained in either of its two constituent parts.
+Note also that this is not a problem of different permutations of parts producing different sequences of accumulated summaries - this will happen regardless of whether there is an inverse or not.
 
@@
 #| fig-height: 5
-#| fig-cap: "\\label{fig:summax} Monoids can \"collapse\" information about disjoint parts.
-Top: since sum is a group with an inverse (minus), we can recover the value of each disjoint part by subtracting the heights of the preceding segments (i.e.
+#| fig-cap: "\\label{fig:summax} Monoids can \"collapse\" information about disjoint subsets.
+Top: since sum is a group with an inverse (minus), we can recover the value of each part by subtracting the heights of the preceding segments (i.e.
 the part order does not matter).
@@
 
-To summarize, whether our statistic is a monoid or a group determines what kinds of comparisons we can make on stacked objects:
+To summarize, whether our statistic is a monoid or a group determines what kinds of comparisons we can make with stacked objects:
 
@@
 This becomes important when implementing interactive data visualization features such as linked selection.
-Specifically, if we want to compare a single selected group vs.
-the rest of the data, then it's enough for our summary statistic to be a monoid.
-If we want to implement multiple selection groups, then our statistic needs to have an inverse as well.
-If we fail either of these conditions, then we may produce data representations which are either inaccurate (in the case of a non-monoidal statistic), or accurate but not "full" (in the case of a monoid but no inverse).
+Specifically, if we want to represent a single highlighted group vs.
+the rest of the data via stacked objects, then it's enough for our summary statistic to be a monoid.
+If we want to represent multiple highlighted groups, then our statistic needs to have an inverse as well.
+If we fail either of these conditions, we may produce representations which are either inaccurate (in the case of non-monoidal statistics), or accurate but not "full" (in the case of monoids but no inverse).
 
@@
 The connection between summary statistics and algebraic structures is important because it allows us to predict which new combinations of summaries and figures will "work".
-For example, while not generally thought of as a summary statistic, the convex hull algorithm [see e.g.
-@barber1996] provides a way to reduce a set of points to a simpler "summary", see Figure \ref{fig:convexhull}.
-We can recast the operation of finding the convex hull as a monoid: starting with an empty set, we iterate through the data, adding points if they lie outside of the current hull (and removing any points in the interior of the new hull).
-This may not be the most computationally efficient method for finding convex hulls, however, since it does meet the definition of a monoid, this tells us *what convex hulls can be used for*.
-If we combine two convex hulls, we know that the new hull is a valid summary for the union of the underlying data points.
+For example, while not generally thought of as a summary statistic, the convex hull [see e.g.
+@barber1996] provides a way to reduce a set of data points into a simpler "summary", see Figure \ref{fig:convexhull}.
+We can recast convex hull algorithm as a monoid: starting with an empty set, we iterate through the data, adding points if they lie outside of the current hull (and removing any points in the interior of the new hull).
+This may not be the most computationally efficient method of finding convex hulls, however, since it does meet the definition of a monoid, this tells us *what convex hulls can be used for*.
+If we combine two convex hulls, we know that the new hull will be a valid summary for the union of the underlying data points.
 In turn, this allows us to implement interactive features such as single-group highlighting/selection, safe in the knowledge that the underlying structure is preserved.
@@
 This tells us that merging two convex hulls may irrevocably collapse information.
-As such, we cannot use convex hulls to display multiple selections, without resorting to some ad-hoc methods (e.g.
-keeping separate representation of the hulls and using partial transparency).
+As such, we cannot safely use convex hulls to display multiple selection groups, without resorting to some ad-hoc methods.
 
@@
 #| fig-cap: "\\label{fig:convexhull}Convex hull can be used as a valid representation of single-group selection but not multi-group selection.
-Both plots show data from Palmer penguins data set (Horst 2020).
-The light polygon shows the convex hull of the whole dataset, the dark polygon shows the hull of the Gentoo species, the medium-brightness polygon shows the combined hull of Gentoo and Chinstrap species, and the white dashed line shows the outline the hull of the Chinstrap penguins alone.
+The light polygons shows the convex hull of the whole dataset, the dark polygon shows the hull of one selection group, the white dashed line shows the hull of a second selection group, and the medium-brightness polygon shows the combined hull of two selection groups.
 Points show individual cases.
-Left: since convex hull is a monoid, we can use it to show single group selection/highlighting.
-Right: because of the lack of an inverse, we cannot guarantee that hull of the Chinstrap penguins (white-dashed line) will be recoverable from the combined hull of Gentoo and Chinstrap - we do not know how many interior points were \"forgotten\"."
+Left: since convex hull is a monoid, we can use it to show single highlighted group.
+Right: because of the lack of an inverse, we cannot guarantee that we will be able to recover either of the original hulls from the combined hull - we do not know how many interior points were \"forgotten\".
+The source of the data is the Palmer penguins data set, Horst 2020."
 #| message: false
@@
   guides(col = "none") +
-  labs(x = "Body mass", y = "Flipper length") +
+  labs(x = NULL, y = NULL) +
   theme(

### 2024-10-23 paper.tex
Rewrote some sections

@@
 \hypersetup{
-  pdftitle={No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction},
+  pdftitle={No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction},
   pdfkeywords={Interactive data visualization, category theory, abstract
@@
 {
-  \title{\bf No More, No Less Than Sum of its Parts: The Algebra of
-Graphics, Statistics, and Interaction}
+  \title{\bf No More, No Less Than Sum of its Parts: Monoids and The
+Algebra of Graphics, Statistics, and Interaction}
 
@@
   \begin{center}
-    {\LARGE\bf No More, No Less Than Sum of its Parts: The Algebra of
-Graphics, Statistics, and Interaction}
+    {\LARGE\bf No More, No Less Than Sum of its Parts: Monoids and The
+Algebra of Graphics, Statistics, and Interaction}
   \end{center}
@@
 theory: groups and monoids. Specifically, if we want our graphics to
-support interactive features which break the data into multiple parts
-and then combine them back together (such as in the case of linked
-selection), then we need the statistics in our plots to fulfill certain
-algebraic properties. Ultimately, understanding this hidden structure
-may empower data visualization experts to build more flexible and
-expressive interactive data visualization systems.
+support interactive features which split our data into parts and then
+combine these parts back together (such as linked selection), then we
+need the statistics in our plots to fulfill certain algebraic
+properties. Ultimately, understanding this hidden structure may empower
+data visualization experts to build more flexible and expressive
+interactive data visualization systems.
 \end{abstract}
@@
 The essence of data visualization is comparison. When we visualize, we
-split our data into parts, summarize each part one or more summary
+split our data into parts, summarize each part by one or more summary
 statistics, encode the summaries into visual attributes such as
-position, size, or color, and finally use these attributes to draw
+position, size, or color, and, finally, use these attributes to draw
 geometric objects
@@
 based on the levels of some categorical variable, count the number of
-cases in each part, and finally represent the counts by drawing bars of
-the corresponding heights. In this way, each bar becomes a visual
-representation of its own ``small data set'', a subset of the original
-data.
-
-To ground this idea in a more concrete example that will be used
-throughout this paper, imagine going through your bookshelf, sorting the
-books by genre, and stacking them on top of each other. You would
-construct a physical version of the barplot, forming one ``bar'' (stack)
-per genre: one bar for sci-fi novels, one bar for romance, one bar for
-detective fiction, and so on. Even from this very simple example,
-interesting mathematical properties jump out: for example, a single book
-cannot go into two stacks (the subsets of the data are \emph{disjoint})
-and if all of the stacks are combined back together, we should end up
-with the bookshelf we started with \citep[the subsets are
-\emph{surjective} and cover the entirety of the data set,
-see][]{ziemkiewicz2009}.
-
-Now, just as we can learn about the contents of our library by
-organizing books in the way we just described, we can learn about our
-data by splitting it into parts. A natural question then arises: can we
-learn more by subdividing our data into even smaller parts and comparing
-these? For instance, in our book example, we can split each stack of
-books into two smaller stacks based on the gender of the author. Each
-smaller stack will then represent the unique combination (Cartesian
-product) of the two categorical variables (genre and gender). Moreover,
-there arises a hierarchical relationship between the stacks: by
-combining two genre-gender stacks (child components) together, we can
-recover the original genre stack (parent component). Similar situation
-often arises in interactive graphics, where we use techniques such as
-linked selection to highlight specific subsets of our data. Either way,
-by forming these finer partitions, we can represent our data with
+cases in each part, and finally represent these counts by drawing bars
+of corresponding heights. In this way, each geometric object (bar)
+becomes a representation of its own small data set, a subset of the
+original data.
+
+To ground this idea in a concrete example that will be used throughout
+this paper, imagine going through your bookshelf, sorting the books by
+genre, and stacking them on top of each other. You will construct a
+physical version of the barplot, forming one ``bar'' (stack) per genre:
+one bar for sci-fi novels, one bar for romance, one bar for detective
+fiction, and so on. Even this very simple example reveals some
+interesting mathematical properties. For example, a single book cannot
+go into two stacks (the subsets of the data are \emph{disjoint}) and if
+all of the stacks are combined back together, we should end up with the
+bookshelf we started with \citep[the subsets are \emph{surjective} and
+cover the entirety of the data set, see][]{ziemkiewicz2009}.
+
+Now, just as we can learn about the contents of our library by sorting
+books into stacks, we can learn about our data by splitting it into
+parts. A natural question then arises: can we learn more by subdividing
+our data into even smaller parts? For instance, in our book example, we
+can split each stack of books into smaller stacks based on the gender of
+the author. Each smaller stack then represents the unique combination
+(Cartesian product) of the two categorical variables (genre and gender).
+Moreover, there arises a hierarchical relationship between the stacks:
+by combining genre-gender stacks (child components), we can recover the
+original genre stack (parent component). Similar situation often arises
+in interactive graphics, where we use techniques such as linked
+selection to highlight specific subsets of our data. Either way, by
+forming these finer partitions of our data, we can represent it with
 greater granularity and discover trends that may otherwise hide in the
@@
 components, we need to represent the information in a way that visually
-preserves the hierarchical nature of this relationship. In data
-visualization, there are two popular methods for doing this, known as
-``stacking'' and ``dodging'' (see Figure \ref{fig:playfair}). These
-methods have existed at least since the time of William Playfair
+preserves this hierarchical relationship. In data visualization, there
+are two popular methods for doing this, known as ``stacking'' and
+``dodging'' (see Figure \ref{fig:playfair}). These methods have existed
+at least since the time of William Playfair
 \citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
-briefly illustrate them on the example of the classical barplot. In a
-stacked barplot, bar segments representing the product of the two factor
+briefly illustrate them on the example of the barplot. In a stacked
+barplot, bar segments representing the product of the two factor
 variables are plotted vertically on top of each other (``stacked''; the
@@
 In contrast, in a dodged barplot, bars are plotted side-by-side, as
-clusters of bars. Finally, there is also a third popular method, known
+``clusters''. Finally, there is also a third more recent method, known
 as layering, where we plot bar segments in separate graphical layers and
-use partial transparency to mitigate overplotting (this technique may
-have been a bit challenging for William Playfair but has been made
-simple with computers).
+use partial transparency to mitigate overplotting.
 
@@
 
-\caption{Examples of the three methods for dealing with bars that have subdivided into segments.
+\caption{Examples of three methods for dealing with bars that have split into child components (segments).
 Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822).
@@
 charts. Specifically, since the lower y-axis coordinate of a stacked
-segment is pushed up by the cumulative height of the segments below it,
-it becomes difficult to accurately compare segments' length, both within
+segment is pushed up by the cumulative height of the segments below, it
+becomes difficult to accurately compare segments' length, both within
 and across bars \citep{cleveland1984}. Subsequent research has
 independently validated these findings and expanded upon them \citep[see
-e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
-statistical legibility of stacking has led some authors to advise
-against its use altogether \citep{kosara2016, wilke2019}, or at least
-urge caution \citep[see e.g.][]{byron2008, cairo2014, franconeri2021}.
+e.g.][]{heer2010, thudt2016, quadri2021}. Due to this suboptimal
+statistical legibility, many data visualization researchers have urged
+caution about stacking \citep[see
+e.g.][]{byron2008, cairo2014, franconeri2021}, and some have even
+discouraged its use altogether \citep{kosara2016, wilke2019}.
 
@@
 
-We argue that not only is stacking a useful technique, but that it can
-in fact offer valuable insights about the mathematical structure
-underlying our visualizations. To make the strongest case possible, we
-bring forward evidence from several levels of the visualization process.
-We start by discussing the properties of visual perception and how they
-apply to both static and interactive visualizations. Next, we explore
-the process of computing statistical summaries that form the basis of
-our plots. Finally, we dive into the algebraic properties of the
-functions underlying these statistical summaries. Throughout the text,
-we use the example of the humble barplot to ground our discussion.
+We argue that not only is stacking a useful technique, despite its
+limitations, but that it can also offer valuable insights into the
+mathematical structure underlying our visualizations. To make the
+strongest case possible, we bring forward evidence from several levels
+of the visualization process. We start by discussing the properties of
+visual perception and how they apply to both static and interactive
+visualizations. Next, we explore the process of computing statistical
+summaries that form the basis of our plots. Finally, we dive into the
+algebraic properties of the functions underlying these statistical
+summaries. Throughout the text, we use the example of the humble barplot
+to ground our discussion.
 
 Our goal is to convince you that, in order to produce coherent and
-efficient interactive graphics, we simply cannot treat graphics,
-statistics, and interaction as independent components. Instead we need
-to consider them holistically, within the context of their respective
-visual and mathematical properties. Understanding this deep web of
-interdependence may in turn empower us to innovate and build new kinds
-of robust and expressive interactive visualizations.
+well-behaved interactive visualizations, we simply cannot treat
+graphics, statistics, and interaction as independent components. Instead
+we need to consider them holistically, within the context of their
+respective visual and mathematical properties. Understanding this deep
+web of interdependence may in turn empower us to innovate and build new
+kinds of robust and expressive interactive visualizations.
 
@@
 Over the course of our evolutionary history, the human brain has evolved
-a sophisticated pipeline for processing visual information. This
-pipeline has been the subject of a great body of research in cognitive
-neuroscience
+a sophisticated network for processing visual input. This visual
+processing pipeline has been the subject of a great body of research in
+cognitive neuroscience
 \citetext{\citealp{goebel2004}; \citealp{knudsen2020}; \citealp[for a
-brief overview, see][20]{ware2019}}. Many of the findings from this
-research are directly applicable to data visualization. For example,
-certain particularly salient visual stimuli are perceived automatically,
-rapidly (\textless10 ms), in parallel, and at a constant speed
-regardless of the number of distractors \citep{treisman1985}. This
-phenomenon, known as pre-attentive processing, is key for data
-visualization, as the right use of visual attributes such as color,
-contrast, or shape can help us design figures which communicate insights
-much more effectively \citep{ware2019}.
-
-However, many of the findings from visual perception research are quite
-specific; to guide effective visualization, we also need some general,
+brief overview, see][20]{ware2019}}, and many of its findings are
+directly applicable to data visualization. For example, certain
+particularly salient visual stimuli are perceived automatically, rapidly
+(\textless10 ms), in parallel, and at a constant speed regardless of the
+number of distractors \citep{treisman1985}. This phenomenon, known as
+pre-attentive processing, is key for data visualization, as the right
+use of visual attributes such as color, contrast, or shape can help us
+design figures which communicate insights much more effectively
+\citep{ware2019}.
+
+However, many of the visual perception findings are quite specific and
+niche; to guide effective visualization, we also need some general,
 high-level rules. One such set of high-level rules are the Gestalt
@@
 These principles formulate a set of fundamental ``laws'' related to the
-perception of patterns, primarily visual. They were discovered by the
+perception of (primarily visual) patterns. They were discovered by the
 founders of the Gestalt school of psychology early in the 20th century,
@@
 as single, unified objects. Finally, the figure and ground principle
-states we tend to see one part of a scene as the central object (the
-``figure'') while the rest is perceived as the background (the
+states that we tend to perceive one part of a visual scene as the
+central object (the ``figure'') and the rest as background (the
 ``ground''). This can be modulated by other Gestalt principle: for
@@
 that clouds of similarly shaped and colored points will be perceived as
-discrete groups by the observer. In other cases, however, our design
-choices can have a significant impact. For instance, another way to have
-objects to be perceived as part of a group is to draw a closed shape
-around them \citep{cairo2012}. Alternatively, if we want to highlight a
-particular part of a data set, we can leverage the figure and ground
-principle and use attributes such as color and saturation to make it
-stand out \citep[for specific examples,
-see][34]{franconeri2021, wilke2019}.
+discrete groups. In other cases, however, our design choices can have a
+significant impact. For instance, another way to make objects be
+perceived as part of a group is to draw a closed shape around them
+\citep{cairo2012}. Alternatively, if we want to highlight a particular
+part of a data set, we can leverage the figure and ground principle and
+use attributes such as color and saturation to make it stand out
+\citep[for specific examples, see][34]{franconeri2021, wilke2019}.
 
@@
 inadvertent optical illusions caused by dense stripe patterns, and the
-overuse of superfluous graphical elements he termed ``chartjunk''.
+overuse of superfluous graphical elements termed ``chartjunk''.
 Similarly, \citet{cairo2019} highlights many cases of graphics that
@@
 time and violate them in another. In fact, interactive visualizations
-can even differ in how well they leverage design principles throughout
-time: the animations that interactions induce may themselves conform to
-good or poor design \citep[see][]{hullman2013, ware2019}.
-
-Thus, interactivity inherently places further constraints the design of
-figures. If we want well-designed interactive figures, we need to ensure
-that they conform to good design principles across all states induced by
-user-interaction. This may not always be easy. We will investigate this
-idea in the following section, on the example of linked selection.
+can even differ in how well they leverage design principles
+\emph{across} time: the animations that interactions induce may
+themselves conform to good or poor design
+\citep[see][]{hullman2013, ware2019}.
+
+Thus, interactivity inherently places additional constraints the design
+of our figures. If we want well-designed interactive figures, we need to
+ensure that they conform to good design principles across all states
+induced by user-interaction. This may not always be easy. We will
+investigate the design challenged posed by interaction in the following
+section, on the example of linked selection.
 
@@
 While these constraints may not seem too restrictive, we argue that they
-pose significant challenges for certain types of plots and visualization
-styles. Specifically, some kinds of plots may produce undesirable visual
-behavior when combined with linked selection. In fact, this may explain
-why stacking is preferred over dodging within the interactive data
-visualization community. Let's explore this idea in more depth by
+pose significant challenges for certain visualization styles.
+Specifically, some kinds of plots may produce undesirable visual
+behavior when combined with linked selection, and this may in fact
+explain why stacking is preferred over dodging within the interactive
+data visualization community. Let's explore this idea in more depth by
 comparing linked selection with a stacked vs.~dodged barplot.
@@
 \caption{\label{fig:display}The visual coherence of stacking vs.~dodging
-during linked selection on the \texttt{mtcars} data set
-\citep{henderson1981}. Bars show the numbers of cars in the 4-, 6-, and
-8-cylinder categories. Dark segments represent selected cases, light
+during linked selection. Dark segments represent selected cases, light
 segments represent unselected cases, and the plots' contour is shown via
-a light outline. Plots left to right show more cars being selected. Top
+a light outline. Plots left to right show more cases being selected. Top
 row: the contour of a stacked barplot remains constant throughout
@@
 the upper y-axis limit and the absence of the lighter segment among the
-6-cylinder cars in the rightmost plot (since all 6-cylinder cars are
-selected, the darker segment has been stretched to fill the available
-space; \texttt{ggplot2} default for handling empty subcategories).}
+bar representing category B in the rightmost plot (the darker segment
+has been stretched to fill the available space; \texttt{ggplot2} default
+for handling empty subcategories). Data source: \citet{henderson1981}.}
 \end{figure}
@@
 First, we propose that stacking is the superior method for
-\emph{displaying} selection (see Figure \ref{fig:display}). In a stacked
-barplot, each stacked bar presents itself as a single visual entity,
-thanks to the four Gestalt principles: proximity, similarity, closure,
-and common region. Since the stacked segments are placed right on top of
-each other, and share the same width and closed border, they are
-perceived as part of a unified whole (the bar). This remains true
-throughout selection - no matter how the heights of the highlighted
-segments within a stacked bar change, the height of the whole bar
-remains constant, and so does the overall outline of the plot.
+\emph{displaying} selection (see Figure \ref{fig:display}). Thanks to
+the four Gestalt principles of proximity, similarity, closure, and
+common region, in a stacked barplot, each bar presents itself as a
+single visual entity. In other words, since the stacked segments are
+placed right on top of each other, and share width and a closed border,
+they are perceived as part of a unified whole (the bar). This remains
+true throughout selection - no matter how the heights of the highlighted
+segments change, the height of the whole stacked bar remains constant,
+and so does the overall outline of the plot.
 
@@
 selection occurs, the jagged contour of the dodged barplot can change
-dramatically as more cases are added or removed. This creates
-distracting visual clutter that makes it difficult to track changes over
-time. Second, since the heights of the bar segments are not bound by any
-value present in the plot (unlike the heights of the segments in the
-stacked barplot, which are bound by the height of the tallest bar), we
-need to decide how the handle the upper y-axis limit in response to
-selection. One option is to make the upper y-axis limit dynamic. If we
-do this, we ensure that the bar segments always fit the plotting region,
-however, we also lose the context that the upper y-axis limit provides
-each time selection occurs. Alternatively, we can assign a static value
-to the upper y-axis limit, however, then we risk the bar segments
+dramatically as cases are added or removed. This creates distracting
+visual clutter that makes it difficult to track changes over time.
+Second, since the heights of the bar segments are not bound by any value
+present in the plot (unlike the heights of the segments in the stacked
+barplot, which are bound by the height of the tallest bar), it is not
+clear how to handle the upper y-axis limit. On one hand, we could make
+the upper y-axis limit dynamic and have it change in response to
+selection. If we do this, we ensure that the bar segments always fit the
+plotting region, however, we also lose the context that the limit
+provides each time selection occurs. Alternatively, we can set the limit
+to a static value, however, then we are at the risk of the bar segments
 growing outside of the bounds of the plotting region.
 
-Stacked barplots offer several other advantages over dodged bars within
+Stacked bars may offer several other advantages over dodged bars within
 the context of linked selection. First, we propose that stacking also
@@
 e.g.][]{cook2023, vegalite2022, vanderplas2018, xie2014, theus2002, urbanek2003, urbanek2011, swayne2003}.
-In dodged barplots, on the other hand, the expectation is less clear:
-when we click on a bar segment in a dodged barplot, will we select the
-entire category or just the cases corresponding to the segment? Second,
-in a stacked barplot, the bar width and inter-bar gap are constant
-throughout selection. In a dodged barplot, it is not clear how to handle
-the gaps when some selection groups are empty. For example, when no
-cases are selected, should we leave gaps for the (currently absent)
-highlighted segments, or draw wide bars and only shrink these in
-response to selection? Finally, by presenting fewer visual entities on
+On the other hand, with dodged bars, the expectation is less clear: when
+we click on a segment in a dodged barplot, do we select the entire
+category or just the cases corresponding to the segment? Second, in a
+stacked barplot, the bar width and inter-bar gap are constant throughout
+selection. In a dodged barplot, it is not clear how to handle the gaps
+when some selection groups are empty - should we leave gaps for the
+(currently absent) highlighted segments, or draw wide bars and only
+shrink them in response to selection? If we do the former, the barplot
+may show asymmetric gaps and illusory clustering due to the absence of
+some segments. If we do the latter, we break the principle that the area
+of the bar should be proportional to the number of cases it represents
+\citep{cleveland1985}. Finally, by presenting fewer visual entities on
 the screen, stacked bars may also reduce cognitive load
@@
 
-To summarize, stacked bars respond gracefully to selection. In contrast,
-dodging produces several undesirable behaviors when combined with
-selection: it produces jagged objects with non-uniform outline that
-changes dramatically with selection. These difference may explain why
-stacking is still popular in the interactive data visualization
-community.
+To summarize, stacked bars exhibit a more graceful and visually pleasing
+behavior when combined with selection. In contrast, dodging produces
+non-uniform objects with jagged outline that changes significantly with
+selection. These difference may explain why stacking is still popular in
+the interactive data visualization community.
 
@@
 
-While data visualizations are made up of geometric objects such as
-points, lines, or areas, these objects alone are not enough to
-constitute data visualizations. A coloured rectangle holds no meaning in
-and of itself; a canvas filled with such rectangles is just an image
-(see Figure \ref{fig:magritte}). To turn an image of rectangles into a
-plot, we need to make sure that each rectangle accurately represents
-some underlying data.
+Data visualizations are made up of geometric objects such as points,
+lines, or areas. However, these objects alone hold no meaning. Without
+context, an image filled with coloured rectangles is just that - an
+image (see Figure \ref{fig:magritte}). To turn the into a plot, we need
+to make sure that each rectangle accurately represents some underlying
+data.
 
@@
 
-\caption{An image is not the thing it represents.
-Left: The Treachery of Images by René Magritte (1929) makes the point that an image of a thing is different from the thing itself.
-Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
+\caption{A representation is not the thing itself.
+Left: The Treachery of Images by René Magritte (1929) makes the point that a painting of a pipe is not the thing it represents (a pipe).
+Right: similarly, an filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
 \end{figure}
 
-However, we can rarely just ``throw'' data at the computer screen, in
-its raw form. Instead, we often first need to translate it into a format
-that the graphics device can understand. In most data visualization
-packages, this is done in two steps: aggregation and scaling.
+Thus, we always have to start with the data. But the raw data is rarely
+something we can just ``throw'' at the computer screen. Instead, we
+often first need to massage and translate it into a format that can be
+understood by the graphics device. In most data visualization packages,
+this is done in two steps: aggregation and scaling.
 
@@
 group is then summarized by some descriptive statistic, such as count,
-sum, mean, or maximum. Similarly, in a typical histogram, the x-axis
+sum, mean, or maximum. Likewise, in a typical histogram, the x-axis
 variable is divided into bins, and each bin is then summarized by the
@@
 These mappings may be subject to non-linear transformations, such as log
-or square-root, and may also apply to discrete values such as barplot
+or square-root, and may also translate discrete values, such as barplot
 categories. Scales and coordinate systems are a rich subject and we have
@@
 
-What does stacking mean in the context of the statistics we are trying
-to represent? To a novice, stacking may appear as a purely graphical
-operation. If they attempt to implement a data visualization pipeline
-from scratch, they may leave stacking to the very end, after the
-statistical summaries have already been translated into screen
-coordinates.
-
-This view of stacking as a purely graphical operation has a certain
-appeal. Many modern data visualization systems follow the grammar-based
-model of visualization
+\begin{quote}
+``This system cannot produce a meaningless graphic, however. This is a
+strong claim, vulnerable to a single counter-example. It is a claim
+based on the formal rules of the system, however, not on the evaluation
+of specific graphics it may produce.''
+
+``Some of the combinations of graphs and statistical methods may be
+degenerate or bizarre, but there is no moral reason to restrict them.''
+
+\citet{wilkinson2012}, The Grammar of Graphics, pp.~15 and 112.
+\end{quote}
+
+But what does stacking mean in the context of the statistics we are
+trying to represent? To a data visualization novice, stacking may appear
+as a purely graphical operation. If they attempt to implement a data
+visualization pipeline from scratch, they may leave stacking to the very
+end, after the statistical summaries have already been translated into
+screen coordinates.
+
+The view of stacking as a purely graphical operation has its appeal.
+Many modern data visualization systems follow the grammar-based model of
+visualization
 \citep{mcnutt2022, kim2022, vanderplas2020, wickham2010, satyanarayan2014, satyanarayan2016, wilkinson2012}.
-This model emphasizes compositionality, allowing plots to be built from
-independent, modular components, which the users can combine freely. For
-example, in the popular \texttt{ggplot2} library \citep{wickham2010},
-plots are built by combining components such as geometric objects
-(called \texttt{geoms}) and statistical summaries (\texttt{stats}).
-These grammar-based approach offers a great deal of expressiveness, and
-has gained widespread popularity as a result.
+This model emphasizes compositionality, and enables the user to build
+plots from independent, modular components. For example, in the popular
+\texttt{ggplot2} library \citep{wickham2010}, plots are built by
+combining components such as geometric objects (called \texttt{geoms})
+and statistical summaries (\texttt{stats}). This grammar-based approach
+offers a great deal of flexibility and expressiveness, and has gained
+widespread popularity as a result.
 
 Thus, it may be tempting to treat stacking just as another modular
-component of a visualization, such that we may be able to stack any
-geometric objects and statistics we like. Unfortunately, this approach
-is deeply flawed. Many data visualization experts have issued explicit
-warnings about this:
+component of a visualization, such that we may be able to freely stack
+any geometric objects and statistics we like. Unfortunately, there is a
+deep flaw in this approach. Many data visualization experts have warned
+about this:
 
@@
 averaging group means, as noted by \citet{wu2022}, does not produce a
-valid summary - the mean of group means differs from the grand mean
-(unless by coincidence). This problem is extends beyond barplots; for
-example, concerns about the statistical validity of stacked density
-plots have also been raised recently \citep{pu2020}.
-
-What exactly makes stacking sums acceptable but stacking other
-statistics such as means not? Wilke, Wills, and Wu et al.~all argue that
-the aggregated value should be ``meaningful''. In the case of sums,
-perhaps another way to phrase this is: ``the whole is equal to the sum
-of its parts''. Or, more specifically:
+valid summary statistic - the mean of group means differs from the grand
+mean (unless by coincidence). This problem is general and applies beyond
+barplots; for example, concerns about the statistical validity of
+stacked density plots have also been recently raised \citep{pu2020}.
+
+What exactly makes stacking counts and sums acceptable but stacking
+other statistics such as means not? Wilke, Wills, and Wu et al.~all
+argue that the aggregated value should be ``meaningful''. In the case of
+sums, there is perhaps a colloquial way to put this: ``the whole is
+equal to the sum of its parts''. Or, more specifically:
 
@@
 
-Then, we find there are other mathematical operators that also behave
-this way. For example, the product of products or the maximum of
+Then, we find there are other mathematical operators that satisfy this
+condition. For example, the product of products or the maximum of
 maximums are also valid overall statistics, see Figure
@@
 particularly accessible introduction, readers are encouraged to seek out
-\citet{fong2019} and \citet{lawvere2009}.
+\citet{fong2019} or \citet{lawvere2009}.
 
 While the term ``monoid'' may sound intimidating, the concept is really
-quite simple. A monoid represents the idea of a ``whole equal to the sum
-of its parts'', if we relax our idea about what it means to ``sum''.
-More formally, a monoid is a tuple \((M, e, \otimes)\) consisting of:
+quite simple. A monoid represents a ``whole equal to the sum of its
+parts'', if we relax our notion of what it means to ``sum''. More
+formally, a monoid is a tuple \((M, e, \otimes)\) consisting of:
 
@@
 In plain words, when we have a monoid, we have some set of elements
-\(m \in M\) and a way to combine them \(\otimes\), such that the order
-in which we do the combining does not matter. Further, among the set of
-elements \(M\), we also have a special element \(e\), called the
-``neutral'' or ``identity'' element, that, when combined with any other
-element \(m \in M\), does nothing (and simply yields back \(m\)). This
-identity element is always unique \citep[the proof is straightforward,
-for reference see e.g.][]{fong2019, lawvere2009}.
-
-Let's return to the book example from Section \ref{sec:intro}. Is it a
-monoid? First of all, what is our set \(M\)? This could be the set of
-all book stacks we can form by combining books in our bookshelf. Second,
-do we have a neutral element? If we are a bit creative, we can imagine
-an ``empty'' stack of zero books. Let's use that as our identity
-element. Third, what is our binary operation? This is just the operation
-of taking one stack of books and putting it on top of another stack.
-
-Now we just need to verify that our structure conforms to the two rules.
-First, is the operation associative? Clearly the answer is ``yes'',
-since a stack of detective novels will be the same height regardless of
-whether we put the novels by Agatha Christie on top of Jo Nesbø novels
-or vice versa. Second, is the operation unital? Again, this is
-self-evidently true, since adding zero books on top of a stack does not
-change its height. Therefore, we do have a monoid.
-
-We can also describe the book example a bit more formally. Specifically,
+(\(m \in M\)), and a way to combine these elements (\(\otimes\)), such
+that the order in which we do the combining does not matter
+(associativity). Further, among the elements, we also have a special
+element (\(e\)), called the ``neutral'' or ``identity'' element, that,
+when combined with any other element, does nothing. This identity
+element is always unique \citep[the proof is straightforward, for
+reference see e.g.][]{fong2019, lawvere2009}.
+
+Let's return to the book example from Section \ref{sec:intro}. Does it
+represent a monoid? First of all, what is our set \(M\)? This could be
+the set of all book stacks we can form by combining books in our
+bookshelf, including an ``empty'' stack of zero books, ``short'' stacks
+made up of a single book each, and the tallest stack formed by taking
+all of the books in the bookshelf. Second, do we have a neutral element?
+Sure, this is the empty stack of zero books. Third, what is our binary
+operation? This is just the operation of taking one stack of books and
+putting it on top of another.
+
+Now we just need to verify that our example conforms to the two rules.
+First, is the operation associative? Yes, since a stack of detective
+novels will be the same height regardless of whether we put the novels
+by Agatha Christie on top of Jo Nesbø books or vice versa. Second, is
+the operation unital? Also yes, since adding zero books on top of a
+stack does not change its height. Therefore, we do have a monoid.
+
+We can describe the book example a bit more formally. Specifically,
 summation on natural numbers \((\mathbb{N}, 0, +)\) is a typical example
 of a monoid, because, as we have shown, it is associative and unital.
-Other examples of monoids include the previously mentioned products of
-real numbers \((\mathbb{R}, 1, \times)\), and the min and max operators
-\((\mathbb{R}, \infty, \min)\) and \((\mathbb{R}, -\infty, \max)\). As a
-counterexample, exponentiation is not a monoid, since it is not
-associative: \(x^{(y^z)} \neq (x^y)^z\).
-
-The definition of a monoid is quite broad however, and applies to more
+Other typical examples of monoids include the previously mentioned
+products of real numbers \((\mathbb{R}, 1, \times)\), and the min and
+max operators \((\mathbb{R}, \infty, \min)\) and
+\((\mathbb{R}, -\infty, \max)\). As a counterexample, exponentiation is
+not a monoid, since it is not associative: \(x^{(y^z)} \neq (x^y)^z\).
+
+The definition of a monoid is quite broad however, and encompasses more
 exotic structures than just operations on numbers. For example, the set
@@
 
-Even very ``non-number-like'' operations such as concatenation of
-strings can be a monoid:
+Even very ``non-number-like'' things such as strings can be a monoid:
 
@@
 algebraic concepts in quick succession may seem like a big ask, a group
-is really just a monoid with the additional property. Specifically, for
-all \(x, y, z \in M\), we have:
+is essentially just a monoid with one additional property. Specifically,
+for all \(x, y, z \in M\), we have:
 
@@
 
-In simpler terms, a group embodies the idea that ``the whole is equal to
+In simple terms, a group embodies the idea that ``the whole is equal to
 the sum of its parts, \emph{in a non-destructive way}''. That is, when
@@
 off the detective novel stack and recover the height of the rest of the
-detective novels stack. Thus, summation on natural numbers is a group:
-\(2 + 3 = 5 \implies 5 - 2 = 3\) (technically speaking, we now have to
-identify the underlying set as integers \(\mathbb{Z}\)). But, for
+books. Thus, summation on natural numbers is a group:
+\(2 + 3 = 5 \implies 5 - 2 = 3\) (technically, we now have to identify
+the underlying set as the set of integers \(\mathbb{Z}\)). But, for
 example, the min and max operators lack an inverse and so do not
@@
 We now arrive at the core idea of the present paper. Suppose that we
-have some data set \(D\), a way of summarizing this data via some
+have some data set \(D\), a way of summarizing this data set via an
 associative binary operation \(\otimes\), and a neutral value with
@@
 
-Now comes the most important idea: combining summaries of two disjoint
-subsets of the data set gives us the same result as summarizing the
-union of the original subsets:
+Now comes the key concept: if the conditions above hold, then combining
+summaries of two disjoint subsets of the data set gives us the same
+result as summarizing the union of the original subsets:
 
@@
 \end{tikzcd}
-\] The top and bottom triangles in the diagram both indicate
-associativity: in the top triangle, we can either first take the union
-of \(A\) and \(B\) and then union that with \(C\), or immediately union
-\(A\) with \(B \cup C\) and the result will be identical. The same
-applies for the monoidal summary in the bottom triangle. By saying that
-the diagram commutes, we mean that any two parallel paths (the same
-start and endpoint) always produce the same result. Practically, this
-means we can either first summarize each set and then combine the
-summaries, or first union the sets and then summarize the union, or do
-some mix of the two, and we will always end up with a valid summary. As
-a sidenote, the diagram above actually corresponds to something called a
-functor or monoid homomorphism, however, that is beyond the scope of the
-present paper \citep[interested reader should
-see][]{fong2019, lawvere2009}.
-
-But this is precisely the idea of a stackable statistic. If our
-statistic conforms to the diagram, then the combined summary of subsets
-of the data is always a valid summary of the whole data set. In Wilke's
-words, we have discovered what makes the combined summary a ``meaningful
-amount''!
+\] The top and bottom triangles in the diagram both show associativity:
+in the top triangle, we can either first take the union of \(A\) and
+\(B\) and then union the result with \(C\), or immediately union \(A\)
+with \(B \cup C\), and the result will be identical. The same applies
+for the monoidal summary in the bottom triangle. By saying that the
+diagram commutes, we mean that any two parallel paths (the same start
+and endpoint) always produce the same result. Practically, this means we
+can either first summarize each set and then combine the summaries,
+first union the sets and then summarize the union, or do some mix of
+both, and we always end up with a valid summary. As a sidenote, the
+diagram above actually corresponds to something called a functor or
+monoid homomorphism, however, that is beyond the scope of the present
+paper \citep[interested reader should see][]{fong2019, lawvere2009}.
+
+But what we have now is precisely the idea of a stackable statistic. If
+our statistic conforms to the diagram, then the combined summary of
+subsets of the data is always a valid summary of the whole data set. In
+Wilke's words, we have discovered what makes the combined summary a
+``meaningful amount''!
 
@@
 i.e.~\(F(A)\), \(F(A \cup B)\), and \(F(A \cup B \cup C)\). This is
-fine, for example, when we do linked selection with a single group and
-compare the summary on the highlighted subset (\(A\)) vs.~the entirety
-the data (\(A \cup B\)).
+fine, for example, when we do linked selection with a single highlighted
+group and want to compare the summary on the highlighted subset (\(A\))
+vs.~the entirety the data (\(A \cup B\)). The resulting statistics give
+us what we need.
 
@@
 
-Importantly, \(F(A) \otimes F(B)\) is still a valid summary of
-\(A \cup B\) even when the inverse does not exist. It is just that
-\(F(A) \otimes F(B)\) may ``forget'' the information contained in either
-of its two parts. Note also that this is not a problem of different
-permutations of parts producing different sequences of accumulated
-summaries - this will happen regardless of whether or not there is an
-inverse.
+Importantly, even when the operation does not have an inverse,
+\(F(A) \otimes F(B)\) is still a valid summary of \(A \cup B\). It is
+just that \(F(A) \otimes F(B)\) may ``forget'' the information contained
+in either of its two constituent parts. Note also that this is not a
+problem of different permutations of parts producing different sequences
+of accumulated summaries - this will happen regardless of whether there
+is an inverse or not.
 
@@
 \caption{\label{fig:summax} Monoids can ``collapse'' information about
-disjoint parts. Top: since sum is a group with an inverse (minus), we
-can recover the value of each disjoint part by subtracting the heights
-of the preceding segments (i.e.~the part order does not matter). Bottom:
-since maximum lacks an inverse, we may lose information about the
-individual parts under certain permutations (note the absence of
-segments in the middle and rightmost plot).}
+disjoint subsets. Top: since sum is a group with an inverse (minus), we
+can recover the value of each part by subtracting the heights of the
+preceding segments (i.e.~the part order does not matter). Bottom: since
+maximum lacks an inverse, we may lose information about the individual
+parts under certain permutations (note the absence of segments in the
+middle and rightmost plot).}
 \end{figure}
@@
 To summarize, whether our statistic is a monoid or a group determines
-what kinds of comparisons we can make on stacked objects:
+what kinds of comparisons we can make with stacked objects:
 
@@
 This becomes important when implementing interactive data visualization
-features such as linked selection. Specifically, if we want to compare a
-single selected group vs.~the rest of the data, then it's enough for our
-summary statistic to be a monoid. If we want to implement multiple
-selection groups, then our statistic needs to have an inverse as well.
-If we fail either of these conditions, then we may produce data
-representations which are either inaccurate (in the case of a
-non-monoidal statistic), or accurate but not ``full'' (in the case of a
-monoid but no inverse).
+features such as linked selection. Specifically, if we want to represent
+a single highlighted group vs.~the rest of the data via stacked objects,
+then it's enough for our summary statistic to be a monoid. If we want to
+represent multiple highlighted groups, then our statistic needs to have
+an inverse as well. If we fail either of these conditions, we may
+produce representations which are either inaccurate (in the case of
+non-monoidal statistics), or accurate but not ``full'' (in the case of
+monoids but no inverse).
 
@@
 summaries and figures will ``work''. For example, while not generally
-thought of as a summary statistic, the convex hull algorithm \citep[see
-e.g.][]{barber1996} provides a way to reduce a set of points to a
-simpler ``summary'', see Figure \ref{fig:convexhull}. We can recast the
-operation of finding the convex hull as a monoid: starting with an empty
-set, we iterate through the data, adding points if they lie outside of
-the current hull (and removing any points in the interior of the new
-hull). This may not be the most computationally efficient method for
-finding convex hulls, however, since it does meet the definition of a
-monoid, this tells us \emph{what convex hulls can be used for}. If we
-combine two convex hulls, we know that the new hull is a valid summary
-for the union of the underlying data points. In turn, this allows us to
+thought of as a summary statistic, the convex hull \citep[see
+e.g.][]{barber1996} provides a way to reduce a set of data points into a
+simpler ``summary'', see Figure \ref{fig:convexhull}. We can recast
+convex hull algorithm as a monoid: starting with an empty set, we
+iterate through the data, adding points if they lie outside of the
+current hull (and removing any points in the interior of the new hull).
+This may not be the most computationally efficient method of finding
+convex hulls, however, since it does meet the definition of a monoid,
+this tells us \emph{what convex hulls can be used for}. If we combine
+two convex hulls, we know that the new hull will be a valid summary for
+the union of the underlying data points. In turn, this allows us to
 implement interactive features such as single-group
@@
 convex hulls may irrevocably collapse information. As such, we cannot
-use convex hulls to display multiple selections, without resorting to
-some ad-hoc methods (e.g.~keeping separate representation of the hulls
-and using partial transparency).
+safely use convex hulls to display multiple selection groups, without
+resorting to some ad-hoc methods.
 
@@
 representation of single-group selection but not multi-group selection.
-Both plots show data from Palmer penguins data set (Horst 2020). The
-light polygon shows the convex hull of the whole dataset, the dark
-polygon shows the hull of the Gentoo species, the medium-brightness
-polygon shows the combined hull of Gentoo and Chinstrap species, and the
-white dashed line shows the outline the hull of the Chinstrap penguins
-alone. Points show individual cases. Left: since convex hull is a
-monoid, we can use it to show single group selection/highlighting.
-Right: because of the lack of an inverse, we cannot guarantee that hull
-of the Chinstrap penguins (white-dashed line) will be recoverable from
-the combined hull of Gentoo and Chinstrap - we do not know how many
-interior points were ``forgotten''.}
+The light polygons shows the convex hull of the whole dataset, the dark
+polygon shows the hull of one selection group, the white dashed line
+shows the hull of a second selection group, and the medium-brightness
+polygon shows the combined hull of two selection groups. Points show
+individual cases. Left: since convex hull is a monoid, we can use it to
+show single highlighted group. Right: because of the lack of an inverse,
+we cannot guarantee that we will be able to recover either of the
+original hulls from the combined hull - we do not know how many interior
+points were ``forgotten''. The source of the data is the Palmer penguins
+data set, Horst 2020.}
 \end{figure}

### 2024-10-23 paper.Rmd
Wrote some more

@@
 This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids.
-Specifically, if we want our graphics to support interactive features which split our data into parts and then combine these parts back together (such as linked selection), then we need the statistics in our plots to fulfill certain algebraic properties.
-Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.
+Specifically, if we want our graphics to support interactive features which split our data into parts and then combine these parts back together (such as linked selection), then the statistics in our plots need to posses certain algebraic properties.
+Ultimately, by grounding our thinking in these concepts, we may be able to build more flexible and expressive interactive data visualization systems.
 bibliography: references.bib
 biblio-style: apalike
-nocite: |
-  @magritte1929
-  @horst2020
 output:
@@
 
+\nocite{horst2020}
+\nocite{magritte1929}
+
 ```{r}
@@
 #| echo: false
-#| fig-align: "center"
-#| fig-cap: "Examples of three methods for dealing with bars that have split into child components (segments).
-Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822).
-Bottom row: layering (created with ggplot2; Wickham, 2010)."
+#| fig-cap: "\\label{fig:playfair}Examples of three methods for dealing with bars that have split into child components (segments). 
+#| Top row: dodging (left) and stacking/highlighting (right), both by William Playfair [-@playfair1801; -@playfair1822]. 
+#| Bottom row: layering, created with `ggplot2` [@wickham2010; the source of the underlying data is the U.S. fuel economy data set, @mpg2024]."
 #| message: false
@@
 
-mpg$manufacturer <- tools::toTitleCase(mpg$manufacturer)
+mpg$manufacturer <- factor(mpg$manufacturer)
+levels(mpg$manufacturer) <- LETTERS[1:length(levels(mpg$manufacturer))]
 
@@
   geom_bar(position = "identity", alpha = 0.5) +
-  scale_x_discrete(labels = NULL) +
   guides(fill = "none") +
@@
 #| in the rightmost plot (the darker segment has been stretched to fill the available space; 
-#| `ggplot2` default for handling empty subcategories). Data source: @henderson1981."
+#| `ggplot2` default for handling empty subcategories). 
+#| The source of the underlying data is the `mtcars` dataset [@henderson1981]."
 
@@
 First, we propose that stacking also provides a simpler mental model for *selection*.
-Since each stacked bar is a single visual entity, it makes intuitive sense that clicking it will select the entire category corresponding to it.
-Indeed, this is how many interactive data visualization systems implement linked selection: clicking on part of an object selects the category corresponding to the entire object [see e.g.
+Since each stacked bar is a single visual entity, it makes intuitive sense that clicking it will select the entire corresponding category.
+Indeed, this is how many interactive data visualization systems implement default linked selection: clicking on part of an object selects all cases corresponding to the whole object [see e.g.
 @cook2023; @vegalite2022; @vanderplas2018; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @swayne2003].
-On the other hand, with dodged bars, the expectation is less clear: when we click on a segment in a dodged barplot, do we select the entire category or just the cases corresponding to the segment?
-Second, in a stacked barplot, the bar width and inter-bar gap are constant throughout selection.
-In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty - should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink them in response to selection?
-If we do the former, the barplot may show asymmetric gaps and illusory clustering due to the absence of some segments.
-If we do the latter, we break the principle that the area of the bar should be proportional to the number of cases it represents [@cleveland1985].
+On the other hand, with dodged bars, since the segments do not form a unified whole, the expectation is less clear: when we click on a segment in a dodged barplot, do we select the entire category or just the cases corresponding to the individual segment?
+Second, stacked barplots maintain constant bar widths and inter-bar gaps throughout selection.
+In a dodged barplot, it is not clear what to do when some selection groups are empty.
+Should we leave gaps for the (absent) highlighted segments, leading to asymmetric gaps and potential illusory clustering?
+Or should we draw wide segments and only shrink them in response to selection, avoiding asymmetry but violating the principle of area proportionality [@cleveland1985]?
 Finally, by presenting fewer visual entities on the screen, stacked bars may also reduce cognitive load [@sweller2019; for a non-technical overview, see @knaflic2015].
 
-To summarize, stacked bars exhibit a more graceful and visually pleasing behavior when combined with selection.
-In contrast, dodging produces non-uniform objects with jagged outline that changes significantly with selection.
-These difference may explain why stacking is still popular in the interactive data visualization community.
+To summarize, stacked bars behave in a fairly graceful and visually pleasing way when combined with selection.
+Dodged bars, on the other hand, lack a bounded uniform outline, and this makes them behave erratically when combined with selection, hurting the interpretability of the plot.
+These difference may explain the continued popularity of stacking within the interactive data visualization community.
 
@@
 #| fig-align: "center"
-#| fig-cap: "A representation is not the thing itself.
-Left: The Treachery of Images by René Magritte (1929) makes the point that a painting of a pipe is not the thing it represents (a pipe).
-Right: similarly, an filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."
+#| fig-cap: "The representation of a thing is not the thing itself.
+Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe.
+Right: similarly, an image filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."
 
@@
 
-The connection between summary statistics and algebraic structures is important because it allows us to predict which new combinations of summaries and figures will "work".
+We can use the connection between summary statistics and algebraic structures to offload some of the cognitive effort required when thinking about new types of interactive visualizations.
 For example, while not generally thought of as a summary statistic, the convex hull [see e.g.
@@
 
+While it is possible to reach the same conclusions via ad hoc reasoning or manual testing, these methods may be time-consuming and have to be applied to each new visualization style separately.
+By grounding our understanding in the simple theory of groups and monoids, we can bypass most of this effort and reason about the properties of the visualization directly.
+
 ```{r}
@@
 #| fig-cap: "\\label{fig:convexhull}Convex hull can be used as a valid representation of single-group selection but not multi-group selection.
-The light polygons shows the convex hull of the whole dataset, the dark polygon shows the hull of one selection group, the white dashed line shows the hull of a second selection group, and the medium-brightness polygon shows the combined hull of two selection groups.
+The light polygons shows the convex hull of the whole dataset, the dark polygon shows the hull of one selection group, the white dashed line shows the hull of a second selection group, and the medium-brightness polygon shows the combined hull of the two selection groups.
 Points show individual cases.
-Left: since convex hull is a monoid, we can use it to show single highlighted group.
-Right: because of the lack of an inverse, we cannot guarantee that we will be able to recover either of the original hulls from the combined hull - we do not know how many interior points were \"forgotten\".
-The source of the data is the Palmer penguins data set, Horst 2020."
+Left: since convex hull is a monoid, we can use it to show a single highlighted group.
+Right: because convex hull lacks an inverse, after combining two hulls, we cannot guarantee we will be able to recover either of the original hulls - we do not know how many interior points were \"forgotten\".
+The source of the underlying data is the Palmer penguins data set [@horst2020]."
 #| message: false
@@
 
-There are also other advantages that groups and monoids offer for visualizing data.
+There are other benefits that groups and monoids offer for visualizing data.
 One of these is the existence of the neutral element $e$ (see Section \ref{sec:monoids}).
@@
 A common approach is to omit the objects representing empty subsets.
-However, with non-monoidal summaries, this approach is flawed and leads ambiguity.
+However, with non-monoidal summaries, this approach is flawed and leads to ambiguity.
 Suppose we are drawing the bars of averages.
-If we omit bars representing the empty categories, then the missingness of a bar can indicate either of two things: either there are *no* cases in the category, or there *are* cases and their average is equal to the lower y-axis limit.
+If we omit bars representing empty categories, then the missingness of a bar can indicate one of two things: either there are *no* cases in the category, or there *are* cases and their average is equal to the lower y-axis limit.
 There is just no way to tell from the graphic alone.
-In contrast, with monoidal summaries such as sums, the message of an object's absence is unambiguous.
-Regardless of whether there are zero or hundred cases in a category, if the bar representing them is missing, this means that their sum is zero.
-
-The neutral element can also provide a new perspective on a perennial debate in data visualization: should the base of a barplot always start at zero?
-Some authors argue that starting bars from a non-zero base makes the lengths of the bars meaningless [@cleveland1985].
-However, @wilkinson2012, for example, suggests that the bar base is really a statement about scales rather than about graphics, and as such there is no compulsion for starting it at zero.
-We offer the following new perspective: often, it might make sense for the base of the barplot to start at $e$, the neutral element, since this represents the default state of doing nothing.
-In other words, the bar base is tied to the properties of the *statistic*, rather than the just the graphic or scale.
-
-In a barplot of sums or counts, this approach aligns with the traditional prescription, since the neutra element is zero.
-However, in a barplot of products, the neutral element is one.
-Thus, we may should our bars at one.
-One might object that this invalidates the y-axis as a ratio scale [@stevens1946], since a bar representing the value 4 will be no longer twice as tall as the bar representing the value 2.
-And indeed, if we start the base at one, the bar representing the value 2 will be only 2 - 1 = 1 units tall, whereas the bar representing the value 4 will be 4 - 1 = 3 units tall.
-However, we need to think about what the values represent.
-The bar representing the value 2 shows a 200% change or 100% increase; the bar representing 4 shows 400% change or 300% increase.
-Therefore, it makes sense that the bar representing 4 is three times taller.
-Also, this scale still fits the @stevens1946 definition of a ratio scale, since we can still determine equality of ratios: 1.5 is to 2 (100% / 50% = 2, double the increase) as 2 is to 3 (200% / 100% = 2, double the increase).
-
-This example provides further evidence to support the claim that we cannot separate statistics from other data visualization components such as graphics and scales.
+In contrast, with monoidal summaries such as sums, the absence of an object (bar) sends a clear message.
+Regardless of whether there are zero or hundred cases in a category, if the corresponding bar is missing, then we know their sum is zero.
+
+The neutral element can also provide a new perspective on the longstanding debate in data visualization: should the base of a barplot always start at zero?
+Some authors argue that starting from a non-zero base renders bar lengths meaningless [@cleveland1985].
+On the other hand, @wilkinson2012, for example, suggests that the bar base is really a statement about scales rather than graphics, and as such it does not have to be zero.
+We offer the following new perspective: often, it might make sense to start the base of the barplot at $e$, the neutral element, since this represents the default state of "doing nothing".
+In other words, the bar base is tied to the properties of the *statistic*, rather than just the graphic or the scale.
+For instance, in a logarithmic barplot, the underlying monoidal operation is multiplication, and so the base of the bars should start at one.
+
+These example provides further support for the claim that we cannot separate statistics from other components such as graphics and scales.
 Even before we apply any transformations, the method we have used to derive our summary statistics determines important properties of our visualization.
@@
 
-Groups and monoids also come with some inherent computational advantages.
+Groups and monoids also offer some inherent computational advantages.
 Specifically, if a summary statistic is a monoid, associativity guarantees that it can be computed in a single pass through the data (i.e.
@@
 @lammel2008; @lin2013].
-This is why monoids are popular in functional programming, where they are used alongside the so-called `fold` or `reduce` functions [see e.g.
+This is makes monoids popular in functional programming, where they are known as the `fold` or `reduce` functions [see e.g.
 @milewski2018; @abelson2022].
-To be fair, computational efficiency is usually not the primary concern in data visualization: most data visualization systems are designed to work in a single-threaded environment, and rely on summary statistics that are fairly cheap to compute.
-Nevertheless, the guarantee of inherent efficiency can still be useful when visualizing larger data sets.
-
-The computational advantages of groups and monoids become significant in interactive data visualization, however.
-Specifically, when designing an interactive data visualization, we may be able to exploit the fact that summaries on fine-grained subsets of the data "add up" to the summaries on coarse-grained subsets (their unions) to avoid unnecessary computation.
-Consider a linked histogram with reactive axes and interactive manipulation of binwidth and anchor.
-When the user engages in selection, the counts within the individual highlighted segments may change but the counts across the whole stacked bars will not.
-Thus, there is no need to recompute and update the upper y-axis limit - we know this will stay constant throughout selection.
-However, if the user interactively changes the binwidth or anchor, then both the counts within the whole bars *and* the highlighted segments may change and so we need to update the graphic appropriately.
+
+While computational efficiency is usually not the number one concern in static visualization, it becomes significant in the context of interactive graphics.
+Here, we may be able to use the properties of groups and monoids to avoid unnecessary computation.
+Specifically, when using groups and monoids, we can exploit associativity: the fact that summaries on fine-grained subsets of the data "add up" to the summaries on coarse-grained subsets.
+
+Consider a stacked histogram, with linked selection, reactive axes, and interactive manipulation of binwidth and anchor.
+When the user engages in selection, the counts within the individual highlighted segments may change but the counts within the whole bars will stay constant.
+Therefore, there is no need to update the upper y-axis limit.
+However, if the user updates either the binwidth or the anchor, then both the counts within the whole bars *and* within the highlighted segments may change, and so we need to update the graphic appropriately.
 
 In other words, associativity allows us to implement strategic updating.
-Fine-grained summaries (individual segments) need to be updated when the more coarse-grained summaries (whole stacked bars) change, but not vice versa.
+Fine-grained summaries (individual segments) need to be updated whenever coarse-grained summaries (whole stacked bars) change, but not vice versa.
 By pre-computing summaries on different levels of aggregation, we may be able to reduce computational effort.
@@
 
-However, while the inverse does bring advantages, it is also the case that fewer summary statistics have it.
+However, while the presence of the inverse does bring some advantages, it is also the case that fewer summary statistics have an inverse.
 For instance, as was mentioned before, while maximum operator is a valid monoid, it lacks an inverse and as such does not meet the definition of a group.
@@
 And there are many other operations (such as the convex hull in Figure \ref{fig:convexhull}) which lack an inverse.
-As such, when designing and interactive data visualization library, we need to carefully weigh the ability to support different kinds of summary statistics (general, monoids, groups) versus supporting different kinds of interactions (e.g.
+As such, when designing and interactive data visualization library, we need to carefully weigh the ability to support different kinds of summary statistics (general, monoids, groups) against the ability to support wide range of interactions (e.g.
 single-group selection vs.
@@
 
-# Conclusion
+# Discussion
 
 We hope we have demonstrated that, when designing interactive graphics, algebraic thinking is invaluable.
-To support specific interactive features, the statistics underlying our plots must possess appropriate algebraic properties.
-For instance, if we want to display the result of splitting our data into parts and combining these parts back together, then we need either of two classes of mathematical structures.
-If we want to compare nested subsets of our data (for example, via single-group linked selection) then our summaries need to be monoids.
-If we want to compare disjoint parts of our data (multi-group linked selection), then our summaries need to be (algebraic) groups.
-The properties of these structures ensure that our figures are statistically sound, computationally efficient, and adhere to good design principles.
-
-We have focused on unitality, associativity, and inverses, however, it may be useful to consider other algebraic properties as well.
-For instance, many of the visual attributes that we use to represent our data can only *grow* in one direction.
-Such is the case with, for example, width, height, and area (there is no such thing as "negative area").
-If the visual attribute can only grow in one direction, then our summary statistic should be *monotonic*, such that if $x_1 \leq y_1$ and and $x_2 \leq y_2$ then $x_1 \otimes x_2 \leq y_1 \otimes y_2$.
+To support certain interactive features such as linked selection, the statistics underlying our plots must possess appropriate algebraic properties.
+If we want to display the result of splitting our data into parts and then combining these parts back together, then we need either monoids or groups.
+Monoids are for comparing nested subsets of our data (single-group linked selection) whereas groups are for comparing disjoint parts of our data (multi-group linked selection).
+The properties of these structures ensure that our figures will be statistically sound, computationally efficient, and adhere to good design principles.
+
+We do not advocate for exclusive use of groups and monoids.
+Rather, we want to highlight the inherent trade-off present in the choice of a summary statistic.
+Few summary statistics meet the definition of a group, but, if we choose such a statistic, then we can do a lot of things with it.
+With a monoid, our options are more limited, and with a non-monoidal statistic, even more so.
+Like many past authors [such as @tufte2001; @tukey1977; @wilkinson2012], we want to emphasize the importance of thinking critically about the mathematical structure underlying our graphics.
+Geometric objects, statistics, and interaction are all deeply connected, and groups and monoids merely allow us to carve out a small space of fairly well-behaved visualizations.
+
+On that note, while we have focused on unitality, associativity, and inverses, it may be useful to consider other algebraic properties as well.
+For instance, many of the visual attributes that we use to represent our data, such as width, height, or area, can only *grow* in one direction (there is no such thing as "negative area").
+If this is the case, then our summary statistic should be *monotonic*, such that if $x_1 \leq y_1$ and and $x_2 \leq y_2$ then $x_1 \otimes x_2 \leq y_1 \otimes y_2$.
 Similarly, in many types of plots, the order of the data points does not matter, however, in some, such as time-series plots, it does.
@@
 
-We do not advocate for exclusively visualizing groups and monoids.
-Rather, we want to highlight the trade-off present in the choice of the summary statistic.
-Few summary statistics meet the definition of a group, but, if we choose such a statistic, then we can do a lot of things with it.
-With a monoid, our options are more limited, and with a non-monoidal statistic, even more so.
-Thus, like many past authors [such as @tufte2001; @tukey1977; @wilkinson2012], we simply want to point to the importance of thinking critically about the mathematical structure underlying our graphics.
-Geometric objects, statistics, and interaction are all deeply connected, and a full elaboration on this web of interdependence may require a lot more work.
-Groups and monoids merely allow us to carve out a small space of well-behaved graphics.
-
-On a more practical note, we also need a framework for efficiently transforming raw data into summary statistics that can be displayed by interactive graphics.
-A decade ago, @wickham2009 noted the lack of such a formal pipeline (or "plumbing"), and, despite some attempts at implementing such pipelines [see e.g.
-@lawrence2009], that same sentiment was echoed again recently [@vanderplas2020].
-Ideally, such a framework would minimize recomputation during user interaction.
-The computational "shortcuts" that groups and monoids provide may be invaluable here.
-We envision a visualization pipeline structured as a hierarchy of reactive partitions, where only the downstream (finer) partitions need to be recomputed when the upstream (coarser) partitions change, but not vice versa.
-We are currently exploring this concept in an R package called `plotscaper`^[`https://github.com/bartonicek/plotscaper`] (the project still has far to go to be truly generic).
+On a more practical note, interactive graphics require efficient pipelines for transforming raw data into summary statistics, and, currently, we still do not possess a formal framework for doing this.
+A decade ago, @wickham2009 noted the lack of such data pipelines, and, despite some attempts at implementation since then [see e.g.
+@lawrence2009], the same sentiment was echoed again recently [@vanderplas2020].
+The ultimate goal is to minimize recomputation during user interaction, and this is where the computational shortcuts that groups and monoids provide may prove invaluable.
+We envision a visualization pipeline structured as a hierarchy of reactive partitions, where only the downstream (fine) partitions need to be recomputed when the upstream (coarse) partitions change, but not vice versa.
+We are currently exploring this idea in an R package called `plotscaper`^[`https://github.com/bartonicek/plotscaper`] (note that the project is still under development).
 
@@
 Nevertheless, we believe that simple algebraic thinking offers a valuable foundation.
-Concepts like monoids and groups provide a rigorous framework for reasoning about graphics, interactive or otherwise.
-Ultimately, these concepts may empower us to build a broader class of robust interactive data visualizations.
+Concepts like groups and monoids provide a solid framework for reasoning about graphics, interactive or otherwise.
+By grounding our thinking in these concepts, we may be able to offload some of the required cognitive effort and envision new types of robust and expressive interactive graphics.
 

### 2024-10-23 paper.tex
Wrote some more

@@
 support interactive features which split our data into parts and then
-combine these parts back together (such as linked selection), then we
-need the statistics in our plots to fulfill certain algebraic
-properties. Ultimately, understanding this hidden structure may empower
-data visualization experts to build more flexible and expressive
-interactive data visualization systems.
+combine these parts back together (such as linked selection), then the
+statistics in our plots need to posses certain algebraic properties.
+Ultimately, by grounding our thinking in these concepts, we may be able
+to build more flexible and expressive interactive data visualization
+systems.
 \end{abstract}
@@
 \spacingset{1.9} % DON'T change the spacing!
+
+\nocite{horst2020} \nocite{magritte1929}
 
@@
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/playfair-1}
-
-}
-
-\caption{Examples of three methods for dealing with bars that have split into child components (segments).
-Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822).
-Bottom row: layering (created with ggplot2; Wickham, 2010).}\label{fig:playfair}
+\centering
+\includegraphics{paper_files/figure-latex/playfair-1.pdf}
+\caption{\label{fig:playfair}Examples of three methods for dealing with
+bars that have split into child components (segments). Top row: dodging
+(left) and stacking/highlighting (right), both by William Playfair
+\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Bottom row:
+layering, created with \texttt{ggplot2}
+\citetext{\citealp{wickham2010}; \citealp[the source of the underlying
+data is the U.S. fuel economy data set,][]{mpg2024}}.}
 \end{figure}
@@
 has been stretched to fill the available space; \texttt{ggplot2} default
-for handling empty subcategories). Data source: \citet{henderson1981}.}
+for handling empty subcategories). The source of the underlying data is
+the \texttt{mtcars} dataset \citep{henderson1981}.}
 \end{figure}
@@
 bar is a single visual entity, it makes intuitive sense that clicking it
-will select the entire category corresponding to it. Indeed, this is how
-many interactive data visualization systems implement linked selection:
-clicking on part of an object selects the category corresponding to the
-entire object \citep[see
+will select the entire corresponding category. Indeed, this is how many
+interactive data visualization systems implement default linked
+selection: clicking on part of an object selects all cases corresponding
+to the whole object \citep[see
 e.g.][]{cook2023, vegalite2022, vanderplas2018, xie2014, theus2002, urbanek2003, urbanek2011, swayne2003}.
-On the other hand, with dodged bars, the expectation is less clear: when
-we click on a segment in a dodged barplot, do we select the entire
-category or just the cases corresponding to the segment? Second, in a
-stacked barplot, the bar width and inter-bar gap are constant throughout
-selection. In a dodged barplot, it is not clear how to handle the gaps
-when some selection groups are empty - should we leave gaps for the
-(currently absent) highlighted segments, or draw wide bars and only
-shrink them in response to selection? If we do the former, the barplot
-may show asymmetric gaps and illusory clustering due to the absence of
-some segments. If we do the latter, we break the principle that the area
-of the bar should be proportional to the number of cases it represents
-\citep{cleveland1985}. Finally, by presenting fewer visual entities on
+On the other hand, with dodged bars, since the segments do not form a
+unified whole, the expectation is less clear: when we click on a segment
+in a dodged barplot, do we select the entire category or just the cases
+corresponding to the individual segment? Second, stacked barplots
+maintain constant bar widths and inter-bar gaps throughout selection. In
+a dodged barplot, it is not clear what to do when some selection groups
+are empty. Should we leave gaps for the (absent) highlighted segments,
+leading to asymmetric gaps and potential illusory clustering? Or should
+we draw wide segments and only shrink them in response to selection,
+avoiding asymmetry but violating the principle of area proportionality
+\citep{cleveland1985}? Finally, by presenting fewer visual entities on
 the screen, stacked bars may also reduce cognitive load
@@
 
-To summarize, stacked bars exhibit a more graceful and visually pleasing
-behavior when combined with selection. In contrast, dodging produces
-non-uniform objects with jagged outline that changes significantly with
-selection. These difference may explain why stacking is still popular in
-the interactive data visualization community.
+To summarize, stacked bars behave in a fairly graceful and visually
+pleasing way when combined with selection. Dodged bars, on the other
+hand, lack a bounded uniform outline, and this makes them behave
+erratically when combined with selection, hurting the interpretability
+of the plot. These difference may explain the continued popularity of
+stacking within the interactive data visualization community.
 
@@
 
-\caption{A representation is not the thing itself.
-Left: The Treachery of Images by René Magritte (1929) makes the point that a painting of a pipe is not the thing it represents (a pipe).
-Right: similarly, an filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
+\caption{The representation of a thing is not the thing itself.
+Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe.
+Right: similarly, an image filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
 \end{figure}
@@
 
-The connection between summary statistics and algebraic structures is
-important because it allows us to predict which new combinations of
-summaries and figures will ``work''. For example, while not generally
-thought of as a summary statistic, the convex hull \citep[see
-e.g.][]{barber1996} provides a way to reduce a set of data points into a
-simpler ``summary'', see Figure \ref{fig:convexhull}. We can recast
-convex hull algorithm as a monoid: starting with an empty set, we
-iterate through the data, adding points if they lie outside of the
-current hull (and removing any points in the interior of the new hull).
-This may not be the most computationally efficient method of finding
-convex hulls, however, since it does meet the definition of a monoid,
-this tells us \emph{what convex hulls can be used for}. If we combine
-two convex hulls, we know that the new hull will be a valid summary for
-the union of the underlying data points. In turn, this allows us to
-implement interactive features such as single-group
+We can use the connection between summary statistics and algebraic
+structures to offload some of the cognitive effort required when
+thinking about new types of interactive visualizations. For example,
+while not generally thought of as a summary statistic, the convex hull
+\citep[see e.g.][]{barber1996} provides a way to reduce a set of data
+points into a simpler ``summary'', see Figure \ref{fig:convexhull}. We
+can recast convex hull algorithm as a monoid: starting with an empty
+set, we iterate through the data, adding points if they lie outside of
+the current hull (and removing any points in the interior of the new
+hull). This may not be the most computationally efficient method of
+finding convex hulls, however, since it does meet the definition of a
+monoid, this tells us \emph{what convex hulls can be used for}. If we
+combine two convex hulls, we know that the new hull will be a valid
+summary for the union of the underlying data points. In turn, this
+allows us to implement interactive features such as single-group
 highlighting/selection, safe in the knowledge that the underlying
@@
 resorting to some ad-hoc methods.
+
+While it is possible to reach the same conclusions via ad hoc reasoning
+or manual testing, these methods may be time-consuming and have to be
+applied to each new visualization style separately. By grounding our
+understanding in the simple theory of groups and monoids, we can bypass
+most of this effort and reason about the properties of the visualization
+directly.
 
@@
 shows the hull of a second selection group, and the medium-brightness
-polygon shows the combined hull of two selection groups. Points show
+polygon shows the combined hull of the two selection groups. Points show
 individual cases. Left: since convex hull is a monoid, we can use it to
-show single highlighted group. Right: because of the lack of an inverse,
-we cannot guarantee that we will be able to recover either of the
-original hulls from the combined hull - we do not know how many interior
-points were ``forgotten''. The source of the data is the Palmer penguins
-data set, Horst 2020.}
+show a single highlighted group. Right: because convex hull lacks an
+inverse, after combining two hulls, we cannot guarantee we will be able
+to recover either of the original hulls - we do not know how many
+interior points were ``forgotten''. The source of the underlying data is
+the Palmer penguins data set \citep{horst2020}.}
 \end{figure}
@@
 
-There are also other advantages that groups and monoids offer for
-visualizing data. One of these is the existence of the neutral element
-\(e\) (see Section \ref{sec:monoids}). While seemingly inconspicuous,
-the neutral element can actually be quite useful.
+There are other benefits that groups and monoids offer for visualizing
+data. One of these is the existence of the neutral element \(e\) (see
+Section \ref{sec:monoids}). While seemingly inconspicuous, the neutral
+element can actually be quite useful.
 
@@
 However, with non-monoidal summaries, this approach is flawed and leads
-ambiguity. Suppose we are drawing the bars of averages. If we omit bars
-representing the empty categories, then the missingness of a bar can
-indicate either of two things: either there are \emph{no} cases in the
+to ambiguity. Suppose we are drawing the bars of averages. If we omit
+bars representing empty categories, then the missingness of a bar can
+indicate one of two things: either there are \emph{no} cases in the
 category, or there \emph{are} cases and their average is equal to the
 lower y-axis limit. There is just no way to tell from the graphic alone.
-In contrast, with monoidal summaries such as sums, the message of an
-object's absence is unambiguous. Regardless of whether there are zero or
-hundred cases in a category, if the bar representing them is missing,
-this means that their sum is zero.
-
-The neutral element can also provide a new perspective on a perennial
-debate in data visualization: should the base of a barplot always start
-at zero? Some authors argue that starting bars from a non-zero base
-makes the lengths of the bars meaningless \citep{cleveland1985}.
-However, \citet{wilkinson2012}, for example, suggests that the bar base
-is really a statement about scales rather than about graphics, and as
-such there is no compulsion for starting it at zero. We offer the
-following new perspective: often, it might make sense for the base of
-the barplot to start at \(e\), the neutral element, since this
-represents the default state of doing nothing. In other words, the bar
-base is tied to the properties of the \emph{statistic}, rather than the
-just the graphic or scale.
-
-In a barplot of sums or counts, this approach aligns with the
-traditional prescription, since the neutra element is zero. However, in
-a barplot of products, the neutral element is one. Thus, we may should
-our bars at one. One might object that this invalidates the y-axis as a
-ratio scale \citep{stevens1946}, since a bar representing the value 4
-will be no longer twice as tall as the bar representing the value 2. And
-indeed, if we start the base at one, the bar representing the value 2
-will be only 2 - 1 = 1 units tall, whereas the bar representing the
-value 4 will be 4 - 1 = 3 units tall. However, we need to think about
-what the values represent. The bar representing the value 2 shows a
-200\% change or 100\% increase; the bar representing 4 shows 400\%
-change or 300\% increase. Therefore, it makes sense that the bar
-representing 4 is three times taller. Also, this scale still fits the
-\citet{stevens1946} definition of a ratio scale, since we can still
-determine equality of ratios: 1.5 is to 2 (100\% / 50\% = 2, double the
-increase) as 2 is to 3 (200\% / 100\% = 2, double the increase).
-
-This example provides further evidence to support the claim that we
-cannot separate statistics from other data visualization components such
-as graphics and scales. Even before we apply any transformations, the
-method we have used to derive our summary statistics determines
-important properties of our visualization. We need to be mindful of
-this.
+In contrast, with monoidal summaries such as sums, the absence of an
+object (bar) sends a clear message. Regardless of whether there are zero
+or hundred cases in a category, if the corresponding bar is missing,
+then we know their sum is zero.
+
+The neutral element can also provide a new perspective on the
+longstanding debate in data visualization: should the base of a barplot
+always start at zero? Some authors argue that starting from a non-zero
+base renders bar lengths meaningless \citep{cleveland1985}. On the other
+hand, \citet{wilkinson2012}, for example, suggests that the bar base is
+really a statement about scales rather than graphics, and as such it
+does not have to be zero. We offer the following new perspective: often,
+it might make sense to start the base of the barplot at \(e\), the
+neutral element, since this represents the default state of ``doing
+nothing''. In other words, the bar base is tied to the properties of the
+\emph{statistic}, rather than just the graphic or the scale. For
+instance, in a logarithmic barplot, the underlying monoidal operation is
+multiplication, and so the base of the bars should start at one.
+
+These example provides further support for the claim that we cannot
+separate statistics from other components such as graphics and scales.
+Even before we apply any transformations, the method we have used to
+derive our summary statistics determines important properties of our
+visualization. We need to be mindful of this.
 
@@
 
-Groups and monoids also come with some inherent computational
-advantages. Specifically, if a summary statistic is a monoid,
-associativity guarantees that it can be computed in a single pass
-through the data (i.e.~\(O(n)\) complexity), and in a
-distributed/parallel fashion \citep[see e.g.][]{lammel2008, lin2013}.
-This is why monoids are popular in functional programming, where they
-are used alongside the so-called \texttt{fold} or \texttt{reduce}
-functions \citep[see e.g.][]{milewski2018, abelson2022}. To be fair,
-computational efficiency is usually not the primary concern in data
-visualization: most data visualization systems are designed to work in a
-single-threaded environment, and rely on summary statistics that are
-fairly cheap to compute. Nevertheless, the guarantee of inherent
-efficiency can still be useful when visualizing larger data sets.
-
-The computational advantages of groups and monoids become significant in
-interactive data visualization, however. Specifically, when designing an
-interactive data visualization, we may be able to exploit the fact that
+Groups and monoids also offer some inherent computational advantages.
+Specifically, if a summary statistic is a monoid, associativity
+guarantees that it can be computed in a single pass through the data
+(i.e.~\(O(n)\) complexity), and in a distributed/parallel fashion
+\citep[see e.g.][]{lammel2008, lin2013}. This is makes monoids popular
+in functional programming, where they are known as the \texttt{fold} or
+\texttt{reduce} functions \citep[see e.g.][]{milewski2018, abelson2022}.
+
+While computational efficiency is usually not the number one concern in
+static visualization, it becomes significant in the context of
+interactive graphics. Here, we may be able to use the properties of
+groups and monoids to avoid unnecessary computation. Specifically, when
+using groups and monoids, we can exploit associativity: the fact that
 summaries on fine-grained subsets of the data ``add up'' to the
-summaries on coarse-grained subsets (their unions) to avoid unnecessary
-computation. Consider a linked histogram with reactive axes and
+summaries on coarse-grained subsets.
+
+Consider a stacked histogram, with linked selection, reactive axes, and
 interactive manipulation of binwidth and anchor. When the user engages
 in selection, the counts within the individual highlighted segments may
-change but the counts across the whole stacked bars will not. Thus,
-there is no need to recompute and update the upper y-axis limit - we
-know this will stay constant throughout selection. However, if the user
-interactively changes the binwidth or anchor, then both the counts
-within the whole bars \emph{and} the highlighted segments may change and
-so we need to update the graphic appropriately.
+change but the counts within the whole bars will stay constant.
+Therefore, there is no need to update the upper y-axis limit. However,
+if the user updates either the binwidth or the anchor, then both the
+counts within the whole bars \emph{and} within the highlighted segments
+may change, and so we need to update the graphic appropriately.
 
 In other words, associativity allows us to implement strategic updating.
-Fine-grained summaries (individual segments) need to be updated when the
-more coarse-grained summaries (whole stacked bars) change, but not vice
+Fine-grained summaries (individual segments) need to be updated whenever
+coarse-grained summaries (whole stacked bars) change, but not vice
 versa. By pre-computing summaries on different levels of aggregation, we
@@
 
-However, while the inverse does bring advantages, it is also the case
-that fewer summary statistics have it. For instance, as was mentioned
-before, while maximum operator is a valid monoid, it lacks an inverse
-and as such does not meet the definition of a group. Once we summarize
-some data with the maximum, we cannot retrieve the second or third
-highest value - that information is gone forever. And there are many
-other operations (such as the convex hull in Figure
-\ref{fig:convexhull}) which lack an inverse. As such, when designing and
-interactive data visualization library, we need to carefully weigh the
-ability to support different kinds of summary statistics (general,
-monoids, groups) versus supporting different kinds of interactions
-(e.g.~single-group selection vs.~multi-group selection).
-
-\section{Conclusion}\label{conclusion}
+However, while the presence of the inverse does bring some advantages,
+it is also the case that fewer summary statistics have an inverse. For
+instance, as was mentioned before, while maximum operator is a valid
+monoid, it lacks an inverse and as such does not meet the definition of
+a group. Once we summarize some data with the maximum, we cannot
+retrieve the second or third highest value - that information is gone
+forever. And there are many other operations (such as the convex hull in
+Figure \ref{fig:convexhull}) which lack an inverse. As such, when
+designing and interactive data visualization library, we need to
+carefully weigh the ability to support different kinds of summary
+statistics (general, monoids, groups) against the ability to support
+wide range of interactions (e.g.~single-group selection vs.~multi-group
+selection).
+
+\section{Discussion}\label{discussion}
 
 We hope we have demonstrated that, when designing interactive graphics,
-algebraic thinking is invaluable. To support specific interactive
-features, the statistics underlying our plots must possess appropriate
-algebraic properties. For instance, if we want to display the result of
-splitting our data into parts and combining these parts back together,
-then we need either of two classes of mathematical structures. If we
-want to compare nested subsets of our data (for example, via
-single-group linked selection) then our summaries need to be monoids. If
-we want to compare disjoint parts of our data (multi-group linked
-selection), then our summaries need to be (algebraic) groups. The
-properties of these structures ensure that our figures are statistically
-sound, computationally efficient, and adhere to good design principles.
-
-We have focused on unitality, associativity, and inverses, however, it
-may be useful to consider other algebraic properties as well. For
-instance, many of the visual attributes that we use to represent our
-data can only \emph{grow} in one direction. Such is the case with, for
-example, width, height, and area (there is no such thing as ``negative
-area''). If the visual attribute can only grow in one direction, then
-our summary statistic should be \emph{monotonic}, such that if
-\(x_1 \leq y_1\) and and \(x_2 \leq y_2\) then
+algebraic thinking is invaluable. To support certain interactive
+features such as linked selection, the statistics underlying our plots
+must possess appropriate algebraic properties. If we want to display the
+result of splitting our data into parts and then combining these parts
+back together, then we need either monoids or groups. Monoids are for
+comparing nested subsets of our data (single-group linked selection)
+whereas groups are for comparing disjoint parts of our data (multi-group
+linked selection). The properties of these structures ensure that our
+figures will be statistically sound, computationally efficient, and
+adhere to good design principles.
+
+We do not advocate for exclusive use of groups and monoids. Rather, we
+want to highlight the inherent trade-off present in the choice of a
+summary statistic. Few summary statistics meet the definition of a
+group, but, if we choose such a statistic, then we can do a lot of
+things with it. With a monoid, our options are more limited, and with a
+non-monoidal statistic, even more so. Like many past authors \citep[such
+as][]{tufte2001, tukey1977, wilkinson2012}, we want to emphasize the
+importance of thinking critically about the mathematical structure
+underlying our graphics. Geometric objects, statistics, and interaction
+are all deeply connected, and groups and monoids merely allow us to
+carve out a small space of fairly well-behaved visualizations.
+
+On that note, while we have focused on unitality, associativity, and
+inverses, it may be useful to consider other algebraic properties as
+well. For instance, many of the visual attributes that we use to
+represent our data, such as width, height, or area, can only \emph{grow}
+in one direction (there is no such thing as ``negative area''). If this
+is the case, then our summary statistic should be \emph{monotonic}, such
+that if \(x_1 \leq y_1\) and and \(x_2 \leq y_2\) then
 \(x_1 \otimes x_2 \leq y_1 \otimes y_2\). Similarly, in many types of
@@
 
-We do not advocate for exclusively visualizing groups and monoids.
-Rather, we want to highlight the trade-off present in the choice of the
-summary statistic. Few summary statistics meet the definition of a
-group, but, if we choose such a statistic, then we can do a lot of
-things with it. With a monoid, our options are more limited, and with a
-non-monoidal statistic, even more so. Thus, like many past authors
-\citep[such as][]{tufte2001, tukey1977, wilkinson2012}, we simply want
-to point to the importance of thinking critically about the mathematical
-structure underlying our graphics. Geometric objects, statistics, and
-interaction are all deeply connected, and a full elaboration on this web
-of interdependence may require a lot more work. Groups and monoids
-merely allow us to carve out a small space of well-behaved graphics.
-
-On a more practical note, we also need a framework for efficiently
-transforming raw data into summary statistics that can be displayed by
-interactive graphics. A decade ago, \citet{wickham2009} noted the lack
-of such a formal pipeline (or ``plumbing''), and, despite some attempts
-at implementing such pipelines \citep[see e.g.][]{lawrence2009}, that
-same sentiment was echoed again recently \citep{vanderplas2020}.
-Ideally, such a framework would minimize recomputation during user
-interaction. The computational ``shortcuts'' that groups and monoids
-provide may be invaluable here. We envision a visualization pipeline
-structured as a hierarchy of reactive partitions, where only the
-downstream (finer) partitions need to be recomputed when the upstream
-(coarser) partitions change, but not vice versa. We are currently
-exploring this concept in an R package called
+On a more practical note, interactive graphics require efficient
+pipelines for transforming raw data into summary statistics, and,
+currently, we still do not possess a formal framework for doing this. A
+decade ago, \citet{wickham2009} noted the lack of such data pipelines,
+and, despite some attempts at implementation since then \citep[see
+e.g.][]{lawrence2009}, the same sentiment was echoed again recently
+\citep{vanderplas2020}. The ultimate goal is to minimize recomputation
+during user interaction, and this is where the computational shortcuts
+that groups and monoids provide may prove invaluable. We envision a
+visualization pipeline structured as a hierarchy of reactive partitions,
+where only the downstream (fine) partitions need to be recomputed when
+the upstream (coarse) partitions change, but not vice versa. We are
+currently exploring this idea in an R package called
 \texttt{plotscaper}\footnote{\texttt{https://github.com/bartonicek/plotscaper}}
-(the project still has far to go to be truly generic).
+(note that the project is still under development).
 
@@
 believe that simple algebraic thinking offers a valuable foundation.
-Concepts like monoids and groups provide a rigorous framework for
-reasoning about graphics, interactive or otherwise. Ultimately, these
-concepts may empower us to build a broader class of robust interactive
-data visualizations.
+Concepts like groups and monoids provide a solid framework for reasoning
+about graphics, interactive or otherwise. By grounding our thinking in
+these concepts, we may be able to offload some of the required cognitive
+effort and envision new types of robust and expressive interactive
+graphics.
 

### 2024-10-23 paper.Rmd
Wrote some more

@@
 #| Top row: dodging (left) and stacking/highlighting (right), both by William Playfair [-@playfair1801; -@playfair1822]. 
-#| Bottom row: layering, created with `ggplot2` [@wickham2010; the source of the underlying data is the U.S. fuel economy data set, @mpg2024]."
+#| Bottom row: layering, created with `ggplot2` 
+#| [@wickham2010; the source of the underlying data is the U.S. fuel economy data set, @mpg2024]."
 #| message: false
@@
 
-1. **Be selected**
-2. **Represent selection**
+1. Be selected
+2. Represent selection
 3. (both in a visually and statistically sound way).
@@
 
-```{r magritte}
+```{r}
 #| echo: false
+#| fig-cap: "\\label{fig:magritte}The representation of a thing is not the thing itself. 
+#| Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe. 
+#| Right: similarly, an image filled with coloured rectangles is not a barplot, 
+#| unless the rectangles accurately represent some underlying data."
 #| fig-height: 2.5
 #| fig-show: "hold"
-#| fig-align: "center"
-#| fig-cap: "The representation of a thing is not the thing itself.
-Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe.
-Right: similarly, an image filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."
 
@@
 #| fig-height: 2.5
-#| fig-cap: "\\label{fig:sumprodmax} Different operators can be used for stacking.
+#| fig-cap: "\\label{fig:sumprodmax}Different operators can be used for stacking.
 The values `1.2`, `1.8`, `NA`, and `1.5` were assigned to a 2-by-2 table and stacked using three different operators.
@@
 Left: since convex hull is a monoid, we can use it to show a single highlighted group.
-Right: because convex hull lacks an inverse, after combining two hulls, we cannot guarantee we will be able to recover either of the original hulls - we do not know how many interior points were \"forgotten\".
+Right: because convex hull lacks an inverse, after combining two hulls, we cannot guarantee that we will be able to recover either of the original hulls - we do not know how many interior points were \"forgotten\".
 The source of the underlying data is the Palmer penguins data set [@horst2020]."
@@
   geom_point(col = "grey20", size = 0.25) +
-  labs(title = "Single group: Valid")
+  labs(title = "Valid representation")
 
@@
                col = "white", fill = NA) +
-  labs(title = "Multiple groups: Invalid")
+  labs(title = "Not a valid representation")
 

### 2024-10-23 paper.tex
Wrote some more

@@
 \item
-  \textbf{Be selected}
+  Be selected
 \item
-  \textbf{Represent selection}
+  Represent selection
 \item
@@
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/magritte-1}
-
-}
-
-\caption{The representation of a thing is not the thing itself.
-Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe.
-Right: similarly, an image filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
+\centering
+\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{\label{fig:magritte}The representation of a thing is not the
+thing itself. Left: The Treachery of Images by René Magritte
+\citeyearpar{magritte1929} makes the point that a painting of a pipe is
+not a pipe. Right: similarly, an image filled with coloured rectangles
+is not a barplot, unless the rectangles accurately represent some
+underlying data.}
 \end{figure}
@@
 \includegraphics{paper_files/figure-latex/sumprodmax-1.pdf}
-\caption{\label{fig:sumprodmax} Different operators can be used for
+\caption{\label{fig:sumprodmax}Different operators can be used for
 stacking. The values \texttt{1.2}, \texttt{1.8}, \texttt{NA}, and
@@
 \centering
-\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}
+\includegraphics{paper_files/figure-latex/unnamed-chunk-3-1.pdf}
 \caption{\label{fig:convexhull}Convex hull can be used as a valid
@@
 show a single highlighted group. Right: because convex hull lacks an
-inverse, after combining two hulls, we cannot guarantee we will be able
-to recover either of the original hulls - we do not know how many
+inverse, after combining two hulls, we cannot guarantee that we will be
+able to recover either of the original hulls - we do not know how many
 interior points were ``forgotten''. The source of the underlying data is
